### 2020-11-16 main.tex
initial commit

\PassOptionsToPackage{utf8}{inputenc}
%\documentclass{article}

\documentclass[11pt,hidelinks]{article}
%\usepackage[margin=0.75in,top=0.75in,footskip=0.5in]{geometry}
\usepackage{comment}

\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}

\usepackage[draft]{hyperref}
\usepackage{authblk}
\usepackage{fullpage}

\newcommand{\vocab}{\textbf}

\begin{document}

\title{Hogwild graph genome layout by stochastic gradient descent}

\author{ %[Heumos, \textit{et~al}.]{
  Simon Heumos\,$^{1 \dagger}$,
  Andrea Guarracino\,$^{2 \dagger}$,
  Pjotr Prins\,$^3$,
  and Erik Garrison\,$^{3 *}$
}

\affil{\small$^1$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076 \\
  $^2$University of Rome ``Tor Vergata'', Rome, Italy \\
  $^3$UTHSC, Memphis, TN, USA \\
  $^\ast$To whom correspondence should be addressed. \\
  $^\dagger$Contributed equally.}

\maketitle

\abstract{We sort graphs in 1D and 2D using hogwild stochastic gradient descent (SGD).}

\section{Introduction}

Pangenome graphs \cite{computational2016computational}.

\section{Implementation}

\section{Evaluation}

\section{Discussion}

\section*{Funding}

\bibliography{document}
\bibliographystyle{ieeetr}

\end{document}

### 2022-09-15 main.tex
add abstract

@@
 
-\title{Hogwild graph genome layout by stochastic gradient descent}
+\title{Pangenome graph layout by HOGWILD! path-guided stochastic gradient descent}
 
@@
 
-\abstract{We sort graphs in 1D and 2D using hogwild stochastic gradient descent (SGD).}
-
+\abstract{Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.
+These structures can introduce difficulty in downstream analyses, visualization, and interpretation.
+Graph sorting aims to find the best node order for a 1D and 2D layout to simplify these complex regions.
+Pangenome graphs embed pangenomic sequences as paths in the graph, but it remains an open challenge to consider this biological information during the sorting.
+Moreover, existing 2D layout methods struggle to deal with large graphs.
+We propose a new layout algorithm that orders the nodes of a pangenome graph using a HOGWILD!
+path-guided stochastic gradient descent approach: PG-SGD.
+Our implementation demonstrates that we can efficiently compute the latent structure of gigabase-scale pangenome graphs revealing their underlying biology.
+}
 \section{Introduction}

### 2022-09-19 main.tex
update abstract

--- 2020-11-16 main.tex
@@
 
-\abstract{We sort graphs in 1D and 2D using hogwild stochastic gradient descent (SGD).}
-
+\abstract{High-quality low-cost genome assemblies led to sequencing of whole populations whose genomic variation can be comparatively studied in a graphical pangenome.
+In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
+Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.
+These nonlinear structures can introduce difficulty in downstream analyses, visualization, and interpretation.
+We propose a new layout algorithm that orders the nodes of a pangenome graph using a HOGWILD!
+path-guided stochastic gradient descent approach: PG-SGD.
+Our implementation demonstrates that we can efficiently compute the latent structure of gigabase-scale pangenome graphs revealing their underlying biology.
+}
 \section{Introduction}

### 2022-09-19 main.tex
update affiliations

@@
 \author{ %[Heumos, \textit{et~al}.]{
-  Simon Heumos\,$^{1 \dagger}$,
-  Andrea Guarracino\,$^{2 \dagger}$,
-  Pjotr Prins\,$^3$,
-  and Erik Garrison\,$^{3 *}$
+  Simon Heumos\,$^{1,2 \dagger}$,
+  Andrea Guarracino\,$^{3 \dagger}$,
+  Sven Nahnsen\,$^{1,2}$,
+  Pjotr Prins\,$^4$,
+  and Erik Garrison\,$^{4 *}$
 }
 
-\affil{\small$^1$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076 \\
-  $^2$University of Rome ``Tor Vergata'', Rome, Italy \\
-  $^3$UTHSC, Memphis, TN, USA \\
+\affil{\small$^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
+  $^2$ Biomedical Data Science, Department of Computer Science, University of Tübingen , Tübingen 72076, Germany \\ 
+  $^3$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
+  $^4$
+  Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
   $^\ast$To whom correspondence should be addressed. \\

### 2022-09-19 main.tex
follow GB outline
draft outline

@@
 }
-\section{Introduction}
+\section{Background}
 
-Pangenome graphs \cite{computational2016computational}.
+Pangenome graphs \cite{computational2016computational}. SGD-layout. Layout in general.
 
-\section{Implementation}
+\section{Results}
 
-\section{Evaluation}
+Algorithm 1: How does our node update algorithm work in general, single threaded.
+\paragraph{}
+Fig. 1: Describe how our approach works. Explanation of 1D graph sorting and metric with Simon's handmade figures. SGD figure. Zipfian distribution figure.
+\paragraph{}
+Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes. Time + RAM by number of threads. Chr6 HPRC graph?
+\\
+1D (max. threads) vs. ALIBI; we should compare time + RAM. But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper. 
+\\
+2D (max. threads) + odgi draw vs. Bandage.
+\paragraph{}
+Fig.
+3: Cool quantitative 1D sortings and 2D layouts: biological implications.
+Randomly sorted.
+PG-SGD sorted.
+Ygs sorted.
+We want a pipeline of sortings.
+2D layout of the whole HPRC.
+Chr6 HPRC HLA graph?
+Chr8 beta-defensin gene cluster HPRC?
+Whole HPRC?
+\paragraph{Bonus Section}
+Fig.
+4: Detect tension.
+Relax a graph.
+Detect tension afterwards.
+I need to test this on a new data set I got from Erik.
+I need to establish a fixed lower boundary for the tension from which on we don't relax anymore.
 
@@
 
-\section*{Funding}
+\section{Conclusion}
+
+\section{Methods}
 
@@
 
+\section*{Acknowledgements}
+
+\section*{Funding}
+
+\section*{Competing interests}
+The authors declare that they have no competing interests.
+
+\section*{Availability of data and materials}
+
 \end{document}

### 2022-09-19 main.tex
drafting the discussion

@@
 
-Pangenome graphs \cite{computational2016computational}. SGD-layout. Layout in general.
+Pangenome graphs \cite{computational2016computational}. blubb.
+
+variation graph model
+
+SGD-layout. Layout in general.
 
@@
 
+\paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
 Algorithm 1: How does our node update algorithm work in general, single threaded.
-\paragraph{}
-Fig. 1: Describe how our approach works. Explanation of 1D graph sorting and metric with Simon's handmade figures. SGD figure. Zipfian distribution figure.
-\paragraph{}
+\\
+Fig.
+1: Describe how our approach works.
+Explanation of 1D graph sorting and metric with Simon's handmade figures.
+SGD figure.
+Zipfian distribution figure.
+One basic 1D and 2D viz to explain how to actually read these.
+\paragraph{Performance evaluation}
 Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes. Time + RAM by number of threads. Chr6 HPRC graph?
@@
 2D (max. threads) + odgi draw vs. Bandage.
-\paragraph{}
+\paragraph{Latent graph structure reveals underlying biology}
 Fig.
@@
 
+We propose / implemented ....
+\paragraph{}
+Difference to other existing methods, are there possible improvements of our method possible?
+\paragraph{}
+Discuss performance eval
+@Jiajie + Niklas: What about going GPU?
+\paragraph{}
+The algorithm allows to inspect a pangenome graph on base-level, as a whole.
+\paragraph{from the ODGI paper:}
+Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.
+However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
+Recent interactive pangenome graph browsers are reference-centric (Beyer et al., 2019; Yokoyama et al., 2019), have a limited predefined coordinate system (Durant et al., 2021), or focus primarily on 2D representations (Gonnella et al., 2019; Wick et al., 2015).
+Our graph sorting and layout algorithms can provide the foundation for future tools of this type.
+We plan to focus on using these learned models to detect structural variation and assembly errors.
+\paragraph{}
+The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
+A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
+Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
+Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
+\paragraph{}
+What about the future? 
+\paragraph{}
+How can other scientists benefit from this work?
 \section{Conclusion}
@@
 \section*{Acknowledgements}
+The authors thank members of the HPRC Pangenome Working Group for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
 
 \section*{Funding}
+S.H.
+acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
+S.N.
+acknowledges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT)—EXC 2180–390900677.
+This work was supported by the BMBF-funded de.
+NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) [031A537B, 031A533A, 031A538A, 031A533B, 031A535A, 031A537C, 031A534A and 031A532B].
 

### 2022-09-19 main.tex
filling some gaps

@@
 Pangenome graphs \cite{computational2016computational}. blubb.
-
+\\
 variation graph model
-
+\\
 SGD-layout. Layout in general. 
-
+\\
 \section{Results}
@@
 \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
+Our algorithm is inspired by Zheng et al.
+(\url{https://doi.org/hkv4}) and it is applicable in 1D and in 2D.
+In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
+Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
+This allows the PG-SGD to be applied to gigabase-scale pangenome graphs.
+\\
+Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}).
+The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
+Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}), or ALIBI (\url{https://doi.org/hkv3}).
+\\
 Algorithm 1: How does our node update algorithm work in general, single threaded.
@@
 I need to establish a fixed lower boundary for the tension from which on we don't relax anymore.
+\\
+With a high quality layout, we can measure the discrepancy of the path layout position versus the expected path nucleotide position, the “tension” of a graph.
+The greater the “tension”, the greater is the possibility of a biologically meaningless alignment.
+This allows us to detect telomere collapsing alignment errors and hopefully (pangenome) assembly errors, subsequently correcting them.
 

### 2022-09-21 manuscript/main.tex
update affiliations

--- 2022-09-19 main.tex
@@
   Andrea Guarracino\,$^{3 \dagger}$,
+  Jan-Niklas Manuel Schmelzle\,$^{4,5}$,
+  Jiajie Li\,$^{4,5}$,
   Sven Nahnsen\,$^{1,2}$,
-  Pjotr Prins\,$^4$,
-  and Erik Garrison\,$^{4 *}$
+  Pjotr Prins\,$^6$,
+  and Erik Garrison\,$^{6 *}$
 }
@@
 \affil{\small$^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
-  $^2$ Biomedical Data Science, Department of Computer Science, University of Tübingen , Tübingen 72076, Germany \\ 
+  $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen , Tübingen 72076, Germany \\ 
   $^3$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
-  $^4$
-  Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
+  $^4$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
+  $^5$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
+  $^6$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
   $^\ast$To whom correspondence should be addressed. \\

### 2022-09-21 manuscript/main.tex
update affiliations

@@
   Jiajie Li\,$^{4,5}$,
+  Zhiru Zhang\,$^{5}$,
   Sven Nahnsen\,$^{1,2}$,
@@
 \section*{Funding}
+JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.
 S.H.

### 2022-09-21 manuscript/main.tex
update affiliations

@@
   Jan-Niklas Manuel Schmelzle\,$^{4,5}$,
-  Jiajie Li\,$^{4,5}$,
+  Jiajie Li\,$^{5}$,
   Zhiru Zhang\,$^{5}$,

### 2022-09-22 manuscript/fig_algorithm.tex
figure 1 outline

\begin{figure*}[!htb]
	\centering
		%\hline
		\begin{subfigure}[t]{0.4\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_metrics_explained.pdf}
			\label{fig:metrics_explained}
		\end{subfigure}
		\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
		\label{fig:sorting_explained}
		\end{subfigure}
		%\smallskip
		\begin{subfigure}[t]{0.4\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_pg_sgd_update_sketch.png}
			\label{fig:pg_sgd_update_sketch}
		\end{subfigure}
		\begin{subfigure}[t]{0.4\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
			\label{fig:fig:zipf_curve}
		\end{subfigure}
	\caption{
		Explain metrics and algorithms.
		\textbf{(a)} Sketch of 1D sorting operation. \textbf{(b)} Sketch of metrics calculation. \textbf{(c)} Term updated sketch \textbf{(d)} Zipfian distribution.
	}
	\label{fig:algorithm}
\end{figure*}

### 2022-09-22 manuscript/main.tex
figure 1 outline

@@
 \usepackage{fullpage}
+% singlelinecheck=false puts subcaptions on the left
+\usepackage[singlelinecheck=false]{subcaption}
+%\usepackage{tikz,graphics,color,fullpage,float,epsf,caption,subcaption}
+\usepackage{graphicx}
+
+\usepackage{amsfonts}
+\usepackage{booktabs}
+\usepackage{siunitx}
 
@@
 Algorithm 1: How does our node update algorithm work in general, single threaded.
-\\
+
 Fig.
-1: Describe how our approach works.
-Explanation of 1D graph sorting and metric with Simon's handmade figures.
-SGD figure.
-Zipfian distribution figure.
-One basic 1D and 2D viz to explain how to actually read these.
+1: Describe how our approach works (Fig.
+\ref{fig:algorithm}).
+Explanation of 1D graph sorting and metric with Simon's handmade figures \ref{fig:sorting_explained},\ref{fig:metrics_explained}.
+SGD figure \ref{fig:pg_sgd_update_sketch}.
+Zipfian distribution figure \ref{fig:fig:zipf_curve}.
+
+\input{fig_algorithm}
+
 \paragraph{Performance evaluation}
@@
 Ygs sorted.
+Reference sorted.
 We want a pipeline of sortings.
@@
 Whole HPRC?
+\\
+We could also try to build a gastric cancer pangenome graph with data from \url{https://www.nature.com/articles/s41467-022-33073-7} from up to 185 samples.
+However, we would have to request access to the data.
+\\
+\\
+Table 1: Metrics of the sorted graphs in Fig. 3.
 \paragraph{Bonus Section}

### 2022-09-22 manuscript/fig_latent_graph_structure.tex
outline finished

\begin{figure*}[!htb]
	\centering
	%\hline
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/DRB1-3123.fa.gz.c666522.417fcdf.seqwish.og.r.og.png}
		\label{fig:random}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/DRB1-3123.fa.gz.c666522.417fcdf.seqwish.og.r.og.ud.png}
		\label{fig:random_pos}
	\end{subfigure}
	%\smallskip
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/DRB1-3123.fa.gz.c666522.417fcdf.seqwish.og.Y.og.png}
		\label{fig:sorted}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/DRB1-3123.fa.gz.c666522.417fcdf.seqwish.og.Y.og.ud.png}
		\label{fig:sorted_pos}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/DRB1-3123.fa.gz.c666522.417fcdf.seqwish.og.Ygs.og.png}
		\label{fig:pipeline}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/DRB1-3123.fa.gz.c666522.417fcdf.seqwish.og.Ygs.og.ud.png}
		\label{fig:pipeline_pos}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/DRB1-3123.fa.gz.c666522.417fcdf.seqwish.og.YH.og.png}
		\label{fig:ref}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/DRB1-3123.fa.gz.c666522.417fcdf.seqwish.og.YH.og.ud.png}
		\label{fig:ref_pos}
	\end{subfigure}
		\begin{subfigure}[t]{0.8\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/latent_graph_structure/TODO_2d.png}
		\label{fig:2d}
	\end{subfigure}
	\caption{
		Latent graph structures.
		\textbf{(a)} Randomly sorted graph. \textbf{(b)} Randomly sorted graph by position. \textbf{(c)} PG-SGD sorted graph. \textbf{(d)} PG-SGD sorted graph by position. \textbf{(e)} Ygs sorted graph. \textbf{(f)} Ygs sorted graph by position. \textbf{(g)} Reference sorted graph. \textbf{(h)} Reference sorted graph by position. \textbf{(i)} 2D layout of graph.
	}
	\label{fig:latent_graph_structure}
\end{figure*}

### 2022-09-22 manuscript/fig_performance.tex
outline finished

\begin{figure*}[!htb]
	\centering
	%\hline
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/performance/TODO_by_haps_eval_time_ram.pdf}
		\label{fig:haps_time_ram}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/performance/TODO_by_threads_eval_time_ram.pdf}
		\label{fig:threads_time_ram}
	\end{subfigure}
	%\smallskip
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/performance/TODO_1d_alibi_time_ram.pdf}
		\label{fig:alibi}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/performance/TODO_by_2d_draw_bandage_time_ram.pdf}
		\label{fig:bandage}
	\end{subfigure}
	\caption{
		Performance evaluations.
		\textbf{(a)} PG-SGD 1D and 2D by haplotypes. \textbf{(b)} PG-SGD 1D and 2D by threads. \textbf{(c)} PG-SGD vs. ALIBI. \textbf{(d)} PG-SGD + draw vs. BANDAGE.
	}
	\label{fig:performance}
\end{figure*}

### 2022-09-22 manuscript/fig_tension.tex
outline finished

\begin{figure*}[!htb]
	\centering
	%\hline
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/tension/tension_bed.png}
		\label{fig:tension_bed}
	\end{subfigure}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/tension/tension_bed_relaxed.png}
		\label{fig:tension_extracted}
	\end{subfigure}
\\
	%\smallskip
	\begin{subfigure}[t]{0.1\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/tension/layout.png}
		\label{fig:tension_draw}
	\end{subfigure}
	\begin{subfigure}[t]{0.1\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/tension/layout_relaxed.png}
		\label{fig:tension_draw_extracted}
	\end{subfigure}
	\caption{
		Detecting tension and relaxing a pangenome graph.
		\textbf{(a)} Tension detection before relaxation. \textbf{(b)} Tension detection after relaxation. \textbf{(c)} Folded 2D before relaxation. \textbf{(d)} Linearized 2D after relaxation.
	}
	\label{fig:tension}
\end{figure*}

### 2022-09-22 manuscript/main.tex
outline finished

@@
 \usepackage{siunitx}
+
+\usepackage{algorithm2e}
+\SetAlgoLined
+\SetKwProg{MyStruct}{For each}{:}{end}
 
@@
 \\
-Algorithm 1: How does our node update algorithm work in general, single threaded.
+We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
+\begin{algorithm}
+	\MyStruct{Iteration}{
+		Perform term update.
+	}
+	\caption{TODO: PG-SGD algorithm.}
+	\label{alg:pg_sgd}
+\end{algorithm}
 
@@
 \paragraph{Performance evaluation}
-Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes. Time + RAM by number of threads. Chr6 HPRC graph?
+Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Chr6 HPRC graph?
 \\
-1D (max. threads) vs. ALIBI; we should compare time + RAM. But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper. 
+1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}. But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1. 
 \\
-2D (max. threads) + odgi draw vs. Bandage.
+2D (max. threads) + odgi draw vs. Bandage in Fig.~\ref{fig:bandage}. 
+\input{fig_performance}
+\\
+\\
+Table 1: Metrics of the ALIBI and PG-SGD graphs from Fig. 2.
+\input{tab_performance.tex}
 \paragraph{Latent graph structure reveals underlying biology}
@@
 \\
-Table 1: Metrics of the sorted graphs in Fig. 3.
+\input{fig_latent_graph_structure}
+Table 2: Metrics of the sorted graphs in Fig. 3.
+\input{tab_metrics_latent_graph_structure}
 \paragraph{Bonus Section}
@@
 This allows us to detect telomere collapsing alignment errors and hopefully (pangenome) assembly errors, subsequently correcting them.
+\input{fig_tension}
 

### 2022-09-23 manuscript/main.tex
Germanize row length

@@
 
-\abstract{High-quality low-cost genome assemblies led to sequencing of whole populations whose genomic variation can be comparatively studied in a graphical pangenome.
+\abstract{
+    High-quality low-cost genome assemblies led to sequencing of whole populations whose genomic variation can be comparatively studied in a graphical pangenome.
     In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
@@
     These nonlinear structures can introduce difficulty in downstream analyses, visualization, and interpretation.
-We propose a new layout algorithm that orders the nodes of a pangenome graph using a HOGWILD!
-path-guided stochastic gradient descent approach: PG-SGD.
+    We propose a new layout algorithm that orders the nodes of a pangenome graph using a HOGWILD! path-guided stochastic gradient descent approach: PG-SGD.
     Our implementation demonstrates that we can efficiently compute the latent structure of gigabase-scale pangenome graphs revealing their underlying biology.
 }
+
 \section{Background}
@@
 \\
+
 \section{Results}
@@
 \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
-Our algorithm is inspired by Zheng et al.
-(\url{https://doi.org/hkv4}) and it is applicable in 1D and in 2D.
+Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is applicable in 1D and in 2D.
 In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
 Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
-This allows the PG-SGD to be applied to gigabase-scale pangenome graphs.
-\\
+This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
 Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}).
@@
 \\
-1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}. But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1. 
+1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}.
+But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1.
 \\
@@
 \paragraph{Latent graph structure reveals underlying biology}
-Fig.
-3: Cool quantitative 1D sortings and 2D layouts: biological implications.
-Randomly sorted.
-PG-SGD sorted.
-Ygs sorted.
-Reference sorted.
-We want a pipeline of sortings.
-2D layout of the whole HPRC.
-Chr6 HPRC HLA graph?
-Chr8 beta-defensin gene cluster HPRC?
-Whole HPRC?
+Fig. 3: Cool quantitative 1D sortings and 2D layouts: biological implications.
+andomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
+We want a pipeline of sortings. 2D layout of the whole HPRC.
+Chr6 HPRC HLA graph? Chr8 beta-defensin gene cluster HPRC? Whole HPRC?
 \\
@@
 \paragraph{Bonus Section}
-Fig.
-4: Detect tension.
-Relax a graph.
-Detect tension afterwards.
+Fig. 4: Detect tension. Relax a graph. Detect tension afterwards.
 I need to test this on a new data set I got from Erik.
-I need to establish a fixed lower boundary for the tension from which on we don't relax anymore.
-\\
+I need to establish a fixed lower boundary for the tension from which on we don't relax anymore. \\
 With a high quality layout, we can measure the discrepancy of the path layout position versus the expected path nucleotide position, the “tension” of a graph.

### 2022-09-23 manuscript/main.tex
AG acknowledges someone
tweaks

@@
 \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
-Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is applicable in 1D and in 2D.
+Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
+In this work, we describe the results in 1D and 2D.
 In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
@@
 Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}).
-The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
-Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}), or ALIBI (\url{https://doi.org/hkv3}).
+%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
+Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
 \\
@@
 \paragraph{}
-How can other scientists benefit from this work?
+How can other scientists benefit from this work? Graph visualization (together with annotation) is key for understanding variations.
 \section{Conclusion}
@@
 \section*{Acknowledgements}
-The authors thank members of the HPRC Pangenome Working Group for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
+We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
 
@@
 JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.
-S.H.
-acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
-S.N.
-acknowledges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT)—EXC 2180–390900677.
+S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
+S.N. acknowledges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT)—EXC 2180–390900677.
 This work was supported by the BMBF-funded de.
 NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) [031A537B, 031A533A, 031A538A, 031A533B, 031A535A, 031A537C, 031A534A and 031A532B].
+A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
 

### 2022-09-27 manuscript/main.tex
tweaks
more outline

@@
     High-quality low-cost genome assemblies led to sequencing of whole populations whose genomic variation can be comparatively studied in a graphical pangenome.
-    In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
+    %In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
     Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.
@@
 \section{Background}
-
+A pangenome1 models the entire set of genomic elements of a given population.
+It can efficiently be encoded2 in the form of a variation graph, which embeds the linear sequences of the pangenome as paths in the graphs themselves.
+\\
+In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
+\\
 Pangenome graphs \cite{computational2016computational}. blubb.
@@
 This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
-Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}).
+Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
+\begin{itemize}
+	\item The first node Xi of a pair is a uniform path step pick from all nodes.
+	\item The second node Xj of a pair is sampled from the same path following a Zipfian distribution. 
+	\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes. The magnitude r of the update depends on the current learning rate of the SGD.
+\end{itemize}
 %The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
-Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
-\\
 We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
@@
 \paragraph{Performance evaluation}
-Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Chr6 HPRC graph?
+Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
 \\
@@
 \input{tab_performance.tex}
+Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
 \paragraph{Latent graph structure reveals underlying biology}
 Fig. 3: Cool quantitative 1D sortings and 2D layouts: biological implications.
-andomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
+Randomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
 We want a pipeline of sortings. 2D layout of the whole HPRC.
@@
 \paragraph{}
+I'd like to argue that this projection presents biologically relevant structures that we can work with downstream
+both for human interface
+and for other algorithms that look at and classify variation
+\paragraph{}
 What about the future?
 \paragraph{}
-How can other scientists benefit from this work? Graph visualization (together with annotation) is key for understanding variations.
+How can other scientists benefit from this work?
+Graph visualization (together with annotation) is key for understanding variations.
+This projection presents biologically relevant structures that we can work with downstream for researches and algorithms that look at and classify variation,
 \section{Conclusion}

### 2023-01-23 manuscript/fig_algorithm.tex
outline finished
sketching the sketches

@@
 		%\hline
-		\begin{subfigure}[t]{0.4\textwidth}
+		\begin{subfigure}[t]{0.32\textwidth}
 			\centering
@@
 			\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_metrics_explained.pdf}
-			\label{fig:metrics_explained}
+			\label{fig:1d_before_update}
 		\end{subfigure}
-		\begin{subfigure}[t]{0.4\textwidth}
+		\begin{subfigure}[t]{0.32\textwidth}
 		\centering
@@
 		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
-		\label{fig:sorting_explained}
+		\label{fig:1d_before_update_highlighted}
+		\end{subfigure}
+		\begin{subfigure}[t]{0.32\textwidth}
+		\centering
+		\caption{}
+		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
+		\label{fig:1d_after_update}
 		\end{subfigure}	
 		%\smallskip
-		\begin{subfigure}[t]{0.4\textwidth}
+		\begin{subfigure}[t]{0.32\textwidth}
 			\centering
@@
 			\includegraphics[width=\linewidth]{fig/algorithm/TODO_pg_sgd_update_sketch.png}
-			\label{fig:pg_sgd_update_sketch}
+			\label{fig:2d_before_update}
 		\end{subfigure}
-		\begin{subfigure}[t]{0.4\textwidth}
+		\begin{subfigure}[t]{0.32\textwidth}
 			\centering
@@
 			\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
-			\label{fig:fig:zipf_curve}
+			\label{fig:2d_before_update_highlighted}
+		\end{subfigure}
+		\begin{subfigure}[t]{0.32\textwidth}
+		\centering
+		\caption{}
+		\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
+		\label{fig:2d_after_update}
 		\end{subfigure}	
 	\caption{
-		Explain metrics and algorithms.
-		\textbf{(a)} Sketch of 1D sorting operation. \textbf{(b)} Sketch of metrics calculation. \textbf{(c)} Term updated sketch \textbf{(d)} Zipfian distribution.
+		PG-SGD update operation sketches.
+\textbf{(a)-(c)} 1D layout sketch of an update operation moving a pair of nodes at a time.
+Edges are not shown.
+\textbf{(d)-(f)} 2D layout sketch of an update operation moving a pair of nodes at a time.
+Edges are drawn as black links.
+		\textbf{(a)} State of the graph before the 1D update.
+\textbf{(b)} State of the graph before the update highlighting the discrepancy of the layout distance versus the path nucleotide distance of the blue nodes in 1D.
+\textbf{(c)} State of the graph after the 1D update.
+\textbf{(d)} State of the graph before the 2D update.
+\textbf{(e)} State of the graph before the update highlighting the discrepancy of the layout distance versus the path nucleotide distance of the blue nodes in 2D.
+\textbf{(f)} State of the graph after the 2D update.
 	}
-	\label{fig:algorithm}
+	\label{fig:sketches}
 \end{figure*}

### 2023-01-23 manuscript/main.tex
sketching the sketches

@@
 SGD-layout. Layout in general.
-\\
+\\ Very light background.
 
@@
 Fig.
-1: Describe how our approach works (Fig.
-\ref{fig:algorithm}).
-Explanation of 1D graph sorting and metric with Simon's handmade figures \ref{fig:sorting_explained},\ref{fig:metrics_explained}.
-SGD figure \ref{fig:pg_sgd_update_sketch}.
-Zipfian distribution figure \ref{fig:fig:zipf_curve}.
+1: Describe how our approach works, especially a single update operation (Fig.
+\ref{fig:sketches}).
+Explanation of 1D graph updating in Figures \ref{fig:1d_before_update}-\ref{fig:1d_after_update}.
+Explanation of 2D graph updating in Figures \ref{fig:2d_before_update}-\ref{fig:2d_after_update}.
+Zipfian distribution.
 
@@
 
+% something for the supplementary
 \paragraph{Performance evaluation}

### 2023-01-24 manuscript/fig_sketches.tex
some wild sketches appear!

--- 2023-01-23 manuscript/fig_algorithm.tex
@@
 			\caption{}
-			\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_metrics_explained.pdf}
+			\includegraphics[width=\linewidth]{fig/sketches/1D_sketch_before_update.pdf}
 			\label{fig:1d_before_update}
@@
 		\caption{}
-		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
+		\includegraphics[width=\linewidth]{fig/sketches/1D_sketch_before_update_highlighted.pdf}
 		\label{fig:1d_before_update_highlighted}
@@
 		\caption{}
-		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
+		\includegraphics[width=\linewidth]{fig/sketches/1D_sketch_after_update.pdf}
 		\label{fig:1d_after_update}
@@
 			\caption{}
-			\includegraphics[width=\linewidth]{fig/algorithm/TODO_pg_sgd_update_sketch.png}
+			\includegraphics[width=\linewidth]{fig/sketches/2D_sketch_before_update.pdf}
 			\label{fig:2d_before_update}
@@
 			\caption{}
-			\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
+			\includegraphics[width=\linewidth]{fig/sketches/2D_sketch_before_update_highlighted.pdf}
 			\label{fig:2d_before_update_highlighted}
@@
 		\caption{}
-		\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
+		\includegraphics[width=\linewidth]{fig/sketches/2D_sketch_after_update.pdf}
 		\label{fig:2d_after_update}
@@
 Edges are not shown.
-\textbf{(d)-(f)} 2D layout sketch of an update operation moving a pair of nodes at a time.
-Edges are drawn as black links.
+%https://docs.google.com/presentation/d/1Htp2bkCNIXea0QnWuKpi2YzvR4fXppNfLDsihCjcmWM/edit#slide=id.p
+		\textbf{(d)-(f)} 2D layout sketch of an update operation moving a pair of nodes at a time. Edges are drawn as black links.
 		\textbf{(a)} State of the graph before the 1D update.

### 2023-01-24 manuscript/main.tex
some wild sketches appear!
1/2 PG-SGD pseudocode

@@
 \usepackage{algorithm2e}
+\usepackage[usenames,dvipsnames]{xcolor}
 \SetAlgoLined
-\SetKwProg{MyStruct}{For each}{:}{end}
+\SetKwProg{foreach}{For each}{:}{end}
+\SetKwProg{for}{For}{:}{end}
+\SetKwProg{pgsgd}{PG-SGD}{:}{end}
+\usepackage{bm}
 
@@
 \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
-Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
+Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is theoretically applicable in any number of dimensions.
 In this work, we describe the results in 1D and 2D.
@@
 %The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
-We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
+We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread. We can have several worker threads but only one checker thread!
 \begin{algorithm}
-	\MyStruct{Iteration}{
-		Perform term update.
+\pgsgd{\textbf{(}V\textbf{)}}{
+	\textbf{input:} variation graph $V = (N, E, P)$ \\
+	%https://tex.stackexchange.com/questions/22643/how-to-write-letters-in-bold-in-the-math-mode
+	\textbf{output:} $k$-dimensional layout \boldsymbol{$L$} with $n$ nodes \\
+    \boldsymbol{$P$} $\leftarrow PathIndex(G)$ \\ 
+	\boldsymbol{$L$} $\leftarrow RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization}
+	\for{$\eta$ $in$ $annealing$ $schedule$}{
+		\foreach{$term$ $update$	} {
+			DO SOME MAGIC.
 		}
-	\caption{TODO: PG-SGD algorithm.}
+	}
+}
+\caption{Pseudocode of the PG-SGD algorithm. Explain $\eta$. Explain $PathIndex$ in the text. Explain $RandomLayout$ in the text. Explain annealing schedule in the text or rephrase this wording.}
 \label{alg:pg_sgd}
@@
 
-\input{fig_algorithm}
+\input{fig_sketches}
 

### 2023-02-21 manuscript/main.tex
algorithm sketch

@@
     \boldsymbol{$P$} $\leftarrow PathIndex(G)$ \\ 
-	\boldsymbol{$L$} $\leftarrow RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization}
+    \boldsymbol{$Z$} $\leftarrow ZipfZetas(G,P)$ \\ %@Andrea is this correct like this?
+	\boldsymbol{$L$} $\leftarrow RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization, 1D starts with the current node order}
+	% atomic positions initialization?
+	% TODO for somplicity reasons I would only describe the 2D one here
 	\for{$\eta$ $in$ $annealing$ $schedule$}{
-		\foreach{$term$ $update$	} {
-			DO SOME MAGIC.
+		\foreach{$term$ $update$} {
+			random path step from all steps \\
+			if cooling $||$ flip \\
+			uniform step across the path \\
+			else \\
+			forward \\
+				use Z to generate max jump length \\
+				newsteprank = cursteprank + sampledzipfvalue \\
+			backward \\
+				use Z to generate jump length \\
+				newsteprank = cursteprank + sampledzipfvalue \\
+			get path positions via path index \\	
+			$nuc\_dist=abs(pos\_in\_path\_a - pos\_in\_path\_b)$ \\
+			$term\_weight = 1.0 / term\_dist$ \\
+			$w\_ij = term\_weight$ \\
+			$mu = eta.load() * w_{ij}$ \\
+			$dx = X[i].load() - X[j].load()$ \tcp{calculate layout distance }
+			$mag = std::abs(dx)$ \\
+			$Delta = mu * (mag - d_ij) / 2$ \\
+			$ Delta_abs = std::abs(Delta)$ \\
+			\tcp{potentially store new Delta\_max}
+			$r = Delta / mag$ \\
+			$ r_x = r * dx$ \\
+			\tcp{new X = old X - r\_x}
+			\tcp{new Y = old Y - r\_Y, 2D only}
 		}
@@
 }
-\caption{Pseudocode of the PG-SGD algorithm. Explain $\eta$. Explain $PathIndex$ in the text. Explain $RandomLayout$ in the text. Explain annealing schedule in the text or rephrase this wording.}
+\caption{Pseudocode of the PG-SGD algorithm.
+Explain $\eta$.
+Explain $PathIndex$ in the text.
+Explain $ZipfZetas$ in the text.
+Explain $RandomLayout$ in the text.
+Explain annealing schedule in the text or rephrase this wording.}
 \label{alg:pg_sgd}

### 2023-05-27 manuscript/main.tex
tweak abstract

@@
 
-\title{Pangenome graph layout by HOGWILD! path-guided stochastic gradient descent}
+\title{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
 
@@
   Simon Heumos\,$^{1,2 \dagger}$,
-  Andrea Guarracino\,$^{3 \dagger}$,
+  Andrea Guarracino\,$^{3,6 \dagger}$,
   Jan-Niklas Manuel Schmelzle\,$^{4,5}$,
@@
 \affil{\small$^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
-  $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen , Tübingen 72076, Germany \\
+  $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen, Tübingen 72076, Germany \\
   $^3$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
@@
 \abstract{
-    High-quality low-cost genome assemblies led to sequencing of whole populations whose genomic variation can be comparatively studied in a graphical pangenome.
-    %In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
-    Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.
-    These nonlinear structures can introduce difficulty in downstream analyses, visualization, and interpretation.
-    We propose a new layout algorithm that orders the nodes of a pangenome graph using a HOGWILD! path-guided stochastic gradient descent approach: PG-SGD.
-    Our implementation demonstrates that we can efficiently compute the latent structure of gigabase-scale pangenome graphs revealing their underlying biology.
+    \textbf{Motivation:}
+	The availability of many high-quality genomes has led to the need for models that allow the study of genetic variability within entire populations.
+	Pangenome graphs can represent the full diversity between multiple genomes, but they may exhibit complex structures due to common patterns of genome variation.
+    These nonlinear structures introduce difficulty in downstream analyses, visualization, and interpretation.
+    \textbf{Results:}
+	To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
+	Our approach orders the nodes of a pangenome graph by leveraging the HOGWILD! strategy and it is guided by the genomes represented in the graph.
+	We show that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, exposing their biological features.
 }
@@
 \\ Very light background.
+%Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.
 

### 2023-05-29 manuscript/main.tex
switch to Bx template

@@
 \PassOptionsToPackage{utf8}{inputenc}
-%\documentclass{article}
+\documentclass{bioinfo}
 
-\documentclass[11pt,hidelinks]{article}
-%\usepackage[margin=0.75in,top=0.75in,footskip=0.5in]{geometry}
+\usepackage{makecell}
+
+\usepackage{floatrow}
+
 \usepackage{comment}
 
-\usepackage{helvet}
-\renewcommand{\familydefault}{\sfdefault}
+\usepackage{siunitx}
 
-\usepackage[draft]{hyperref}
-\usepackage{authblk}
-\usepackage{fullpage}
 % singlelinecheck=false puts subcaptions on the left
 \usepackage[singlelinecheck=false]{subcaption}
-%\usepackage{tikz,graphics,color,fullpage,float,epsf,caption,subcaption}
-\usepackage{graphicx}
+
+\usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
 
@@
 \usepackage{booktabs}
-\usepackage{siunitx}
 
@@
 \SetAlgoLined
-\SetKwProg{foreach}{For each}{:}{end}
 \SetKwProg{for}{For}{:}{end}
 \SetKwProg{pgsgd}{PG-SGD}{:}{end}
+\SetKwProg{each}{For each}{:}{end}
 \usepackage{bm}
 
+% we squeeze our figures even more together
+\captionsetup{belowskip=-2pt}
+
+\SetAlgoLined
+\SetKwProg{MyStruct}{Struct}{ contains}{end}
+
 \newcommand{\vocab}{\textbf}
+\newcommand{\red}[1]{{\textcolor{Red}{#1}}}
+\newcommand{\FIXME}[1]{\red{[FIXME: #1]}}
+
+\usepackage{orcidlink}
+\hypersetup{hidelinks}
+
+\def\labelitemi{--}
+
+\copyrightyear{2023} \pubyear{XXXX}
+
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Genome Analysis}
 
 \begin{document}
+\firstpage{1}
 
-\title{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
+\subtitle{Genome Analysis}
 
-\author{ %[Heumos, \textit{et~al}.]{
-  Simon Heumos\,$^{1,2 \dagger}$,
-  Andrea Guarracino\,$^{3,6 \dagger}$,
-  Jan-Niklas Manuel Schmelzle\,$^{4,5}$,
-  Jiajie Li\,$^{5}$,
-  Zhiru Zhang\,$^{5}$,
-  Sven Nahnsen\,$^{1,2}$,
-  Pjotr Prins\,$^6$,
-  and Erik Garrison\,$^{6 *}$
+\title[Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
+\author[Heumos, Guarracino \textit{et~al}.]{
+Simon~Heumos\,$^{\orcidlink{0000-0003-3326-817X}1,2,\dagger,*}$,
+Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}3,4,\dagger}$,
+Jan-Niklas Manuel Schmelzle\,$^{\orcidlink{0000-0001-8566-4049}5,6}$,
+Jiajie Li\,$^{\orcidlink{0000-0002-9467-0659}6}$,
+Zhiru Zhang\,$^{\orcidlink{0000-0002-0778-0308}6}$,
+Sven Nahnsen\,$^{\orcidlink{0000-0002-4375-0691}1,2}$,
+Pjotr Prins\,$^{\orcidlink{0000-0002-8021-9162}3}$,
+Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 3},*}$
 }
 
-\affil{\small$^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
+\address{
+$^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
 $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen, Tübingen 72076, Germany \\
-  $^3$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
-  $^4$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
-  $^5$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
-  $^6$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
+$^3$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
+$^4$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
+$^5$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
+$^6$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
+}
+
+\corresp{
 $^\ast$To whom correspondence should be addressed. \\
-  $^\dagger$Contributed equally.}
+$^{\dagger}$The authors wish it to be known that, in their opinion, the first two authors should be regarded as Joint First Authors.\
+}
 
-\maketitle
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
 
@@
 Pangenome graphs can represent the full diversity between multiple genomes, but they may exhibit complex structures due to common patterns of genome variation.
-    These nonlinear structures introduce difficulty in downstream analyses, visualization, and interpretation.
+These nonlinear structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
 \textbf{Results:}
 To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-	Our approach orders the nodes of a pangenome graph by leveraging the HOGWILD! strategy and it is guided by the genomes represented in the graph.
-	We show that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, exposing their biological features.
+Our approach orders the nodes of a pangenome graph by leveraging the HOGWILD! strategy and it is guided by the path position information of the genomes represented in the graph.
+We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, exposing their biological features. \\
+\textbf{Availability:}
+The PG-SGD algorithm is implemented in \textit{odgi} which is published as free software under the MIT open source license.
+Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}.
+\textit{odgi} can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
+\textbf{Contact:} \href{simon.heumos@qbic.uni-tuebingen.de}{simon.heumos@qbic.uni-tuebingen.de},
+\href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
+%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
 }
+
+\maketitle
 
@@
 
-\section{Results}
+\section{Algorithm}
 
@@
 We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread. We can have several worker threads but only one checker thread!
+
 \begin{algorithm}
+	\caption{Pseudocode of the PG-SGD algorithm.
+Explain $\eta$.
+Explain $PathIndex$ in the text.
+Explain $ZipfZetas$ in the text.
+Explain $RandomLayout$ in the text.
+Explain annealing schedule in the text or rephrase this wording.}
+	\label{alg:pg_sgd}
 	\pgsgd{\textbf{(}V\textbf{)}}{
@@
 		\for{$\eta$ $in$ $annealing$ $schedule$}{
-		\foreach{$term$ $update$} {
+			\each{$term$ $update$} {
 				random path step from all steps \\
@@
 	}
-\caption{Pseudocode of the PG-SGD algorithm.
-Explain $\eta$.
-Explain $PathIndex$ in the text.
-Explain $ZipfZetas$ in the text.
-Explain $RandomLayout$ in the text.
-Explain annealing schedule in the text or rephrase this wording.}
-\label{alg:pg_sgd}
 \end{algorithm}
@@
 \input{fig_sketches}
+
+\section{Results}
+\label{sec:results}
+
+%\subsection{Experimental insights}
+
+%\subsection{Limitations}
 
@@
 \section{Discussion}
+\label{sec:discussion}
 
@@
 This projection presents biologically relevant structures that we can work with downstream for researches and algorithms that look at and classify variation,
-\section{Conclusion}
 
-\section{Methods}
+\section*{Acknowledgments}
 
-\bibliography{document}
-\bibliographystyle{ieeetr}
-
-\section*{Acknowledgements}
 We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
+We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
 
 \section*{Funding}
+
 JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.
@@
 S.N. acknowledges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT)—EXC 2180–390900677.
-This work was supported by the BMBF-funded de.
-NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) [031A537B, 031A533A, 031A538A, 031A533B, 031A535A, 031A537C, 031A534A and 031A532B].
+This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A532B, 031A533A, 031A533B, 031A534A, 031A535A, 031A537A, 031A537B, 031A537C, 031A537D and 031A538A).
 A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
@@
 
-\section*{Availability of data and materials}
+\section*{Data availability}
+
+Code and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/pangenome/sorting-paper}.
+
+\bibliographystyle{natbib}
+
+\bibliography{document}
 

### 2023-05-29 manuscript/main.tex
more complete formula

@@
 \SetAlgoLined
-\SetKwProg{for}{For}{:}{end}
-\SetKwProg{pgsgd}{PG-SGD}{:}{end}
-\SetKwProg{each}{For each}{:}{end}
 \usepackage{bm}
@@
 Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
-This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
+This allows the PG-SGD to be applied manuscriptto gigabase-scale pangenome graphs. \\
 Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
@@
 
-\begin{algorithm}
-	\caption{Pseudocode of the PG-SGD algorithm.
-Explain $\eta$.
-Explain $PathIndex$ in the text.
-Explain $ZipfZetas$ in the text.
-Explain $RandomLayout$ in the text.
-Explain annealing schedule in the text or rephrase this wording.}
-	\label{alg:pg_sgd}
-	\pgsgd{\textbf{(}V\textbf{)}}{
-		\textbf{input:} variation graph $V = (N, E, P)$ \\
-		%https://tex.stackexchange.com/questions/22643/how-to-write-letters-in-bold-in-the-math-mode
-		\textbf{output:} $k$-dimensional layout \boldsymbol{$L$} with $n$ nodes \\
-		\boldsymbol{$P$} $\leftarrow PathIndex(G)$ \\ 
-		\boldsymbol{$Z$} $\leftarrow ZipfZetas(G,P)$ \\ %@Andrea is this correct like this?
-		\boldsymbol{$L$} $\leftarrow RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization, 1D starts with the current node order}
-		% atomic positions initialization?
-		% TODO for somplicity reasons I would only describe the 2D one here
-		\for{$\eta$ $in$ $annealing$ $schedule$}{
-			\each{$term$ $update$} {
-				random path step from all steps \\
-				if cooling $||$ flip \\
-				uniform step across the path \\
-				else \\
-				forward \\
-				use Z to generate max jump length \\
-				newsteprank = cursteprank + sampledzipfvalue \\
-				backward \\
-				use Z to generate jump length \\
-				newsteprank = cursteprank + sampledzipfvalue \\
-				get path positions via path index \\	
-				$nuc\_dist=abs(pos\_in\_path\_a - pos\_in\_path\_b)$ \\
-				$term\_weight = 1.0 / term\_dist$ \\
-				$w\_ij = term\_weight$ \\
-				$mu = eta.load() * w_{ij}$ \\
-				$dx = X[i].load() - X[j].load()$ \tcp{calculate layout distance }
-				$mag = std::abs(dx)$ \\
-				$Delta = mu * (mag - d_ij) / 2$ \\
-				$ Delta_abs = std::abs(Delta)$ \\
-				\tcp{potentially store new Delta\_max}
-				$r = Delta / mag$ \\
-				$ r_x = r * dx$ \\
-				\tcp{new X = old X - r\_x}
-				\tcp{new Y = old Y - r\_Y, 2D only}
-			}
-		}
-	}
-\end{algorithm}
+\input{pg-sgd_pseudocode}
 

### 2023-05-29 manuscript/pg-sgd_pseudocode.tex
more complete formula

\SetKwProg{for}{for}{:}{end}
\SetKwProg{pgsgd}{PG-SGD}{:}{end}
\SetKwProg{each}{for each}{:}{end}
\SetKwProg{IF}{if}{:}{end}

\SetKwFunction{PathIndex}{PathIndex}
\SetKwFunction{RandomLayout}{RandomLayout}
\SetKwFunction{InitZipf}{InitZipf}
\SetKwFunction{Unif}{Unif}
\SetKwFunction{Zipf}{Zipf}
\SetKwFunction{StepCount}{StepCount}
\SetKwFunction{PathId}{PathId}
\SetKwFunction{StepPos}{StepPos}

\begin{algorithm}
	\pgsgd{\textbf{(}$\mathcal{V}$\textbf{)}}{
		\textbf{input:} variation graph $\mathcal{V} = (\mathcal{N}, \mathcal{E}, \mathcal{P})$ \\
		%https://tex.stackexchange.com/questions/22643/how-to-write-letters-in-bold-in-the-math-mode
		\textbf{output:} $k$-dimensional layout $\mathcal{L}$ with $n$ nodes \\
		$\mathcal{XP}$ $\gets \PathIndex(\mathcal{V})$ \tcp{index for efficient path position lookup}
		%\boldsymbol{$Z$} $\gets ZipfZetas(G,P)$ \\ %@Andrea is this correct like this?
		$\mathcal{L}$ $\gets \RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization, 1D starts with the current node order}
		$\mathcal{Z} \gets \InitZipf(\mathcal{V},\mathcal{XP})$ \tcp{generate Zipfian distribution}
		% atomic positions initialization?
		% TODO for somplicity reasons I would only describe the 2D one here
		\for{$\eta$ $in$ $annealing$ $schedule$}{ %our "schedule" actually is the number paths.... we should specify this I would say
			\each{$\{i,j: i<j\}$ in uniform order} { % I think we can remove i<j, because we don't care about that
				$ps_i \gets \Unif(\mathcal{XP})$ \tcp{uniform path step from all steps}
				$path \gets \PathId(ps_i,\mathcal{XP})$ \tcp{the path id of the sampled step}
				\uIf{$(coolling$ $||$ $flip)$} {
					$ps_j \gets \Unif(\StepCount(path, \mathcal{XP}))$ \tcp{uniform path step from specific path}
					$p_j \gets \StepPos(ps_j)$ \tcp{nucleotide position of step}
				} \Else {
					$ps_j \gets \Zipf(path)$ \tcp{}
					use Z to generate max jump length \\
					newsteprank = cursteprank + sampledzipfvalue \\
					backward \\
					use Z to generate jump length \\
					newsteprank = cursteprank + sampledzipfvalue \\
				}
				$p_i \gets \StepPos(ps_i)$ \tcp{nucleotide position of step}
				$nd \gets ||p_i - p_j||$ \tcp{calculate nucleotide distance} %mag is nx
				$w_{ij} \gets \frac{1.0}{nd}$ \tcp{term weight} 
				$\mu \gets w_{ij}\eta$ \tcp{learning rate}  % current learning rate is given by term weight and step size
				\IF{$\mu>1$} {
				$\mu \gets 1$
				}
				$ld \gets ||\mathcal{L}_{i}-\mathcal{L}_j||$ \tcp{calculate layout distance}
				$\delta \gets \mu \cdot \frac{ld -nd}{2}$ \tcp{the actual delta}
				% TODO stop early?
				% TODO potentially store new delta max?
				$r \gets \delta - ld$ \tcp{size of the term update}
				$\mathcal{L}_i \gets \mathcal{L}_i - r\cdot ld$ \\
				$\mathcal{L}_j \gets \mathcal{L}_j - r\cdot ld$
			}
		}
	}
	\caption{Pseudocode of the PG-SGD algorithm. Explain $\eta$. Explain $PathIndex$ in the text. Explain $ZipfZetas$ in the text. Explain $RandomLayout$ in the text. Explain annealing schedule in the text or rephrase this wording.}
	\label{alg:pg_sgd}
\end{algorithm}

### 2023-05-30 manuscript/pg-sgd_pseudocode.tex
long algorithm

@@
 		% TODO for somplicity reasons I would only describe the 2D one here
-		\for{$\eta$ $in$ $annealing$ $schedule$}{ %our "schedule" actually is the number paths.... we should specify this I would say
-			\each{$\{i,j: i<j\}$ in uniform order} { % I think we can remove i<j, because we don't care about that
-				$ps_i \gets \Unif(\mathcal{XP})$ \tcp{uniform path step from all steps}
-				$path \gets \PathId(ps_i,\mathcal{XP})$ \tcp{the path id of the sampled step}
+		\for{$planned$ $term$ $updates$}{ %our "schedule" actually is the number paths.... we should specify this I would say
+			\each{$\eta$ $in$ $annealing$ $schedule$} { % I think we can remove i<j, because we don't care about that
+				$ps_a \gets \Unif(\mathcal{XP})$ \tcp{uniform path step from all steps}
+				$path \gets \PathId(ps_a,\mathcal{XP})$ \tcp{the path id of the sampled step}
 				\uIf{$(coolling$ $||$ $flip)$} {
-					$ps_j \gets \Unif(\StepCount(path, \mathcal{XP}))$ \tcp{uniform path step from specific path}
-					$p_j \gets \StepPos(ps_j)$ \tcp{nucleotide position of step}
+					$ps_b \gets \Unif(\StepCount(path, \mathcal{XP}))$ \tcp{uniform path step from specific path}
+					$p_b \gets \StepPos(ps_b)$ \tcp{nucleotide position of step}
 				} \Else {
-					$ps_j \gets \Zipf(path)$ \tcp{}
-					use Z to generate max jump length \\
-					newsteprank = cursteprank + sampledzipfvalue \\
-					backward \\
-					use Z to generate jump length \\
-					newsteprank = cursteprank + sampledzipfvalue \\
+					$ps_b \gets \Zipf(path)$ \tcp{Zipfian path step from specific path}
 				}
-				$p_i \gets \StepPos(ps_i)$ \tcp{nucleotide position of step}
-				$nd \gets ||p_i - p_j||$ \tcp{calculate nucleotide distance} %mag is nx
-				$w_{ij} \gets \frac{1.0}{nd}$ \tcp{term weight} 
-				$\mu \gets w_{ij}\eta$ \tcp{learning rate}  % current learning rate is given by term weight and step size
+				$p_a \gets \StepPos(ps_a)$ \tcp{nucleotide position of step}
+				$p_b \gets \StepPos(ps_b)$ \tcp{nucleotide position of step}
+				$nd \gets ||p_a - p_b||$ \tcp{calculate nucleotide distance} %mag is nx
+				$ld \gets ||\mathcal{L}_{a}-\mathcal{L}_b||$ \tcp{calculate layout distance}
+				$w_{ab} \gets \frac{1.0}{nd}$ \tcp{term weight} 
+				$\mu \gets w_{ab}\eta$ \tcp{learning rate}  % current learning rate is given by term weight and step size
 				\IF{$\mu>1$} {
@@
 				}
-				$ld \gets ||\mathcal{L}_{i}-\mathcal{L}_j||$ \tcp{calculate layout distance}
 				$\delta \gets \mu \cdot \frac{ld -nd}{2}$ \tcp{the actual delta}
+				\uIf{$\delta <= 0$} {
+					$STOP$ \tcp{we can't optimize it any better, so we stop here}
+				}
 				% TODO stop early?
@@
 				$r \gets \delta - ld$ \tcp{size of the term update}
-				$\mathcal{L}_i \gets \mathcal{L}_i - r\cdot ld$ \\
-				$\mathcal{L}_j \gets \mathcal{L}_j - r\cdot ld$
+				$\mathcal{L}_a \gets \mathcal{L}_a - r\cdot ld$ \tcp{update the 1. term}
+				$\mathcal{L}_b \gets \mathcal{L}_b - r\cdot ld$ \tcp{update the 2. term}
 			}

### 2023-06-01 manuscript/pg-sgd_pseudocode.tex
correct for loops in algo2

@@
 		% TODO for somplicity reasons I would only describe the 2D one here
-		\for{$planned$ $term$ $updates$}{ %our "schedule" actually is the number paths.... we should specify this I would say
-			\each{$\eta$ $in$ $annealing$ $schedule$} { % I think we can remove i<j, because we don't care about that
+		\for{$\eta$ $in$ $annealing$ $schedule$}{ %our "schedule" actually is the number paths.... we should specify this I would say
+			\each{$planned$ $term$ $updates$} { % I think we can remove i<j, because we don't care about that
 				$ps_a \gets \Unif(\mathcal{XP})$ \tcp{uniform path step from all steps}

### 2023-06-15 manuscript/main.tex
back to 3 pages

@@
 \textbf{Motivation:}
-The availability of many high-quality genomes has led to the need for models that allow the study of genetic variability within entire populations.
+The availability of many high-quality genomes has led to the need for models that allow the study of genomic variability within entire populations.
 Pangenome graphs can represent the full diversity between multiple genomes, but they may exhibit complex structures due to common patterns of genome variation.
@@
 To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-Our approach orders the nodes of a pangenome graph by leveraging the HOGWILD! strategy and it is guided by the path position information of the genomes represented in the graph.
+Our approach arranges the nodes of a pangenome graph by leveraging the HOGWILD! strategy and it is guided by the path position information of the genomes represented in the graph.
 We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, exposing their biological features. \\
@@
 
+\iffalse
 Fig.
@@
 \input{fig_sketches}
+\fi
 
+\iffalse
 \section{Results}
@@
 \input{fig_tension}
+\fi
 

### 2023-06-16 scripts/primates_chr8.sh
initial primates14 chr8 script

@@ -1,0 +1,19 @@
+screen -R p14_chr8_p95_s5k_k79
+mkdir p14_chr8_p95_s5k_k79
+cd p14_chr8_p95_s5k_k79
+nextflow run nf-core/pangenome -r dev -profile cfc --input ~/data/primates/primates14.chr8.fa.gz -c /home-link/afaaj01/hprc_chr16_brave_sing.config --n_haplotypes 14 --wfmash_map_pct_id 95 --wfmash_segment_length 5000 --wfmash_chunks 20 -w work_primates14_chr8_p95_s5k_k79 --outdir results_primates14_chr8_p95_s5k_k79 --seqwish_min_match_length 79
+
+screen -R p14_chr8_p90_s5k_k79
+mkdir p14_chr8_p90_s5k_k79
+cd p14_chr8_p90_s5k_k79
+nextflow run nf-core/pangenome -r dev -profile cfc --input ~/data/primates/primates14.chr8.fa.gz -c /home-link/afaaj01/hprc_chr16_brave_sing.config --n_haplotypes 14 --wfmash_map_pct_id 90 --wfmash_segment_length 5000 --wfmash_chunks 20 -w work_primates14_chr8_p90_s5k_k79 --outdir results_primates14_chr8_p90_s5k_k79 --seqwish_min_match_length 79
+
+screen -R p14_chr8_p95_s5k_k311
+mkdir p14_chr8_p95_s5k_k311
+cd p14_chr8_p95_s5k_k311
+nextflow run nf-core/pangenome -r dev -profile cfc --input ~/data/primates/primates14.chr8.fa.gz -c /home-link/afaaj01/hprc_chr16_brave_sing.config --n_haplotypes 14 --wfmash_map_pct_id 95 --wfmash_segment_length 5000 --wfmash_chunks 20 -w work_primates14_chr8_p95_s5k_k311 --outdir results_primates14_chr8_p95_s5k_k311 --seqwish_min_match_length 311
+
+screen -R p14_chr8_p90_s5k_k311
+mkdir p14_chr8_p90_s5k_k311
+cd p14_chr8_p90_s5k_k311
[…]

### 2023-06-18 scripts/random.sh
beta-defensin coordinates and inject/paths/viz

@@ -1,0 +1,5 @@
+pggb -i primates14.chr8.fa.gz -n 14 -o primates14.chr8 -D /scratch
+
+odgi inject -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.og -b beta_defensin.bed -o primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.og -t 8 -P
+odgi paths -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.og -L | cut -f 1,2 -d '#' | sort | uniq > prefixes.txt
+odgi viz -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.og -M prefixes.txt -o primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.merged.png

### 2023-06-18 scripts/random.sh
more sortings

@@ -1,5 +1,13 @@
 pggb -i primates14.chr8.fa.gz -n 14 -o primates14.chr8 -D /scratch
 
-odgi inject -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.og -b beta_defensin.bed -o primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.og -t 8 -P
+odgi inject -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.og -b beta_defensin.bed -o primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.og -P
 odgi paths -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.og -L | cut -f 1,2 -d '#' | sort | uniq > prefixes.txt
-odgi viz -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.og -M prefixes.txt -o primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.merged.png
+odgi viz -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.og -M prefixes.txt -o primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.inject.merged.z.png -z
+
+odgi sort -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.og -t 48 -o primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.Ygs.sort_by_ref.og -p Ygs -P --temp-dir /scratch -H <(echo grch38#chr8)
+odgi inject -i primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.Ygs.sort_by_ref.og -b beta_defensin.bed -o primates14.chr8/primates14.chr8.fa.gz.667b9b6.417fcdf.8abe853.smooth.final.Ygs.sort_by_ref.inject.og -P
[…]

### 2023-06-20 manuscript/main.tex
abstract and introduction

@@
 \textbf{Motivation:}
-The availability of many high-quality genomes has led to the need for models that allow the study of genomic variability within entire populations.
-Pangenome graphs can represent the full diversity between multiple genomes, but they may exhibit complex structures due to common patterns of genome variation.
-These nonlinear structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
+The increasing availability of high-quality genomes demands models that allow for the study of genomic variability within entire populations.
+Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
+These structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
 \textbf{Results:}
 To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-Our approach arranges the nodes of a pangenome graph by leveraging the HOGWILD! strategy and it is guided by the path position information of the genomes represented in the graph.
-We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, exposing their biological features. \\
+Our method, driven by the HOGWILD! strategy, arranges the nodes of a pangenome graph using the path position information of the genomes represented in the graph.
+We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, revealing their biological features. \\
 \textbf{Availability:}
-The PG-SGD algorithm is implemented in \textit{odgi} which is published as free software under the MIT open source license.
+The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
 Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}.
-\textit{odgi} can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
-\textbf{Contact:} \href{simon.heumos@qbic.uni-tuebingen.de}{simon.heumos@qbic.uni-tuebingen.de},
-\href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
-%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
+\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu}
+\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
 }
@@
 
-\section{Background}
-A pangenome1 models the entire set of genomic elements of a given population.
-It can efficiently be encoded2 in the form of a variation graph, which embeds the linear sequences of the pangenome as paths in the graphs themselves.
-\\
-In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
-\\
-Pangenome graphs \cite{computational2016computational}. blubb.
-\\
-variation graph model
-\\
-SGD-layout. Layout in general.
-\\ Very light background.
-%Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.
+\section{Introduction}
+Reference genomes are the most widely used resource in genetics, serving as a foundation for a variety of downstream analyses
+such as gene annotation, read mapping, and variant detection (\FIXME{XXX WE NEED CITATIONS XXX}).
+However, with the availability of thousands of high-quality genomes per species, this linear model has shown its weaknesses.
+A single genome can not faithfully represent the genetic diversity of a species, leading to the reference bias problem (\FIXME{XXX WE NEED CITATIONS XXX}).
+A pangenome models the entire set of genomic elements of a given population \citep{Tettelin_2008,cpang2018,Eizenga_2020}.
+It can be represented by a graph data structure incorporating sequences as nodes and their connections as edges ((\FIXME{XXX A FIGURE MIGHT HELP XXX}).
+These nodes are shared for identical sequences, such as homologs, paralogs, and orthologs.
+In the variation graph model \citep{Garrison:2018} reads, contigs, haplotypes, or individual chromosomes are represented as paths in the graph.
+A bi-directed graph can even contain both strands of DNA as paths. \\
+
+The layout of these pangenome graphs, that is the arrangement of nodes and edges in a representation, provide an accessible representation of genetic variation across multiple genomes.
+Graph sorting aims to find the best node order for a layout to minimizes overlapping nodes or edges, reduces edge crossings, and promote intuitive understanding of the data by displaying structure and important features in an understandable way.
+\FIXME{
+	Existing layout algorithms range from simple, force-directed methods to more complex, hierarchical layouts, each presenting unique strengths and drawbacks.
+	Force-directed methods offer an aesthetic layout based on a physics simulation but may struggle with scalability for larger graphs (XXX WE NEED CITATIONS XXX).
+	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
+}
+
+With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
+efficiency, scalability, readability, and the ability to reveal significant biological patterns.
+Pangenome graphs embed linear sequences as paths in the graph, but to our knowledge, no algorithm takes into account this biological information in the sorting.
+We present a new layout algorithm to simplify a pangenome graph, by using path-guided stochastic gradient descent (\FIXME{CITATION}) to move a single pair of nodes at a time.
+\FIXME{A LITTLE BIT MORE ABOUT THE SGD APPROACH?}
+We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph buildiong and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, and maybe HPRC PAPER}).
 

### 2023-06-20 manuscript/main.tex
acknowledge cfc admin
add important notes

@@
 \hypersetup{hidelinks}
+\usepackage{appendix}
 
@@
 \maketitle
+
+\FIXME{WE ONLY WANT TO TAKE ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
+PG-SGD PRODUCES 1D and 2D LAYOUTS.
+IN 1D WE SORT A GRAPH BY PROJECTING THE ACTUAL 1D LAYOUT INTO A NODE ORDER!}
 
@@
 We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
+A special thanks goes to Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.
 
@@
 
+\begin{appendices}
+	\section{Quantization}
+	\FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.}
+\end{appendices}
+
 \end{document}

### 2023-06-20 manuscript/main.tex
layout, no sorting

@@
 A bi-directed graph can even contain both strands of DNA as paths. \\
-
 The layout of these pangenome graphs, that is the arrangement of nodes and edges in a representation, provide an accessible representation of genetic variation across multiple genomes.
-Graph sorting aims to find the best node order for a layout to minimizes overlapping nodes or edges, reduces edge crossings, and promote intuitive understanding of the data by displaying structure and important features in an understandable way.
+Graph layout methods aim to find the best node positions in the layout to minimizes overlapping nodes or edges, reduces edge crossings, and promote intuitive understanding of the data by displaying structure and important features in an understandable way.
 \FIXME{
@@
 	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
-}
-
+} \\
 With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
@@
 \FIXME{A LITTLE BIT MORE ABOUT THE SGD APPROACH?}
-We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph buildiong and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, and maybe HPRC PAPER}).
+We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph building and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, and maybe HPRC PAPER}).
 
@@
 Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
-This allows the PG-SGD to be applied manuscriptto gigabase-scale pangenome graphs. \\
-Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
+This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
+Our multi threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
 \begin{itemize}
@@
 	\item The second node Xj of a pair is sampled from the same path following a Zipfian distribution. 
-	\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes. The magnitude r of the update depends on the current learning rate of the SGD.
+	\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
+	\item The magnitude r of the update depends on the current learning rate of the SGD.
 \end{itemize}
 %The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
-We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread. We can have several worker threads but only one checker thread!
+We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
+We can have several worker threads but only one checker thread!
 

### 2023-06-21 manuscript/main.tex
fix the indentation

@@
     \author[Heumos, Guarracino \textit{et~al}.]{
-Simon~Heumos\,$^{\orcidlink{0000-0003-3326-817X}1,2,\dagger,*}$,
+        Simon~Heumos\,$^{\orcidlink{0000-0003-3326-817X}1,2,\dagger}$,
         Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}3,4,\dagger}$,
@@
         The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
-Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}.
-\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu}
+        Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
+        \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
         \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
@@
     \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
-Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is theoretically applicable in any number of dimensions.
-In this work, we describe the results in 1D and 2D.
+    Our algorithm is inspired by Zheng et al. ((\FIXME{CITATION} \url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
+    In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D).
     In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
@@
     This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
-Our multi threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
+    Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
     \begin{itemize}
@@
     \iffalse
+
     \section{Results}
@@
 % something for the supplementary
+
     \paragraph{Performance evaluation}
@@
     Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
+
     \paragraph{Latent graph structure reveals underlying biology}
@@
     \input{tab_metrics_latent_graph_structure}
+
     \paragraph{Bonus Section}
@@
     \label{sec:discussion}
+    We presented
+    We propose / implemented ....
 
-We propose / implemented ....
     \paragraph{}
     Difference to other existing methods, are there possible improvements of our method possible?
+
     \paragraph{}
@@
     @Jiajie + Niklas: What about going GPU?
+
     \paragraph{}
     The algorithm allows to inspect a pangenome graph on base-level, as a whole.
+
     \paragraph{from the ODGI paper:}
@@
     We plan to focus on using these learned models to detect structural variation and assembly errors.
+
     \paragraph{}
@@
     Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
+
     \paragraph{}
@@
     and for other algorithms that look at and classify variation
+
     \paragraph{}
     What about the future?
+
     \paragraph{}
@@
 
-JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.
     S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
@@
     A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
+    JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.
 

### 2023-06-24 manuscript/main.tex
stuff and discussion stuff

@@
         \textbf{Motivation:}
-        The increasing availability of high-quality genomes demands models that allow for the study of genomic variability within entire populations.
+        The increasing availability of high-quality genomes demands models for studying genomic variability within entire populations.
         Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
@@
         To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-        Our method, driven by the HOGWILD! strategy, arranges the nodes of a pangenome graph using the path position information of the genomes represented in the graph.
-        We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, revealing their biological features. \\
+        Our method, driven by the HOGWILD! strategy, arranges the nodes of a pangenome graph using the nucleotide sequences of the genomes represented in the graph.
+        We demonstrate that our implementation efficiently computes the latent structure of gigabase-scale pangenome graphs, revealing their biological features. \\
         \textbf{Availability:}
-        The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
-        Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
+        The PG-SGD algorithm is integrated in \text{ODGI} which is released as free software under the MIT open source license.
+        Source code is available at \url{https://github.com/pangenome/odgi}. \\ % and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
         \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
@@
 
-    \FIXME{WE ONLY WANT TO TAKE ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
-PG-SGD PRODUCES 1D and 2D LAYOUTS.
-IN 1D WE SORT A GRAPH BY PROJECTING THE ACTUAL 1D LAYOUT INTO A NODE ORDER!}
+    %\FIXME{WE ONLY WANT TO TAKE ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
+    %PG-SGD PRODUCES 1D and 2D LAYOUTS.
+    %IN 1D WE SORT A GRAPH BY PROJECTING THE ACTUAL 1D LAYOUT INTO A NODE ORDER!}
 
@@
     \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
+    \FIXME{SPECIFY THAT THE IMPLEMENTATION IS IN ODGI AND CITE IT}
     Our algorithm is inspired by Zheng et al. ((\FIXME{CITATION} \url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
@@
     \label{sec:discussion}
-    We presented
-    We propose / implemented ....
 
-    \paragraph{}
-    Difference to other existing methods, are there possible improvements of our method possible?
-
-    \paragraph{}
-    Discuss performance eval
-    @Jiajie + Niklas: What about going GPU?
-
-    \paragraph{}
-    The algorithm allows to inspect a pangenome graph on base-level, as a whole.
-
-    \paragraph{from the ODGI paper:}
-    Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.
-    However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-    Recent interactive pangenome graph browsers are reference-centric (Beyer et al., 2019; Yokoyama et al., 2019), have a limited predefined coordinate system (Durant et al., 2021), or focus primarily on 2D representations (Gonnella et al., 2019; Wick et al., 2015).
-    Our graph sorting and layout algorithms can provide the foundation for future tools of this type.
-    We plan to focus on using these learned models to detect structural variation and assembly errors.
-
-    \paragraph{}
-    The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
-    A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
-    Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
-    Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
-
-    \paragraph{}
-    I'd like to argue that this projection presents biologically relevant structures that we can work with downstream
-    both for human interface
-    and for other algorithms that look at and classify variation
-
-    \paragraph{}
-    What about the future?
-
-    \paragraph{}
-    How can other scientists benefit from this work?
-    Graph visualization (together with annotation) is key for understanding variations.
-    This projection presents biologically relevant structures that we can work with downstream for researches and algorithms that look at and classify variation,
+    We presented the Path-Guided Stochastic Gradient Descent (PG-SGD), a layout algorithm for pangenome graphs that
+    leverages the inherent biological information available within the graph's represented genomic sequences.
+    Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
+    \FIXME{Difference to other existing methods, are there possible improvements of our method possible?}
+    \FIXME{Discuss performance eval. @Jiajie + Niklas: What about going GPU?}
+    We implemented the PG-SGD algorithm to generate layouts in 1 and 2 dimensions, but the algorithm can be extended in any number of dimensions.
+    Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented 
+    high-level perspective on variation in pangenomes. These layouts are critical for pangenome graph building (\FIXME{CITE PGG and HPRC MAIN PAPER}) and 
+    making new biological discoveries (\FIXME{CITE ACROCENTRIC PAPER}).
+    %The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
+    %A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
+    %Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
+    %Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
+    However, there remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
+    Our PG-SGD algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts.
+    Futhermore, projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of 
+    algorithms that use the graph layout to look at variant classification.
+    Our future research involve exploiting these learned projections to detect structural variations and assembly errors.
+    \FIXME{PROTEIN?}
 

### 2023-06-26 manuscript/main.tex
better readable abstract

--- 2023-06-21 manuscript/main.tex
@@
 \textbf{Motivation:}
-        The increasing availability of high-quality genomes demands models that allow for the study of genomic variability within entire populations.
+The increasing availability of high quality genomes demands models that allow for the study of genomic variability within entire populations.
 Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
@@
 To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-        Our method, driven by the HOGWILD! strategy, arranges the nodes of a pangenome graph using the path position information of the genomes represented in the graph.
+Our HOWGWILD! applying method uses the path position information of the genomes represented in the pangenome graph to move pairs of nodes in parallel.
 We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, revealing their biological features. \\
@@
 A bi-directed graph can even contain both strands of DNA as paths. \\
+
 The layout of these pangenome graphs, that is the arrangement of nodes and edges in a representation, provide an accessible representation of genetic variation across multiple genomes.
-    Graph layout methods aim to find the best node positions in the layout to minimizes overlapping nodes or edges, reduces edge crossings, and promote intuitive understanding of the data by displaying structure and important features in an understandable way.
+Graph sorting aims to find the best node order for a layout to minimizes overlapping nodes or edges, reduces edge crossings, and promote intuitive understanding of the data by displaying structure and important features in an understandable way.
 \FIXME{
@@
 	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
-    } \\
+}
+
 With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
@@
 \FIXME{A LITTLE BIT MORE ABOUT THE SGD APPROACH?}
-    We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph building and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, and maybe HPRC PAPER}).
+We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph buildiong and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, and maybe HPRC PAPER}).
 
@@
 \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
-    Our algorithm is inspired by Zheng et al. ((\FIXME{CITATION} \url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
-    In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D).
+Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is theoretically applicable in any number of dimensions.
+In this work, we describe the results in 1D and 2D.
 In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
 Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
-    This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
-    Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
+This allows the PG-SGD to be applied manuscriptto gigabase-scale pangenome graphs. \\
+Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
 \begin{itemize}
@@
 	\item The second node Xj of a pair is sampled from the same path following a Zipfian distribution. 
-        \item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
-        \item The magnitude r of the update depends on the current learning rate of the SGD.
+	\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes. The magnitude r of the update depends on the current learning rate of the SGD.
 \end{itemize}
 %The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
-    We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
-    We can have several worker threads but only one checker thread!
+We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread. We can have several worker threads but only one checker thread!
 
@@
 \iffalse
-
 \section{Results}
@@
 % something for the supplementary
-
 \paragraph{Performance evaluation}
@@
 Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
-
 \paragraph{Latent graph structure reveals underlying biology}
@@
 \input{tab_metrics_latent_graph_structure}
-
 \paragraph{Bonus Section}
@@
 \label{sec:discussion}
-    We presented
+
 We propose / implemented ....
-
 \paragraph{}
 Difference to other existing methods, are there possible improvements of our method possible?
-
 \paragraph{}
@@
 @Jiajie + Niklas: What about going GPU?
-
 \paragraph{}
 The algorithm allows to inspect a pangenome graph on base-level, as a whole.
-
 \paragraph{from the ODGI paper:}
@@
 We plan to focus on using these learned models to detect structural variation and assembly errors.
-
 \paragraph{}
@@
 Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
-
 \paragraph{}
@@
 and for other algorithms that look at and classify variation
-
 \paragraph{}
 What about the future?
-
 \paragraph{}
@@
 
+JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.
 S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
@@
 A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
-    JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.
 

### 2023-06-26 manuscript/main.tex
halfway introduction

@@
 To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-Our HOWGWILD! applying method uses the path position information of the genomes represented in the pangenome graph to move pairs of nodes in parallel.
+Our HOWGWILD! method uses the path position information of the genomes represented in the pangenome graph to move pairs of nodes in parallel.
 We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, revealing their biological features. \\
@@
 
-\FIXME{WE ONLY WANT TO TAKE ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
-PG-SGD PRODUCES 1D and 2D LAYOUTS.
+\FIXME{WE ONLY WANT TO TALK ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
+PG-SGD PRODUCES 1D AND 2D LAYOUTS.
 IN 1D WE SORT A GRAPH BY PROJECTING THE ACTUAL 1D LAYOUT INTO A NODE ORDER!}
@@
 \section{Introduction}
-Reference genomes are the most widely used resource in genetics, serving as a foundation for a variety of downstream analyses
-such as gene annotation, read mapping, and variant detection (\FIXME{XXX WE NEED CITATIONS XXX}).
-However, with the availability of thousands of high-quality genomes per species, this linear model has shown its weaknesses.
-A single genome can not faithfully represent the genetic diversity of a species, leading to the reference bias problem (\FIXME{XXX WE NEED CITATIONS XXX}).
-A pangenome models the entire set of genomic elements of a given population \citep{Tettelin_2008,cpang2018,Eizenga_2020}.
-It can be represented by a graph data structure incorporating sequences as nodes and their connections as edges ((\FIXME{XXX A FIGURE MIGHT HELP XXX}).
+Reference genomes are the most widely used resource in genetics, serving as a foundation for a variety of elemental analyses
+such as gene annotation, read mapping, and variant detection \citep{Singh2022}.
+With the availability of hundresds to thousands of high quality genomes per species, this linear model has become obsolete.
+A single genome can not faithfully represent the genetic diversity of a species, leading to the reference bias problem \citep{Ballouz2019}.
+A pangenome models the entire set of genomic elements of a given population \citep{Tettelin_2008,cpang2018,Eizenga_2020, Sherman_2020}.
+It can be represented by a sequence graph incorporating sequences as nodes and their connections as edges \citep{Hein1989}.
+%@Andrea: I don't think we need a figure which describes nodes, edges, paths.
+And how a variation graph looks like.
+Because, there already exists tons of them and we would just make another such figure.
 These nodes are shared for identical sequences, such as homologs, paralogs, and orthologs.
-In the variation graph model \citep{Garrison:2018} reads, contigs, haplotypes, or individual chromosomes are represented as paths in the graph.
-A bi-directed graph can even contain both strands of DNA as paths. \\
+In the variation graph model \citep{Garrison:2018} reads, contigs, haplotypes, or individual chromosomes are encoded as paths visiting the nodes in the graph.
+%A bidirected graph can even contain both strands of DNA as paths. \\
 
-The layout of these pangenome graphs, that is the arrangement of nodes and edges in a representation, provide an accessible representation of genetic variation across multiple genomes.
-Graph sorting aims to find the best node order for a layout to minimizes overlapping nodes or edges, reduces edge crossings, and promote intuitive understanding of the data by displaying structure and important features in an understandable way.
+A pangenome graph layout is the arrangement of nodes and edges in n-dimensional space such that a human accessible representation of genetic variation across multiple genomes can be drawn.
+The layout challenge is to find the best node positions in order to minimize overlapping nodes or edges, reduce edge crossings, and promote intuitive understanding by displaying structure and important features in a human readable way.
 \FIXME{
@@
 efficiency, scalability, readability, and the ability to reveal significant biological patterns.
-Pangenome graphs embed linear sequences as paths in the graph, but to our knowledge, no algorithm takes into account this biological information in the sorting.
-We present a new layout algorithm to simplify a pangenome graph, by using path-guided stochastic gradient descent (\FIXME{CITATION}) to move a single pair of nodes at a time.
-\FIXME{A LITTLE BIT MORE ABOUT THE SGD APPROACH?}
-We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph buildiong and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, and maybe HPRC PAPER}).
+Pangenome graphs embed linear sequences as paths in the graph, but to our knowledge, no algorithm takes into account this biological information during the layout process.
+We present a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
+ (PG-SGD) to move pairs of nodes in parallel. ( \FIXME{WE NEED TO SAY SOMETHING ABOUT HOGWILD HERE?!}). \\
+\FIXME{A LITTLE BIT MORE ABOUT THE SGD APPROACH?} \\
+We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph buildiong and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, HPRC PAPER, ODGI PAPER}). \\
+\FIXME{2D: waragraph + gfaestus for interactive visualization}.
 

### 2023-06-26 manuscript/main.tex

@@
 
-\section{Algorithm}
-
     \paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
-Our algorithm is inspired by Zheng et al. (\url{https://doi.org/hkv4}) and it is theoretically applicable in any number of dimensions.
-In this work, we describe the results in 1D and 2D.
+    \FIXME{SPECIFY THAT THE IMPLEMENTATION IS IN ODGI AND CITE IT}
+    Our algorithm is inspired by Zheng et al. ((\FIXME{CITATION} \url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
+    In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D).
     In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
     Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
-This allows the PG-SGD to be applied manuscriptto gigabase-scale pangenome graphs. \\
-Our multi-threaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
+    This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
+    Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
     \begin{itemize}
@@
         \item The second node Xj of a pair is sampled from the same path following a Zipfian distribution.
-	\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes. The magnitude r of the update depends on the current learning rate of the SGD.
+        \item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
+        \item The magnitude r of the update depends on the current learning rate of the SGD.
     \end{itemize}
 %The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
-We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread. We can have several worker threads but only one checker thread!
+    We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
+    We can have several worker threads but only one checker thread!
 
@@
 
-We propose / implemented ....
-\paragraph{}
-Difference to other existing methods, are there possible improvements of our method possible?
-\paragraph{}
-Discuss performance eval
-@Jiajie + Niklas: What about going GPU?
-\paragraph{}
-The algorithm allows to inspect a pangenome graph on base-level, as a whole.
-\paragraph{from the ODGI paper:}
-Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.
-However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-Recent interactive pangenome graph browsers are reference-centric (Beyer et al., 2019; Yokoyama et al., 2019), have a limited predefined coordinate system (Durant et al., 2021), or focus primarily on 2D representations (Gonnella et al., 2019; Wick et al., 2015).
-Our graph sorting and layout algorithms can provide the foundation for future tools of this type.
-We plan to focus on using these learned models to detect structural variation and assembly errors.
-\paragraph{}
-The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
-A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
-Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
-Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
-\paragraph{}
-I'd like to argue that this projection presents biologically relevant structures that we can work with downstream
-both for human interface
-and for other algorithms that look at and classify variation
-\paragraph{}
-What about the future?
-\paragraph{}
-How can other scientists benefit from this work?
-Graph visualization (together with annotation) is key for understanding variations.
-This projection presents biologically relevant structures that we can work with downstream for researches and algorithms that look at and classify variation,
+    We presented the Path-Guided Stochastic Gradient Descent (PG-SGD), a layout algorithm for pangenome graphs that
+    leverages the inherent biological information available within the graph's represented genomic sequences.
+    Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
+    \FIXME{Difference to other existing methods, are there possible improvements of our method possible?}
+    \FIXME{Discuss performance eval. @Jiajie + Niklas: What about going GPU?}
+    We implemented the PG-SGD algorithm to generate layouts in 1 and 2 dimensions, but the algorithm can be extended in any number of dimensions.
+    Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented 
+    high-level perspective on variation in pangenomes. These layouts are critical for pangenome graph building (\FIXME{CITE PGG and HPRC MAIN PAPER}) and 
+    making new biological discoveries (\FIXME{CITE ACROCENTRIC PAPER}).
+    %The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
+    %A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
+    %Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
+    %Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
+    However, there remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
+    Our PG-SGD algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts.
+    Futhermore, projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of 
+    algorithms that use the graph layout to look at variant classification.
+    Our future research involve exploiting these learned projections to detect structural variations and assembly errors.
+    \FIXME{PROTEIN?}
 

### 2023-06-27 manuscript/main.tex
rework introduction

@@
 
-\FIXME{WE ONLY WANT TO TALK ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
-PG-SGD PRODUCES 1D AND 2D LAYOUTS.
-IN 1D WE SORT A GRAPH BY PROJECTING THE ACTUAL 1D LAYOUT INTO A NODE ORDER!}
-
 \section{Introduction}
@@
 A pangenome graph layout is the arrangement of nodes and edges in n-dimensional space such that a human accessible representation of genetic variation across multiple genomes can be drawn. 
-The layout challenge is to find the best node positions in order to minimize overlapping nodes or edges, reduce edge crossings, and promote intuitive understanding by displaying structure and important features in a human readable way.
-\FIXME{
-	Existing layout algorithms range from simple, force-directed methods to more complex, hierarchical layouts, each presenting unique strengths and drawbacks.
-	Force-directed methods offer an aesthetic layout based on a physics simulation but may struggle with scalability for larger graphs (XXX WE NEED CITATIONS XXX).
-	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
-}
+\FIXME{We may have to talk about paths and nucleotides here. Or we introduce this later, not sure.}
+The layout challenge is to find suitable node coodinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote intuitive understanding by displaying structure and important features in a human readable way.
+One popular graph layout approach is force-directed graph drawing which produces aesthetically-pleasing layouts.
+It is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations can circumvent that \citep{Zheng2019}.
+Typically, force-directed layouts are hard to compute.
+HOGWILD!
+\citep{Recht2011} SGD presents a parallelizable solution.
+In practice, multidimensional scaling (MDS) using pairwise nodes distances as a (dis)similarity measurement is applied to minimize the energy/stress/secs of a graph.
+Since pangenome graphs embed linear sequences as paths in the graph, a reasonable distance metric would be the path nucleotide distance between a pair of nodes. 
+But to our knowledge, no algorithm takes into account this biological information during the layout process.
 
-With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
-efficiency, scalability, readability, and the ability to reveal significant biological patterns.
-Pangenome graphs embed linear sequences as paths in the graph, but to our knowledge, no algorithm takes into account this biological information during the layout process.
 We present a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
- (PG-SGD) to move pairs of nodes in parallel. ( \FIXME{WE NEED TO SAY SOMETHING ABOUT HOGWILD HERE?!}). \\
-\FIXME{A LITTLE BIT MORE ABOUT THE SGD APPROACH?} \\
-We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph buildiong and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, HPRC PAPER, ODGI PAPER}). \\
-\FIXME{2D: waragraph + gfaestus for interactive visualization}.
+(PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner. The algorithm aims to adjust the graph layout to best reflect the path nucleotide positions.
+We exemplify how our implementation is able to produce human readable layouts.
+Furthermore, the 1D PG-SGD already is a key step in pangenome graph construciton \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.
+
+\FIXME{2D: waragraph + gfaestus for interactive visualization? Not sure how to add this here. I am afraid the readers might think we also implemented these tools. Better in the discussion?}. \\
+% We may want to adjust some citations using https://tex.stackexchange.com/questions/585942/how-to-display-the-first-two-authors-in-a-citation-call-out.
+
+%\FIXME{
+%	Existing layout algorithms range from simple, force-directed methods to more complex, hierarchical layouts, each presenting unique strengths and drawbacks.
+%	Force-directed methods offer an aesthetic layout based on a physics simulation but may struggle with scalability for larger graphs (XXX WE NEED CITATIONS XXX).
+%	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
+%}
+
+%With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
+%efficiency, scalability, readability, and the ability to reveal significant biological patterns. \FIXME{MOVE THIS SOMEWHERE ELSE?}
+
+\FIXME{WE ONLY WANT TO TALK ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
+PG-SGD PRODUCES 1D AND 2D LAYOUTS.
+IN 1D WE SORT A GRAPH BY PROJECTING THE ACTUAL 1D LAYOUT INTO A NODE ORDER!}
+
+\section{Algorithm}
 
@@
         \section{Quantization}
-        \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.}
+        \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\
+        \FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y}
     \end{appendices}

### 2023-06-27 manuscript/main.tex
bretzel

@@
 
-We present a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
-(PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner. The algorithm aims to adjust the graph layout to best reflect the path nucleotide positions.
-We exemplify how our implementation is able to produce human readable layouts.
-Furthermore, the 1D PG-SGD already is a key step in pangenome graph construciton \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.
+	This paper describes a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
+	(PG-SGD) to move pairs of nodes in a HOGWILD!
+parallel manner.
+The algorithm aims to adjust the graph layout to best reflect the path nucleotide positions.
+In theory, our algorithm is applicable in any number of dimensions.
+Here,
+	we exemplify how our implementation is able to produce 1D and 2D human readable layouts.
+Already, the 1D PG-SGD is a key step in pangenome graph construction \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.
 
@@
 
-\paragraph{The HOGHWILD! path-guided stochastic gradient descent algorithm}
-\FIXME{SPECIFY THAT THE IMPLEMENTATION IS IN ODGI AND CITE IT}
-Our algorithm is inspired by Zheng et al. ((\FIXME{CITATION} \url{https://doi.org/hkv4}) and it is applicable in any number of dimensions.
-In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D).
+	In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D). In this section we will describe. Then we will. And then. \\
+
+	\FIXME{Do we need a formal definition of a variation graph like in the seqwish paper?}
+
+	Our HOGHWILD! path-guided stochastic gradient descent algorithm is inspired by \cite{Zheng2019} and is theoretically applicable in any number of dimensions. 
+	We implemented the algorithm in ODGI \citep{Guarracino2022}, making use of its already proven interface to pangenome graphs in the standard graphical fragment assebmly (GFA) format (\url{https://github.com/GFA-spec/GFA-spec}).
 	In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
@@
 	We can have several worker threads but only one checker thread!
+
+	\subsection{general algorithm}
+	\subsection{path index}
+	\subsection{parallelization in detail}
+	\subsection{randomization}
+	ZIPF!
+	\subsection{annealing step-size $\eta$}
+	\subsection{convergence / stop criteria, iterations, term updates}
+	\subsection{cooling || flip}
+	\subsection{from layout to 1D node order}
 
@@
 
-    We presented the Path-Guided Stochastic Gradient Descent (PG-SGD), a layout algorithm for pangenome graphs that
-    leverages the inherent biological information available within the graph's represented genomic sequences.
+	We have presented the PG-SGD layout algorithm for pangenome graphs that leverages the inherent biological information available within the graph's represented genomic sequences.
 	Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
@@
 	\FIXME{Discuss performance eval. @Jiajie + Niklas: What about going GPU?}
-    We implemented the PG-SGD algorithm to generate layouts in 1 and 2 dimensions, but the algorithm can be extended in any number of dimensions.
+	We implemented the PG-SGD algorithm to generate layouts in 1D and 2D, but the algorithm can in princible be extended to any number of dimensions.
 	Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented 
@@
 	making new biological discoveries (\FIXME{CITE ACROCENTRIC PAPER}).
+
+	Problem space: We take a look at all paths (FULL HPRC GRAPH ~8 BILLION STEPS ACROSS ~34 THOUSAND PATHS) and not at all nodes (FULL HPRC GRAPH ~100 MILLION).
+So we have a very large problem space compared to just looking at the nodes.
 	%The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
@@
 	%Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
+
 	However, there remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
 	Our PG-SGD algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts.
-    Futhermore, projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of 
-    algorithms that use the graph layout to look at variant classification.
+	\FIXME{2D: waragraph + gfaestus for interactive visualization?}
+	Futhermore, projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout to look at variant classification.
 	Our future research involve exploiting these learned projections to detect structural variations and assembly errors.

### 2023-06-27 manuscript/pg-sgd_pseudocode.tex
bretzel

@@
 		\for{$\eta$ $in$ $annealing$ $schedule$}{ %our "schedule" actually is the number paths.... we should specify this I would say
-			\each{$planned$ $term$ $updates$} { % I think we can remove i<j, because we don't care about that
+			\each{$planned$ $term$ $update$} { % I think we can remove i<j, because we don't care about that
 				$ps_a \gets \Unif(\mathcal{XP})$ \tcp{uniform path step from all steps}
 				$path \gets \PathId(ps_a,\mathcal{XP})$ \tcp{the path id of the sampled step}
-				\uIf{$(coolling$ $||$ $flip)$} {
+				\uIf{$(cooling$ $||$ $flip)$} {
 					$ps_b \gets \Unif(\StepCount(path, \mathcal{XP}))$ \tcp{uniform path step from specific path}
@@
 Explain $PathIndex$ in the text.
-Explain $ZipfZetas$ in the text.
+Explain $InitZipf$ in the text.
 Explain $RandomLayout$ in the text.

### 2023-06-28 manuscript/main.tex
more compact and less jargony abstract

@@
 	\textbf{Motivation:}
-	The increasing availability of high quality genomes demands models that allow for the study of genomic variability within entire populations.
+	The increasing availability of high quality genomes demands models to study genomic variability within entire populations.
 	Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
@@
 	\textbf{Results:}
-	To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-	Our HOWGWILD! method uses the path position information of the genomes represented in the pangenome graph to move pairs of nodes in parallel.
-	We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, revealing their biological features. \\
+	To address these challenges, we propose a novel graph layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
+	Our HOWGWILD! method uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel.
+	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, revealing their biological features. \\
 	\textbf{Availability:}
 	The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
-	Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
+	Source code is available at \url{https://github.com/pangenome/odgi}. \\ % and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
 	\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\

### 2023-06-28 manuscript/main.tex
intro

@@
 	\section{Introduction}
-	Reference genomes are the most widely used resource in genetics, serving as a foundation for a variety of elemental analyses
+	Reference genomes are the most widely used resource in genetics, serving as a foundation for a variety of analyses
 	such as gene annotation, read mapping, and variant detection \citep{Singh2022}.
-	With the availability of hundresds to thousands of high quality genomes per species, this linear model has become obsolete.
+	However, with the availability of hundreds to thousands of high quality genomes per species, this linear model has become obsolete.
 	A single genome can not faithfully represent the genetic diversity of a species, leading to the reference bias problem \citep{Ballouz2019}.
-	A pangenome models the entire set of genomic elements of a given population \citep{Tettelin_2008,cpang2018,Eizenga_2020, Sherman_2020}.
-	It can be represented by a sequence graph incorporating sequences as nodes and their connections as edges \citep{Hein1989}.
-	%@Andrea: I don't think we need a figure which describes nodes, edges, paths.
-And how a variation graph looks like.
-Because, there already exists tons of them and we would just make another such figure.
+	Rather, a pangenome models the entire set of genomic elements of a given population \citep{Tettelin_2008,cpang2018,Eizenga_2020, Sherman_2020}.
+	Pangenomes can be represented by a sequence graph incorporating sequences as nodes and their connections as edges \citep{Hein1989}.
 	These nodes are shared for identical sequences, such as homologs, paralogs, and orthologs.
-	In the variation graph model \citep{Garrison:2018} reads, contigs, haplotypes, or individual chromosomes are encoded as paths visiting the nodes in the graph.
+	In the variation graph model \citep{Garrison:2018}, genomes are encoded as paths visiting the nodes in the graph.
 	%A bidirected graph can even contain both strands of DNA as paths. \\
 
-	A pangenome graph layout is the arrangement of nodes and edges in n-dimensional space such that a human accessible representation of genetic variation across multiple genomes can be drawn. 
-	\FIXME{We may have to talk about paths and nucleotides here. Or we introduce this later, not sure.}
-	The layout challenge is to find suitable node coodinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote intuitive understanding by displaying structure and important features in a human readable way.
-	One popular graph layout approach is force-directed graph drawing which produces aesthetically-pleasing layouts.
-It is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations can circumvent that \citep{Zheng2019}.
-Typically, force-directed layouts are hard to compute.
-HOGWILD!
-\citep{Recht2011} SGD presents a parallelizable solution.
-In practice, multidimensional scaling (MDS) using pairwise nodes distances as a (dis)similarity measurement is applied to minimize the energy/stress/secs of a graph.
-	Since pangenome graphs embed linear sequences as paths in the graph, a reasonable distance metric would be the path nucleotide distance between a pair of nodes. 
-	But to our knowledge, no algorithm takes into account this biological information during the layout process.
-
-	This paper describes a new pangenome graph layout algorithm which uses path-guided stochastic gradient descent %\citep{Zheng2019}
-	(PG-SGD) to move pairs of nodes in a HOGWILD!
-parallel manner.
-The algorithm aims to adjust the graph layout to best reflect the path nucleotide positions.
-In theory, our algorithm is applicable in any number of dimensions.
-Here,
-	we exemplify how our implementation is able to produce 1D and 2D human readable layouts.
-Already, the 1D PG-SGD is a key step in pangenome graph construction \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.
-
-	\FIXME{2D: waragraph + gfaestus for interactive visualization? Not sure how to add this here. I am afraid the readers might think we also implemented these tools. Better in the discussion?}. \\
-	% We may want to adjust some citations using https://tex.stackexchange.com/questions/585942/how-to-display-the-first-two-authors-in-a-citation-call-out.
-
+	A pangenome graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space such that a human accessible representation of genetic variation across multiple genomes can be drawn.
+	Layout algorithms aim to find suitable node coodinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote intuitive understanding by displaying the graph layout.
+	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetically-pleasing layouts.
+	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations can alleviate such a problem \citep{Zheng2019}.
+	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the HOGWILD! SGD is a parallelizable, therefore scalable solution \citep{Recht2011}.
+	In practice, multidimensional scaling (MDS) is applied to minimize the difference between the visual distance and theoretical graphed distance, and this is accomplished by using pairwise nodes distances to minimize an energy function.
 	%\FIXME{
@@
 	%}
+	Since pangenome graphs represent genomes as paths in the graph, a reasonable distance metric would be the nucleotide distance between a pair of nodes traversed by the same path.
+	But to our knowledge, no algorithm takes into account such biological information to compute the pangenome graph layouts.
 
-	%With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
-	%efficiency, scalability, readability, and the ability to reveal significant biological patterns. \FIXME{MOVE THIS SOMEWHERE ELSE?}
+	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner.
+	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
+	Our implementation produces layouts in 1 dimension (1D) and 2 dimensions (2D), but it can be extended in any number of dimensions.
 
+	% We may want to adjust some citations using https://tex.stackexchange.com/questions/585942/how-to-display-the-first-two-authors-in-a-citation-call-out.
+
+	\section{Algorithm}
 	\FIXME{WE ONLY WANT TO TALK ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
@@
 IN 1D WE SORT A GRAPH BY PROJECTING THE ACTUAL 1D LAYOUT INTO A NODE ORDER!}
-
-	\section{Algorithm}
 
@@
 
+	Already, the 1D PG-SGD is a key step in pangenome graph construction \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.
+	\FIXME{2D: waragraph + gfaestus for interactive visualization? Not sure how to add this here. I am afraid the readers might think we also implemented these tools. Better in the discussion?}. \\
+
 	\section*{Acknowledgments}

### 2023-06-28 manuscript/main.tex
variation graph definition

@@
 	\section{Algorithm}
-	\FIXME{WE ONLY WANT TO TALK ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
+	%\FIXME{WE ONLY WANT TO TALK ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
 PG-SGD PRODUCES 1D AND 2D LAYOUTS.
@@
 
-	In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D). In this section we will describe. Then we will. And then. \\
+	%In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D). In this section we will describe. Then we will. And then. \\
 
-	\FIXME{Do we need a formal definition of a variation graph like in the seqwish paper?}
+\begin{definition}
+	Variation graphs are a mathematical formalism to represent pangenome graphs \citep{Garrison_2019_thesis}.
+	In the variation graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{P})$, nodes (or vertices) $\mathcal{V} = v_1\ldots v_{|\mathcal{V}|}$ contain sequences of DNA.
+	Each node $v_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{v_i}$.
+	The node strand $s$ represents node orientation.
+	Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands ($e_i = ( s_a, s_b )$), defining the graph topology.
+	Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ describe walks over node strands ($p_i = s_1 \ldots s_{|p_i|}$), representing the genomes embedded in the graph.
+\end{definition}
 

### 2023-06-28 manuscript/pg-sgd_pseudocode.tex
more concise and formal pseudocode

@@
 \SetKwFunction{PathIndex}{PathIndex}
-\SetKwFunction{RandomLayout}{RandomLayout}
+\SetKwFunction{LayoutInitialization}{LayoutInitialization}
 \SetKwFunction{InitZipf}{InitZipf}
@@
 \SetKwFunction{StepCount}{StepCount}
-\SetKwFunction{PathId}{PathId}
+\SetKwFunction{Path}{Path}
 \SetKwFunction{StepPos}{StepPos}
@@
 \begin{algorithm}
-	\pgsgd{\textbf{(}$\mathcal{V}$\textbf{)}}{
-		\textbf{input:} variation graph $\mathcal{V} = (\mathcal{N}, \mathcal{E}, \mathcal{P})$ \\
+	\pgsgd{\textbf{(}$\mathcal{G}$\textbf{)}}{
+		\textbf{input:} variation graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{P})$ \\
 		%https://tex.stackexchange.com/questions/22643/how-to-write-letters-in-bold-in-the-math-mode
-		\textbf{output:} $k$-dimensional layout $\mathcal{L}$ with $n$ nodes \\
-		$\mathcal{XP}$ $\gets \PathIndex(\mathcal{V})$ \tcp{index for efficient path position lookup}
+		\textbf{output:} $N$-dimensional layout $\mathcal{L}$ with $|\mathcal{V}|$ nodes \\
+		$\mathcal{XP}$ $\gets \PathIndex(\mathcal{G})$ \tcp{for path position lookup} 
 		%\boldsymbol{$Z$} $\gets ZipfZetas(G,P)$ \\ %@Andrea is this correct like this?
-		$\mathcal{L}$ $\gets \RandomLayout(n, k)$ \tcp{only 2D PG-SGD has a random layout initialization, 1D starts with the current node order}
-		$\mathcal{Z} \gets \InitZipf(\mathcal{V},\mathcal{XP})$ \tcp{generate Zipfian distribution}
+		$\mathcal{L}$ $\gets \LayoutInitialization(\mathcal{V}, N)$ \\ %\tcp{random or deterministic} 
+		$\mathcal{Z} \gets \InitZipf(\mathcal{G},\mathcal{XP})$ \tcp{Zipfian distribution}
 		% atomic positions initialization?
-		% TODO for somplicity reasons I would only describe the 2D one here
+		% TODO for simplicity reasons I would only describe the 2D one here
 		\for{$\eta$ $in$ $annealing$ $schedule$}{ %our "schedule" actually is the number paths.... we should specify this I would say
 			\each{$planned$ $term$ $update$} { % I think we can remove i<j, because we don't care about that
-				$ps_a \gets \Unif(\mathcal{XP})$ \tcp{uniform path step from all steps}
-				$path \gets \PathId(ps_a,\mathcal{XP})$ \tcp{the path id of the sampled step}
+				$s_a \gets \Unif(\mathcal{XP})$ \tcp{uniformly sample a step from $\mathcal{P}$}
+				$p \gets \Path(s_a,\mathcal{XP})$ \tcp{path of $s_a$}
 				\uIf{$(cooling$ $||$ $flip)$} {
-					$ps_b \gets \Unif(\StepCount(path, \mathcal{XP}))$ \tcp{uniform path step from specific path}
-					$p_b \gets \StepPos(ps_b)$ \tcp{nucleotide position of step}
+					$s_b \gets \Unif(\StepCount(p, \mathcal{XP}))$ \tcp{uniformly sample a step from $p$}
+					$p_b \gets \StepPos(s_b)$ \tcp{nuc. position}
 				} \Else {
-					$ps_b \gets \Zipf(path)$ \tcp{Zipfian path step from specific path}
+					$s_b \gets \Zipf(p)$ \tcp{Zipfian sampling of a step from $p$}
 				}
-				$p_a \gets \StepPos(ps_a)$ \tcp{nucleotide position of step}
-				$p_b \gets \StepPos(ps_b)$ \tcp{nucleotide position of step}
-				$nd \gets ||p_a - p_b||$ \tcp{calculate nucleotide distance} %mag is nx
-				$ld \gets ||\mathcal{L}_{a}-\mathcal{L}_b||$ \tcp{calculate layout distance}
+				$p_a \gets \StepPos(s_a)$ \tcp{nuc. position}
+				$p_b \gets \StepPos(s_b)$ \tcp{nuc. position}
+				$nd \gets ||p_a - p_b||$ \tcp{nuc. distance} %mag is nx
+				$ld \gets ||\mathcal{L}_{a}-\mathcal{L}_b||$ \tcp{layout distance}
 				$w_{ab} \gets \frac{1.0}{nd}$ \tcp{term weight} 
@@
 				\uIf{$\delta <= 0$} {
-					$STOP$ \tcp{we can't optimize it any better, so we stop here}
+					$STOP$ \tcp{we can't optimize it more}
 				}
@@
 				$r \gets \delta - ld$ \tcp{size of the term update}
-				$\mathcal{L}_a \gets \mathcal{L}_a - r\cdot ld$ \tcp{update the 1. term}
-				$\mathcal{L}_b \gets \mathcal{L}_b - r\cdot ld$ \tcp{update the 2. term}
+				$\mathcal{L}_a \gets \mathcal{L}_a - r\cdot ld$ \tcp{update the 1st node}
+				$\mathcal{L}_b \gets \mathcal{L}_b - r\cdot ld$ \tcp{update the 2nd node}
 			}
@@
 	\caption{Pseudocode of the PG-SGD algorithm.
-Explain $\eta$.
+\FIXME{Explain $\eta$.
 Explain $PathIndex$ in the text.
@@
 Explain $RandomLayout$ in the text.
-Explain annealing schedule in the text or rephrase this wording.}
+Explain annealing schedule in the text or rephrase this wording.}}
 	\label{alg:pg_sgd}

### 2023-06-28 manuscript/main.tex
more paragraph 2

@@
 	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
-	Our implementation produces layouts in 1 dimension (1D) and 2 dimensions (2D), but it can be extended in any number of dimensions.
+	It can be extended in any number of dimensions. We provide in the ODGI toolkit \citep{Guarracino2022} implementations to produce layouts in 1 dimension (1D) and 2 dimensions (2D).
 
@@
 
-	%In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D). In this section we will describe. Then we will. And then. \\
+	The PG-SGD algorithm is inspired by \cite{Zheng2019}, but we designed it to work on pangenomes in the form of variation graph (\ref{def:vg}).
 
 \begin{definition}
+	\label{def:vg}
 	Variation graphs are a mathematical formalism to represent pangenome graphs \citep{Garrison_2019_thesis}.
-	In the variation graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{P})$, nodes (or vertices) $\mathcal{V} = v_1\ldots v_{|\mathcal{V}|}$ contain sequences of DNA.
+	In the variation graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{P})$, nodes (or vertices) $\mathcal{V} = v_1\ldots v_{|\mathcal{V}|}$ contain nucleotide sequences.
 	Each node $v_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{v_i}$.
@@
 
-	Our HOGHWILD! path-guided stochastic gradient descent algorithm is inspired by \cite{Zheng2019} and is theoretically applicable in any number of dimensions. 
-	We implemented the algorithm in ODGI \citep{Guarracino2022}, making use of its already proven interface to pangenome graphs in the standard graphical fragment assebmly (GFA) format (\url{https://github.com/GFA-spec/GFA-spec}).
-	In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
-	Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
-	This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
-	Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
+	The algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the nucleotide distance of the same nodes as calculated along a path that traverses them.
+	Our path index enables efficient retrieval of path positions to quickly compute nucleotide distances.
+	This eliminates the need to store all pairwise distances between nodes in memory, allowing the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
+	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
+	Our implementation is multithreaded, but for the sake of simplicity, we describe how the algorithm works on a single thread (~\ref{alg:pg_sgd}).
+
+	\subsection{General algorithm}
+
+	\FIXME{
 		\begin{itemize}
@@
 		\end{itemize}
+	}
 	%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
-	We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
-	We can have several worker threads but only one checker thread!
 
-	\subsection{general algorithm}
-	\subsection{path index}
-	\subsection{parallelization in detail}
-	\subsection{randomization}
-	ZIPF!
-	\subsection{annealing step-size $\eta$}
-	\subsection{convergence / stop criteria, iterations, term updates}
-	\subsection{cooling || flip}
-	\subsection{from layout to 1D node order}
+	\subsection{Path index}
+	%\subsection{parallelization in detail}
+	\subsection{Layoit initialization}
+	\subsection{Zipfian distribution}
+	\subsection{Annealing}
+	\subsection{Stop criteria, iterations, term updates}
+	\subsection{Cooling and flip}
+	\subsection{From 1D layout to 1D node order}
 

### 2023-06-28 manuscript/main.tex
discussion and huge gift

@@
 	}
-	%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
 
 	\subsection{Path index}
-	%\subsection{parallelization in detail}
-	\subsection{Layoit initialization}
+	\FIXME{Describe in details}
+	\subsection{Parallelization}
+	\FIXME{Be brief and specify that it works because the pangenome graphs are sparse}
+	\subsection{Layout initialization}
+	\FIXME{1D: random, node order in the input graph; 2D: Random (uniform or gaussian), hilbert}
 	\subsection{Zipfian distribution}
+	\FIXME{Describe what it is and why we use it (SPOILER: it just works)}
 	\subsection{Annealing}
+	\FIXME{Describe the problem (noise aat local level) and how we solve it by 'annealing', that is slowing down the movements}
 	\subsection{Stop criteria, iterations, term updates}
+	\FIXME{Describe briefly}
 	\subsection{Cooling and flip}
+	\FIXME{Describe briefly}
 	\subsection{From 1D layout to 1D node order}
+	\FIXME{Not sure, so very short}
 
@@
 
-    \iffalse
-
     \section{Results}
@@
 
-%\subsection{Experimental insights}
+    \subsection{Performance}
+	\FIXME{Describe the dataset (HPRC chromosomes), runtime and memory usage (specifying the faster/slower chromosome pangenome graph layout and the thinner/fatter in memory), just text no with a table}
+    \FIXME{A comparison with Bandage might be worth, to give readers an idea of the scalability}
 
-%\subsection{Limitations}
+    \subsection{Pangenome graphs layouts reveal biology features}
+	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.
+	In the supp a huge picture with a "karyotype might be hot}
 
-% something for the supplementary
-
-    \paragraph{Performance evaluation}
+	\iffalse
 	Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
@@
 
-	We have presented the PG-SGD layout algorithm for pangenome graphs that leverages the inherent biological information available within the graph's represented genomic sequences.
+	We presented the Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
 	Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
@@
 	\FIXME{Discuss performance eval. @Jiajie + Niklas: What about going GPU?}
-	We implemented the PG-SGD algorithm to generate layouts in 1D and 2D, but the algorithm can in princible be extended to any number of dimensions.
-	Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented 
-	high-level perspective on variation in pangenomes. These layouts are critical for pangenome graph building (\FIXME{CITE PGG and HPRC MAIN PAPER}) and 
-	making new biological discoveries (\FIXME{CITE ACROCENTRIC PAPER}).
 
-	Problem space: We take a look at all paths (FULL HPRC GRAPH ~8 BILLION STEPS ACROSS ~34 THOUSAND PATHS) and not at all nodes (FULL HPRC GRAPH ~100 MILLION).
-So we have a very large problem space compared to just looking at the nodes.
+	Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented high-level perspective on variation in pangenomes.
+	We implemented the PG-SGD algorithm to generate layouts in 1 and 2 dimensions.
+	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of acrocentric chromosomes \citep{Guarracino2023}.
+	Furthermore, they are applied in the creation and analysis of pangenome graphs for any species \citep{Guarracino2022, Garrison2023}.
 	%The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
@@
 	%Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
-	%Already, the 1D PG-SGD is a key step in the PanGenome Graph Building (PGGB) pipeline that we successfully applied to build the first draft human pangenome reference (Liao et al., bioRxiv 2022).
+	%Problem space: We take a look at all paths (FULL HPRC GRAPH ~8 BILLION STEPS ACROSS ~34 THOUSAND PATHS) and not at all nodes (FULL HPRC GRAPH ~100 MILLION).
+So we have a very large problem space compared to just looking at the nodes.
+	There still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
+	Our algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts (\href{https://github.com/chfi/waragraph}{https://github.com/chfi/waragraph}).
 
-	However, there remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
-	Our PG-SGD algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts.
-	\FIXME{2D: waragraph + gfaestus for interactive visualization?}
-	Futhermore, projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout to look at variant classification.
-	Our future research involve exploiting these learned projections to detect structural variations and assembly errors.
-	\FIXME{PROTEIN?}
-
-	Already, the 1D PG-SGD is a key step in pangenome graph construction \citep{Liao2023, Garrison2023} and analysis \citep{Guarracino2022, Guarracino2023}.
-	\FIXME{2D: waragraph + gfaestus for interactive visualization? Not sure how to add this here. I am afraid the readers might think we also implemented these tools. Better in the discussion?}. \\
+	The PG-SGD algorithm can be extended to any number of dimensions.
+	Projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout for variant detection and classification.
+	Our future research involves leveraging these graph projections to detect structural variants and detect and solve assembly errors.
+	Moreover, we consider the extension of the algorithm to RNA and protein sequences.
 

### 2023-06-29 manuscript/main.tex
Algorithm + Implementation

@@
 \usepackage{appendix}
+\newcommand{\description}{}
+\usepackage[inline]{enumitem}
 
@@
 	The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
-	Source code is available at \url{https://github.com/pangenome/odgi}. \\ % and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
+	Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
 	\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
@@
 	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the HOGWILD! SGD is a parallelizable, therefore scalable solution \citep{Recht2011}.
-	In practice, multidimensional scaling (MDS) is applied to minimize the difference between the visual distance and theoretical graphed distance, and this is accomplished by using pairwise nodes distances to minimize an energy function.
+	In practice, multidimensional scaling (MDS) is applied to minimize the difference between the visual distance and theoretical graph distance.
+This can be accomplished by using pairwise node	 distances to minimize an energy function.
 	%\FIXME{
@@
 	It can be extended in any number of dimensions. We provide in the ODGI toolkit \citep{Guarracino2022} implementations to produce layouts in 1 dimension (1D) and 2 dimensions (2D).
-
 	% We may want to adjust some citations using https://tex.stackexchange.com/questions/585942/how-to-display-the-first-two-authors-in-a-citation-call-out.
@@
 
-	The PG-SGD algorithm is inspired by \cite{Zheng2019}, but we designed it to work on pangenomes in the form of variation graph (\ref{def:vg}).
+	The PG-SGD algorithm is inspired by \cite{Zheng2019}, but we designed it to work on pangenomes in the form of a variation graph introduced in Definition \ref{def:vg}.
 
@@
 
-	The algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the nucleotide distance of the same nodes as calculated along a path that traverses them.
-	Our path index enables efficient retrieval of path positions to quickly compute nucleotide distances.
-	This eliminates the need to store all pairwise distances between nodes in memory, allowing the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
+	In brief, the algorithm moves a single pair of nodes $\mathcal{V}_i$, $\mathcal{V}_j$ at a time, minimizing the disparity between the layout distance $ld_{ij}$ of a node pair and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
 	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
-	Our implementation is multithreaded, but for the sake of simplicity, we describe how the algorithm works on a single thread (~\ref{alg:pg_sgd}).
-
-	\subsection{General algorithm}
-
-	\FIXME{
-		\begin{itemize}
-			\item The first node Xi of a pair is a uniform path step pick from all nodes.
-			\item The second node Xj of a pair is sampled from the same path following a Zipfian distribution.
-			\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
-			\item The magnitude r of the update depends on the current learning rate of the SGD.
-		\end{itemize}
-	}
-
-	\subsection{Path index}
-	\FIXME{Describe in details}
-	\subsection{Parallelization}
-	\FIXME{Be brief and specify that it works because the pangenome graphs are sparse}
-	\subsection{Layout initialization}
-	\FIXME{1D: random, node order in the input graph; 2D: Random (uniform or gaussian), hilbert}
-	\subsection{Zipfian distribution}
-	\FIXME{Describe what it is and why we use it (SPOILER: it just works)}
-	\subsection{Annealing}
-	\FIXME{Describe the problem (noise aat local level) and how we solve it by 'annealing', that is slowing down the movements}
-	\subsection{Stop criteria, iterations, term updates}
-	\FIXME{Describe briefly}
-	\subsection{Cooling and flip}
-	\FIXME{Describe briefly}
-	\subsection{From 1D layout to 1D node order}
-	\FIXME{Not sure, so very short}
+	\begin{enumerate*}[label=(\roman*)]
+		\item The first node $\mathcal{V}_i$ of a pair is a uniform path step pick from all nodes.
+		\item The second node $\mathcal{V}_j$ of a pair is sampled from the same path following a Zipfian distribution \citep{Zipf1932}.
+		\item The path nucleotide distance  $nd_{ij}$ guides the actual layout distance $ld_{ij}$ update of these nodes. Also referred to as \textit{term update}.
+		\item The magnitude $r$ of the update depends on the current learning rate $\mu$ of the SGD.
+	\end{enumerate*}
+	These key steps are sketched in Figure \FIXME{Add puritan figure.}. Pseudocode is shown in Algorithm~\ref{alg:pg_sgd}.
 
 	\input{pg-sgd_pseudocode}
+
+	The number of iterations steer the annealing step size $\eta$ which in turn determines the learning rate $\mu$ of the PG-SGD. 
+	Since variation graphs are very sparse, a large step size in the beginning of the schedule already leads to a globally linear \FIXME{(or planar?)} layout.
+	With decreasing $\eta$ the node layout adjustments become more local, ensuring that the nodes are positioned in away to best reflect the nucleotde distances of the paths down to the single base pair level.
+
+	In standard SGD layout algorithms, the problem space is given by pairwise node comparisons.
+In the PG-SGD, the problem space is given by the pairwise path step comparisons.
+The number of path steps in large variation graphs can be several orders of magnitude higher than the number of nodes.
+Therefore, the number of term updates in the PG-SGD is bound to the number of path steps in the graph.
+	%Should the delta of $nd_{ij}$ and $ld_{ij}$ become 0 before the annealing schedule is finished, the algorithm stops.
+
+	Originating from empirical inspection of word frequency tables, Zipf's law states that a word with rank $n$ occurs $1/n$ times as the most frequent one.
+	It's mathematical formulation is called the Zipfian distribution whichs rank-frequency distribution is related to the inverse power law.
+	Sampling from a Zipfian distribution generated with a path's nucleotide position space increases the possibility to draw a small nucleotide position.
+	So there is a high chance of a small nucleotide distance between node pair $\mathcal{V}_ij$ to refine the layout down to the nodes containing only one base pair.
+
+	To provide balance between global and local term updates, there is a 50\% chance for a \textit{flip}: In half of our term updates the second path step is sampled from a uniform distribution instead of a Zipfian one.
+Additionally, after half the number of iterations, a \textit{cooling} phase skewes the Zipfian distribution so that the possibility of a a small sampled nucleotide distance is significantly increased.
+Again, this boosts the local planarity of the graph layout.
+
+	\section{Implementation}
+
+	We implemented the algorithm in ODGI which serves as an interface to graphs in the GFA (\url{https://github.com/GFA-spec/GFA-spec}) format. 
+	Our path index is a strict subset of the XG index~\citep{Garrison:2018} with the exception that
+	we specifically avoided to use succinct SDSL~\citep{Gog2014} data structures. We only use bit compressed integer vectors. 
+	This enables efficient retrieval of path positions to quickly compute nucleotide distances and eliminates the need to store all pairwise distances between nodes in memory, allowing the PG-SGD to be applied to gigabase-scale pangenome graphs.
+
+	Our implementation is multithreaded, threads can perform term update calculations in parallel.
+Layout values are stored in an atomic integer vector presenting the only multithreading bottleneck.
+Effectively, this presents a modified version of the HOGWILD!
+SGD algorithm which just writes the layout updates to a shared non-atomic integer vector.
+We went for an atomic vector, because this ensures memory saftey and in our experience there was basically no performance loss.
+Algorithm~\ref{alg:pg_sgd}.
+
+	Due to the fact that we have a very high learning rate, our 1D implementation can work with a random layout as input.
+Though, by default, it just makes the given node order the starting layout.
+In 2D, we offer several random layout initializations:
+		\begin{enumerate*}[label=(\roman*)]
+		\item Uniform,
+		\item gaussian,
+		\item or hilbert.
+	\end{enumerate*}
+	\FIXME{1D: random, node order in the input graph; 2D: Random (uniform or gaussian), hilbert}
 
@@
 	Projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout for variant detection and classification.
+	Helpful could be to integrate more layout objectives~\FIXME{Add citation for \url{https://arxiv.org/pdf/2112.01571.pdf}} like neighborhood preservation.
+
 	Our future research involves leveraging these graph projections to detect structural variants and detect and solve assembly errors.
@@
 
+	\section{Animations}
+	\FIXME{ADD LINKS TO OUR ANIMATIONS ON ZENODO}
+
 	\bibliographystyle{natbib}
@@
 	\begin{appendices}
-	    \section{Quantization}
+	    \section{Appendix}
 	    \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\
-	    \FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y}
+	    \FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y} \\
+	    \FIXME{From 1D layout to 1D node order} \\
+	    \FIXME{reference-guided path-guided stochastic gradient descent}
 	\end{appendices}

### 2023-06-29 manuscript/pg-sgd_pseudocode.tex
Algorithm + Implementation

@@
 					$s_b \gets \Unif(\StepCount(p, \mathcal{XP}))$ \tcp{uniformly sample a step from $p$}
-					$p_b \gets \StepPos(s_b)$ \tcp{nuc. position}
 				} \Else {
@@
 				\uIf{$\delta <= 0$} {
-					$STOP$ \tcp{we can't optimize it more}
+					$STOP$ \tcp{we can't optimize more}
 				}
@@
 	}
-	\caption{Pseudocode of the PG-SGD algorithm.
-\FIXME{Explain $\eta$.
-Explain $PathIndex$ in the text.
-Explain $InitZipf$ in the text.
-Explain $RandomLayout$ in the text.
-Explain annealing schedule in the text or rephrase this wording.}}
+	\caption{Pseudocode of the PG-SGD algorithm.}
 	\label{alg:pg_sgd}

### 2023-06-30 manuscript/main.tex
remove FIXME
titles
add citation and tiny tweaks

@@
 
-    \title[Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
+    %\title[Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
+    %\title[Pangenome graph layout by Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by Path-Guided Stochastic Gradient Descent}
+    %\title[Unveiling genome variation by sequence-guided pangenome graph layouts]{Unveiling genome variation by sequence-guided pangenome graph layouts}
+    %\title[PG-SGD: sequence-guided pangenome graph layouts]{PG-SGD: sequence-guided pangenome graph layouts}
+    %\title[Genome-guided pangenome graph layouts via Stochastic Gradient Descent]{Genome-guided pangenome graph layouts via Stochastic Gradient Descent}
+	%\title[Sequence-guided pangenome graph layouts via Stochastic Gradient Descent]{Sequence-guided pangenome graph layouts via Stochastic Gradient Descent}
+    %\title[Genome-guided pangenome graph layouts]{Genome-guided pangenome graph layouts}
+	\title[Sequence-guided pangenome graph layouts]{Sequence-guided pangenome graph layouts}
+
 	\author[Heumos, Guarracino \textit{et~al}.]{
@@
 	\textbf{Motivation:}
-	The increasing availability of high quality genomes demands models to study genomic variability within entire populations.
+	The increasing availability of complete genomes demands models to study genomic variability within entire populations.
 	Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
@@
 We went for an atomic vector, because this ensures memory saftey and in our experience there was basically no performance loss.
-Algorithm~\ref{alg:pg_sgd}.
 
@@
 		\item gaussian,
-		\item or hilbert.
+		\item or Hilbert.
 	\end{enumerate*}
-	\FIXME{1D: random, node order in the input graph; 2D: Random (uniform or gaussian), hilbert}
 
@@
     \subsection{Performance}
-	\FIXME{Describe the dataset (HPRC chromosomes), runtime and memory usage (specifying the faster/slower chromosome pangenome graph layout and the thinner/fatter in memory), just text no with a table}
+	\FIXME{Describe the dataset (HPRC chromosomes), runtime and memory usage (specifying the fastest/slowest chromosome pangenome graph layout and the thinnest/fattest in memory), just text no with a table}
     \FIXME{A comparison with Bandage might be worth, to give readers an idea of the scalability}
@@
 	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.
-	In the supp a huge picture with a "karyotype might be hot}
+	In the supp a huge picture with a "karyotype", that is all chromosome layout together sorted by chromosome number, might be hot}
 
@@
 	Projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout for variant detection and classification.
-	Helpful could be to integrate more layout objectives~\FIXME{Add citation for \url{https://arxiv.org/pdf/2112.01571.pdf}} like neighborhood preservation.
+	Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.
 
@@
 	    \section{Appendix}
-	    \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\
-	    \FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y} \\
+	    %\FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\
+	    %\FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y} \\
 	    \FIXME{From 1D layout to 1D node order} \\

### 2023-07-01 manuscript/main.tex
We follow the HOWGWILD! strategy; PG-SGD as a name

@@
 	To address these challenges, we propose a novel graph layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-	Our HOWGWILD! method uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel.
+	PG-SGD uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel by applying the HOWGWILD! strategy.
 	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, revealing their biological features. \\
 	\textbf{Availability:}
-	The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
-	Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
+	PG-SGD has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
+	Source code is available at \url{https://github.com/pangenome/odgi}. \\% and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
 	\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
@@
 
-	The PG-SGD algorithm is inspired by \cite{Zheng2019}, but we designed it to work on pangenomes in the form of a variation graph introduced in Definition \ref{def:vg}.
+	PG-SGD is inspired by \cite{Zheng2019}, but we designed our algorithm to work on pangenomes in the form of a variation graph introduced in Definition \ref{def:vg}.
 
@@
 
-	The number of iterations steer the annealing step size $\eta$ which in turn determines the learning rate $\mu$ of the PG-SGD. 
+	The number of iterations steer the annealing step size $\eta$ which in turn determines the learning rate $\mu$ of PG-SGD. 
 	Since variation graphs are very sparse, a large step size in the beginning of the schedule already leads to a globally linear \FIXME{(or planar?)} layout.
@@
 	In standard SGD layout algorithms, the problem space is given by pairwise node comparisons.
-In the PG-SGD, the problem space is given by the pairwise path step comparisons.
+In PG-SGD, the problem space is given by the pairwise path step comparisons.
 The number of path steps in large variation graphs can be several orders of magnitude higher than the number of nodes.
-Therefore, the number of term updates in the PG-SGD is bound to the number of path steps in the graph.
+Therefore, the number of term updates in PG-SGD is bound to the number of path steps in the graph.
 	%Should the delta of $nd_{ij}$ and $ld_{ij}$ become 0 before the annealing schedule is finished, the algorithm stops.
@@
 	we specifically avoided to use succinct SDSL~\citep{Gog2014} data structures. We only use bit compressed integer vectors. 
-	This enables efficient retrieval of path positions to quickly compute nucleotide distances and eliminates the need to store all pairwise distances between nodes in memory, allowing the PG-SGD to be applied to gigabase-scale pangenome graphs.
+	This enables efficient retrieval of path positions to quickly compute nucleotide distances and eliminates the need to store all pairwise distances between nodes in memory, allowing PG-SGD to be applied to gigabase-scale pangenome graphs.
 
@@
     \input{tab_performance.tex}
-    Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
+    Already, PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
 
@@
 
-	We presented the Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
+	We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
 	Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
@@
 
-	Graph visualization is key for understanding genome variations and the large-scale layouts produced by the PG-SGD algorithm offer an unprecedented high-level perspective on variation in pangenomes.
-	We implemented the PG-SGD algorithm to generate layouts in 1 and 2 dimensions.
+	Graph visualization is key for understanding genome variations and the large-scale layouts produced by PG-SGD offer an unprecedented high-level perspective on variation in pangenomes.
+	We implemented PG-SGD to generate layouts in 1 and 2 dimensions.
 	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of acrocentric chromosomes \citep{Guarracino2023}.
@@
 
-	The PG-SGD algorithm can be extended to any number of dimensions.
+	PG-SGD can be extended to any number of dimensions.
 	Projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout for variant detection and classification.

### 2023-07-01 manuscript/main.tex
be picky

@@
 	The increasing availability of complete genomes demands models to study genomic variability within entire populations.
-	Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
+	Pangenome graphs can represent the full genetic diversity between multiple genomes, but their layouts may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
 	These structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
@@
 	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations can alleviate such a problem \citep{Zheng2019}.
-	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the HOGWILD! SGD is a parallelizable, therefore scalable solution \citep{Recht2011}.
+	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the HOGWILD!
+SGD is a highly parallelizable, therefore scalable solution that can be applied when the optimization problem is sparse \citep{Recht2011}.
 	In practice, multidimensional scaling (MDS) is applied to minimize the difference between the visual distance and theoretical graph distance.
@@
 	Since pangenome graphs represent genomes as paths in the graph, a reasonable distance metric would be the nucleotide distance between a pair of nodes traversed by the same path.
-	But to our knowledge, no algorithm takes into account such biological information to compute the pangenome graph layouts.
+	To our knowledge, no algorithm takes into account such biological information to compute the pangenome graph layouts.
 
-	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner.
+	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in parallel with a HOGWILD! strategy.
 	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
-	It can be extended in any number of dimensions. We provide in the ODGI toolkit \citep{Guarracino2022} implementations to produce layouts in 1 dimension (1D) and 2 dimensions (2D).
+	PG-SGD can be extended in any number of dimensions. We provide in the ODGI toolkit \citep{Guarracino2022} implementations to produce layouts in 1 dimension (1D) and 2 dimensions (2D).
 	% We may want to adjust some citations using https://tex.stackexchange.com/questions/585942/how-to-display-the-first-two-authors-in-a-citation-call-out.
@@
 
-	PG-SGD is inspired by \cite{Zheng2019}, but we designed our algorithm to work on pangenomes in the form of a variation graph introduced in Definition \ref{def:vg}.
+	PG-SGD is inspired by \cite{Zheng2019}, but we designed the algorithm to work on pangenomes in the form of a variation graph (Definition \ref{def:vg}).
 

### 2023-07-01 manuscript/main.tex
fix errors and make text appealing for non germans

@@
 		Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ describe walks over node strands ($p_i = s_1 \ldots s_{|p_i|}$), representing the genomes embedded in the graph.
+		\FIXME{EXPLAIN THE STEPS}
 	\end{definition}
 
-	In brief, the algorithm moves a single pair of nodes $\mathcal{V}_i$, $\mathcal{V}_j$ at a time, minimizing the disparity between the layout distance $ld_{ij}$ of a node pair and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
+	PG-SGD's pseudocode is shown in Algorithm~\ref{alg:pg_sgd} and key steps are sketched in Figure \FIXME{Add puritan figure.}.
+	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
 	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
-	\begin{enumerate*}[label=(\roman*)]
-		\item The first node $\mathcal{V}_i$ of a pair is a uniform path step pick from all nodes.
-		\item The second node $\mathcal{V}_j$ of a pair is sampled from the same path following a Zipfian distribution \citep{Zipf1932}.
-		\item The path nucleotide distance  $nd_{ij}$ guides the actual layout distance $ld_{ij}$ update of these nodes. Also referred to as \textit{term update}.
-		\item The magnitude $r$ of the update depends on the current learning rate $\mu$ of the SGD.
-	\end{enumerate*}
-	These key steps are sketched in Figure \FIXME{Add puritan figure.}. Pseudocode is shown in Algorithm~\ref{alg:pg_sgd}.
+	$v_i$ is the node associated with the step $s_i$ sampled uniformly from all the steps in $\mathcal{P}$.
+	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by following a Zipfian distribution \citep{Zipf1932}.
+	The difference between $nd_{ij}$ and $ld_{ij}$ guides the update of the node coordinates in the layout.
+	The magnitude $r$ of the update depends on the learning rate $\mu$ of the SGD.
+	The number of iterations steers the annealing step size $\eta$ which determines the learning rate $\mu$.
+	%A key property of pangenome graph layouts is that their nodes can be arranged so that most edges go between nodes that are close together in the node order \citep{Guarracino2022}.
+	A large $\eta$ in the first iterations leads to a globally linear (in 1D) or planar (in 2D) layout.
+	By decreasing $\eta$, the layout adjustments become more localized, ensuring that the nodes are positioned in a way to best reflect the nucleotide distances in the paths (i.e., in the genomes).
+% LIAR: down to the single base pair level.
 
@@
 
-	The number of iterations steer the annealing step size $\eta$ which in turn determines the learning rate $\mu$ of PG-SGD. 
-	Since variation graphs are very sparse, a large step size in the beginning of the schedule already leads to a globally linear \FIXME{(or planar?)} layout.
-	With decreasing $\eta$ the node layout adjustments become more local, ensuring that the nodes are positioned in away to best reflect the nucleotde distances of the paths down to the single base pair level.
-
-	In standard SGD layout algorithms, the problem space is given by pairwise node comparisons.
+	%In standard SGD layout algorithms, the problem space is given by pairwise node comparisons.
 In PG-SGD, the problem space is given by the pairwise path step comparisons.
@@
 	Originating from empirical inspection of word frequency tables, Zipf's law states that a word with rank $n$ occurs $1/n$ times as the most frequent one.
-	It's mathematical formulation is called the Zipfian distribution whichs rank-frequency distribution is related to the inverse power law.
-	Sampling from a Zipfian distribution generated with a path's nucleotide position space increases the possibility to draw a small nucleotide position.
-	So there is a high chance of a small nucleotide distance between node pair $\mathcal{V}_ij$ to refine the layout down to the nodes containing only one base pair.
+	It's mathematical formulation is called the Zipfian distribution and it rank-frequency distribution is related to the inverse power law.
+	Sampling $s_j$ from a Zipfian distribution fixed in the $s_i$'s path position space increases the possibility to draw a nucleotide position close to $s_i$.
+	So there is a high chance to use small nucleotide distances $nd_{ij}$ to refine the layout of nodes comprising a few base pairs.
+	\FIXME{TO IMPROVE, AS IT DOES NOT EXPLAIN WHY ZIPFIAN AND NOT OTHER FUNCTIONS}
 
-	To provide balance between global and local term updates, there is a 50\% chance for a \textit{flip}: In half of our term updates the second path step is sampled from a uniform distribution instead of a Zipfian one.
-Additionally, after half the number of iterations, a \textit{cooling} phase skewes the Zipfian distribution so that the possibility of a a small sampled nucleotide distance is significantly increased.
-Again, this boosts the local planarity of the graph layout.
+	To provide balance between global and local layout updates, in half of the updates ($flip$ flag in Algorithm~\ref{alg:pg_sgd}) the $s_j$ is sampled uniformly instead from a Zipfian distribution (uniform sampling favours global updates).
+	Additionally, to boost the local linearity (in 1D) or planarity (in 2D) of the graph layout, a \textit{cooling} phase skewes the Zipfian distribution after half the number of iterations, so that the possibility of a a small sampled nucleotide distance is significantly increased.
 
@@
 
-	We implemented the algorithm in ODGI which serves as an interface to graphs in the GFA (\url{https://github.com/GFA-spec/GFA-spec}) format. 
-	Our path index is a strict subset of the XG index~\citep{Garrison:2018} with the exception that
-	we specifically avoided to use succinct SDSL~\citep{Gog2014} data structures. We only use bit compressed integer vectors. 
+	We implemented PG-SGD in ODGI \citep{Guarracino2022}. % which serves as an interface to graphs in the GFA (\url{https://github.com/GFA-spec/GFA-spec}) format. 
+	Our path index is a strict subset of the XG index~\citep{Garrison:2018} with the exception that we specifically avoid using succinct SDSL data structures \citep{Gog2014}.
+	We only use bit compressed integer vectors. 
 	This enables efficient retrieval of path positions to quickly compute nucleotide distances and eliminates the need to store all pairwise distances between nodes in memory, allowing PG-SGD to be applied to gigabase-scale pangenome graphs.
 
-	Our implementation is multithreaded, threads can perform term update calculations in parallel.
+	Our implementation is multithreaded, with threads performing layout updates without any locking.
 	Layout values are stored in an atomic integer vector presenting the only multithreading bottleneck.
-Effectively, this presents a modified version of the HOGWILD!
-SGD algorithm which just writes the layout updates to a shared non-atomic integer vector.
+	Effectively, this presents a modified version of the HOGWILD! SGD algorithm which just writes the layout updates to a shared non-atomic integer vector.
 	We went for an atomic vector, because this ensures memory saftey and in our experience there was basically no performance loss.

### 2023-07-01 manuscript/pg-sgd_pseudocode.tex
fix errors and make text appealing for non germans

@@
 			\each{$planned$ $term$ $update$} { % I think we can remove i<j, because we don't care about that
-				$s_a \gets \Unif(\mathcal{XP})$ \tcp{uniformly sample a step from $\mathcal{P}$}
-				$p \gets \Path(s_a,\mathcal{XP})$ \tcp{path of $s_a$}
+				$s_i \gets \Unif(\mathcal{XP})$ \tcp{uniform sampling of a step from $\mathcal{P}$}
+				$p \gets \Path(s_i,\mathcal{XP})$ \tcp{path of $s_i$}
 				\uIf{$(cooling$ $||$ $flip)$} {
-					$s_b \gets \Unif(\StepCount(p, \mathcal{XP}))$ \tcp{uniformly sample a step from $p$}
+					$s_j \gets \Unif(\StepCount(p, \mathcal{XP}))$ \tcp{uniform sampling of a step from $p$}
 				} \Else {
-					$s_b \gets \Zipf(p)$ \tcp{Zipfian sampling of a step from $p$}
+					$s_j \gets \Zipf(p)$ \tcp{Zipfian sampling of a step from $p$}
 				}
-				$p_a \gets \StepPos(s_a)$ \tcp{nuc. position}
-				$p_b \gets \StepPos(s_b)$ \tcp{nuc. position}
-				$nd \gets ||p_a - p_b||$ \tcp{nuc. distance} %mag is nx
-				$ld \gets ||\mathcal{L}_{a}-\mathcal{L}_b||$ \tcp{layout distance}
-				$w_{ab} \gets \frac{1.0}{nd}$ \tcp{term weight} 
-				$\mu \gets w_{ab}\eta$ \tcp{learning rate}  % current learning rate is given by term weight and step size
+				$p_i \gets \StepPos(s_i)$ \tcp{nuc. position}
+				$p_j \gets \StepPos(s_j)$ \tcp{nuc. position}
+				$nd_{ij} \gets ||p_i - p_j||$ \tcp{nuc. distance} %mag is nx
+				$ld_{ij} \gets ||l_{i} - l_j||$ \tcp{layout distance}
+				$w_{ij} \gets \frac{1.0}{nd_{ij}}$ \tcp{term weight} 
+				$\mu \gets w_{ij}\eta$ \tcp{learning rate}  % current learning rate is given by term weight and step size
 				\IF{$\mu>1$} {
@@
 				}
-				$\delta \gets \mu \cdot \frac{ld -nd}{2}$ \tcp{the actual delta}
+				$\delta \gets \mu \cdot \frac{ld -nd_{ij}}{2}$ \tcp{the actual delta}
 				\uIf{$\delta <= 0$} {
@@
 				% TODO potentially store new delta max?
-				$r \gets \delta - ld$ \tcp{size of the term update}
-				$\mathcal{L}_a \gets \mathcal{L}_a - r\cdot ld$ \tcp{update the 1st node}
-				$\mathcal{L}_b \gets \mathcal{L}_b - r\cdot ld$ \tcp{update the 2nd node}
+				$r \gets \delta - ld_{ij}$ \tcp{size of the update}
+				$l_i \gets l_i - r\cdot ld_{ij}$ \tcp{update $v_i$ coordinates}
+				$l_j \gets l_j - r\cdot ld_{ij}$ \tcp{update $v_j$ coordinates}
 			}
@@
 	}
-	\caption{Pseudocode of the PG-SGD algorithm.}
+	\caption{Pseudocode of PG-SGD.}
 	\label{alg:pg_sgd}

### 2023-07-01 manuscript/main.tex
continue to make the text appealing for non germans

@@
 	We implemented PG-SGD in ODGI \citep{Guarracino2022}. % which serves as an interface to graphs in the GFA (\url{https://github.com/GFA-spec/GFA-spec}) format. 
-	Our path index is a strict subset of the XG index~\citep{Garrison:2018} with the exception that we specifically avoid using succinct SDSL data structures \citep{Gog2014}.
-	We only use bit compressed integer vectors. 
-	This enables efficient retrieval of path positions to quickly compute nucleotide distances and eliminates the need to store all pairwise distances between nodes in memory, allowing PG-SGD to be applied to gigabase-scale pangenome graphs.
+	Our path index is a strict subset of the XG index~\citep{Garrison:2018} where we specifically avoid using succinct SDSL data structures \citep{Gog2014}.
+	Instead, we only use bit compressed integer vectors, enabling efficient retrieval of path positions to quickly compute nucleotide distances without the need to store all pairwise distances between nodes in memory.
+	This allows PG-SGD to scale on large pangenome graphs representing thousands of whole genomes.
 
-	Our implementation is multithreaded, with threads performing layout updates without any locking.
-	Layout values are stored in an atomic integer vector presenting the only multithreading bottleneck.
-	Effectively, this presents a modified version of the HOGWILD! SGD algorithm which just writes the layout updates to a shared non-atomic integer vector.
-	We went for an atomic vector, because this ensures memory saftey and in our experience there was basically no performance loss.
+	Graph layout initialization can significantly influence the quality of the final layout.
+	In the 1D implementation, nodes are placed in the same order as they appear in the input graph by default, although support for random layout initialization is also provided.
+	his random placement ensures that each run of the PG-SGD method could potentially yield a different result, thereby increasing the likelihood of obtaining a meaningful layout.
+	In 2D, we offer several layout initialization techniques.
+	One option is to position nodes in the first layout dimension according to their order in the input graph and to add either uniform or Gaussian noise in the second dimension.
+	Another possibility is to arrange nodes along a Hilbert curve, which often favors the creation of planar final layouts.
 
-	Due to the fact that we have a very high learning rate, our 1D implementation can work with a random layout as input.
-Though, by default, it just makes the given node order the starting layout.
-In 2D, we offer several random layout initializations:
-		\begin{enumerate*}[label=(\roman*)]
-		\item Uniform,
-		\item gaussian,
-		\item or Hilbert.
-	\end{enumerate*}
+	Our implementation is multithreaded and follows the HOGWILD! strategy, that is threads perform layout updates without any locking.
+	This is possible because pangenome graphs are tipically sparse \citep{Guarracino2022}, with low average node degree, meaning most updates only modify small parts of the entire layout.
+	With respect to the HOGWILD! SGD algorithm \citep{Recht2011} which writes the layout updates to a shared non-atomic integer vector, PG-SGD stores node coordinates in an atomic integer vector.
+	We prefer an atomic vector to ensure memory safety because, in our experience, there is no performance loss with respect to the non-atomic counterpart.
 

### 2023-07-01 manuscript/main.tex
more appelaing for non germans

@@
 
-	Our implementation is multithreaded and follows the HOGWILD! strategy, that is threads perform layout updates without any locking.
+	Our implementation is multithreaded and follows the HOGWILD! strategy \citep{Recht2011}, that is threads perform layout updates without any locking.
 	This is possible because pangenome graphs are tipically sparse \citep{Guarracino2022}, with low average node degree, meaning most updates only modify small parts of the entire layout.
-	With respect to the HOGWILD! SGD algorithm \citep{Recht2011} which writes the layout updates to a shared non-atomic integer vector, PG-SGD stores node coordinates in an atomic integer vector.
+	With respect to the HOGWILD! SGD algorithm, which writes the layout updates to a shared non-atomic integer vector, PG-SGD stores node coordinates in an atomic integer vector.
 	We prefer an atomic vector to ensure memory safety because, in our experience, there is no performance loss with respect to the non-atomic counterpart.
@@
 
-	\iffalse
-	Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
-    \\
-    1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}.
-    But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1.
-    \\
-    2D (max. threads) + odgi draw vs. Bandage in Fig.~\ref{fig:bandage}.
-    \input{fig_performance}
-    \\
-    \\
-    Table 1: Metrics of the ALIBI and PG-SGD graphs from Fig. 2.
-    \input{tab_performance.tex}
-    Already, PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
-
-    \paragraph{Latent graph structure reveals underlying biology}
-    Fig. 3: Cool quantitative 1D sortings and 2D layouts: biological implications.
-    Randomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
-    We want a pipeline of sortings. 2D layout of the whole HPRC.
-    Chr6 HPRC HLA graph? Chr8 beta-defensin gene cluster HPRC? Whole HPRC?
-    \\
-    We could also try to build a gastric cancer pangenome graph with data from \url{https://www.nature.com/articles/s41467-022-33073-7} from up to 185 samples.
-    However, we would have to request access to the data.
-    \\
-    \\
-    \input{fig_latent_graph_structure}
-    Table 2: Metrics of the sorted graphs in Fig. 3.
-    \input{tab_metrics_latent_graph_structure}
-
-    \paragraph{Bonus Section}
-    Fig. 4: Detect tension. Relax a graph. Detect tension afterwards.
-    I need to test this on a new data set I got from Erik.
-    I need to establish a fixed lower boundary for the tension from which on we don't relax anymore. \\
-    With a high quality layout, we can measure the discrepancy of the path layout position versus the expected path nucleotide position, the “tension” of a graph.
-    The greater the “tension”, the greater is the possibility of a biologically meaningless alignment.
-    This allows us to detect telomere collapsing alignment errors and hopefully (pangenome) assembly errors, subsequently correcting them.
-    \input{fig_tension}
-    \fi
+	% \iffalse
+	% Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
+    % \\
+    % 1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}.
+    % But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1.
+    % \\
+    % 2D (max. threads) + odgi draw vs. Bandage in Fig.~\ref{fig:bandage}.
+    % \input{fig_performance}
+    % \\
+    % \\
+    % Table 1: Metrics of the ALIBI and PG-SGD graphs from Fig. 2.
+    % \input{tab_performance.tex}
+    % Already, PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
+    % \paragraph{Latent graph structure reveals underlying biology}
+    % Fig. 3: Cool quantitative 1D sortings and 2D layouts: biological implications.
+    % Randomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
+    % We want a pipeline of sortings. 2D layout of the whole HPRC.
+    % Chr6 HPRC HLA graph? Chr8 beta-defensin gene cluster HPRC? Whole HPRC?
+    % \\
+    % We could also try to build a gastric cancer pangenome graph with data from \url{https://www.nature.com/articles/s41467-022-33073-7} from up to 185 samples.
+    % However, we would have to request access to the data.
+    % \\
+    % \\
+    % \input{fig_latent_graph_structure}
+    % Table 2: Metrics of the sorted graphs in Fig. 3.
+    % \input{tab_metrics_latent_graph_structure}
+    % \paragraph{Bonus Section}
+    % Fig. 4: Detect tension. Relax a graph. Detect tension afterwards.
+    % I need to test this on a new data set I got from Erik.
+    % I need to establish a fixed lower boundary for the tension from which on we don't relax anymore. \\
+    % With a high quality layout, we can measure the discrepancy of the path layout position versus the expected path nucleotide position, the “tension” of a graph.
+    % The greater the “tension”, the greater is the possibility of a biologically meaningless alignment.
+    % This allows us to detect telomere collapsing alignment errors and hopefully (pangenome) assembly errors, subsequently correcting them.
+    % \input{fig_tension}
+    % \fi
 
@@
 	We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
-	Our implementation efficiently computes the layout of pangenome graphs representing thousands of individuals at whole-genome scale.
-	\FIXME{Difference to other existing methods, are there possible improvements of our method possible?}
-	\FIXME{Discuss performance eval. @Jiajie + Niklas: What about going GPU?}
+	Our implementation efficiently computes the layout of pangenome graphs representing thousands of whole genomes.
+	\FIXME{Discuss differences to other existing methods? are there possible improvements for our method?}
 
@@
 	We implemented PG-SGD to generate layouts in 1 and 2 dimensions.
-	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of acrocentric chromosomes \citep{Guarracino2023}.
+	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of human acrocentric chromosomes \citep{Guarracino2023}.
 	Furthermore, they are applied in the creation and analysis of pangenome graphs for any species \citep{Guarracino2022, Garrison2023}.
@@
 So we have a very large problem space compared to just looking at the nodes.
-	There still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
-	Our algorithm will be the foundation of new pangenome graph browsers for exploring meaningful graph layouts (\href{https://github.com/chfi/waragraph}{https://github.com/chfi/waragraph}).
+	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
+	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph}).
 
 	PG-SGD can be extended to any number of dimensions.
-	Projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout for variant detection and classification.
-	Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.
-
-	Our future research involves leveraging these graph projections to detect structural variants and detect and solve assembly errors.
-	Moreover, we consider the extension of the algorithm to RNA and protein sequences.
+	It can be seen as a graph embedding algorithm that converts high-dimensional, sparse pangenome graphs into low-dimensional, dense, and continuous vector spaces, while preserving its biologically relevant information.
+	This enables the application of machine learning algorithms that use the graph layout for variant detection and classification. 
+	Our future research involves leveraging these graph projections to detect structural variants and to identify and correct assembly errors
+	Moreover, we are considering extending the algorithm to RNA and protein sequences to support pantranscriptomes and panproteomes, respectively.
+	%Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.
+	\FIXME{@Jiajie + Niklas: What about going GPU?}
 
@@
 
-	\section{Animations}
-	\FIXME{ADD LINKS TO OUR ANIMATIONS ON ZENODO}
-
 	\bibliographystyle{natbib}
@@
 	    %\FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y} \\
+		\FIXME{ADD LINKS TO OUR ANIMATIONS ON ZENODO} \\
 	    \FIXME{From 1D layout to 1D node order} \\

### 2023-07-01 manuscript/main.tex
cite more external groups

@@
 	Our future research involves leveraging these graph projections to detect structural variants and to identify and correct assembly errors
-	Moreover, we are considering extending the algorithm to RNA and protein sequences to support pantranscriptomes and panproteomes, respectively.
+	Moreover, we are considering extending the algorithm to RNA and protein sequences to support pantranscriptomes \citep{sibbesen_haplotype-aware_2023} and panproteomes \citep{dabbaghie_panpa:_2023}, respectively.
 	%Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.
@@
 	\begin{appendices}
-	    \section{Appendix}
+	    \section{Supplementary data}
 	    %\FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\

### 2023-07-01 scripts/hprc.sh
hprc.sh in progress

@@ -1,0 +1,28 @@
+#!/bin/bash
+
+# Prepare the tools
+cd  ~/tools
+git clone --recursive https://github.com/pangenome/odgi.git
+cd odgi
+git checkout cffe70a14ce0cf51aa7e2e1c297b75cdcc33d5e9
+cmake -H. -Bbuild && cmake --build build -- -j 48
+cp bin/odgi bin/odgi-cffe70a14ce0cf51aa7e2e1c297b75cdcc33d5e9
+
+# Prepare the folder
+mkdir -p /lizardfs/guarracino/sorting-paper
+cd /lizardfs/guarracino/sorting-paper
+
+# Download HPRC graphs
+(seq 1 22; echo X; echo Y) | while read i; do echo $i; wget -c https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/freeze/freeze1/pggb/chroms/chr${i}.hprc-v1.0-pggb.gfa.gz; done
+
+# GFA format to ODGI format
+ODGI=/home/guarracino/tools/odgi/bin/odgi-cffe70a14ce0cf51aa7e2e1c297b75cdcc33d5e9
+(seq 1 22; echo X; echo Y) | while read i; do echo $i; sbatch -p workers -c 48 --wrap "hostname; \time -v $ODGI build -g /lizardfs/guarracino/sorting-paper/chr${i}.hprc-v1.0-pggb.gfa.gz -o /lizardfs/guarracino/sorting-paper/chr${i}.hprc-v1.0-pggb.og -t 48 -P"; done
+
+# Compute the layout (both LAY and TSV formats)
+ODGI=/home/guarracino/tools/odgi/bin/odgi-cffe70a14ce0cf51aa7e2e1c297b75cdcc33d5e9
+(seq 1 22; echo X; echo Y) | while read i; do echo $i; sbatch -p 386mem -x octopus07,octopus10 -c 48 --wrap "hostname; cp /lizardfs/guarracino/sorting-paper/chr${i}.hprc-v1.0-pggb.og /scratch/chr${i}.hprc-v1.0-pggb.og; \time -v $ODGI layout -i /lizardfs/guarracino/sorting-paper/chr${i}.hprc-v1.0-pggb.og -o xxx.lay -T xxx.tsv -t 48 --temp-dir xxx -P; move and clean"; done
+
+# Compute the layout (both LAY and TSV formats) with harder parameters
+
+# Draw the layouts

### 2023-07-01 scripts/hprc.sh
continue hprc.sh

@@ -8,21 +8,26 @@ git checkout cffe70a14ce0cf51aa7e2e1c297b75cdcc33d5e9
 cmake -H. -Bbuild && cmake --build build -- -j 48
 cp bin/odgi bin/odgi-cffe70a14ce0cf51aa7e2e1c297b75cdcc33d5e9
 
-# Prepare the folder
-mkdir -p /lizardfs/guarracino/sorting-paper
-cd /lizardfs/guarracino/sorting-paper
+ODGI=/home/guarracino/tools/odgi/bin/odgi-cffe70a14ce0cf51aa7e2e1c297b75cdcc33d5e9
 
-# Download HPRC graphs
-(seq 1 22; echo X; echo Y) | while read i; do echo $i; wget -c https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/freeze/freeze1/pggb/chroms/chr${i}.hprc-v1.0-pggb.gfa.gz; done
+# Download and gunzip HPRC graphs
+mkdir -p /lizardfs/guarracino/sorting-paper/graphs
+cd /lizardfs/guarracino/sorting-paper/graphs
+(seq 1 22; echo X; echo Y) | while read i; do echo $i; wget -c https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/freeze/freeze1/pggb/chroms/chr${i}.hprc-v1.0-pggb.gfa.gz; gunzip chr${i}.hprc-v1.0-pggb.gfa.gz; done
 
 # GFA format to ODGI format
-ODGI=/home/guarracino/tools/odgi/bin/odgi-cffe70a14ce0cf51aa7e2e1c297b75cdcc33d5e9
-(seq 1 22; echo X; echo Y) | while read i; do echo $i; sbatch -p workers -c 48 --wrap "hostname; \time -v $ODGI build -g /lizardfs/guarracino/sorting-paper/chr${i}.hprc-v1.0-pggb.gfa.gz -o /lizardfs/guarracino/sorting-paper/chr${i}.hprc-v1.0-pggb.og -t 48 -P"; done
+cd /lizardfs/guarracino/sorting-paper/graphs
[…]

### 2023-07-03 scripts/hprc.sh
no P

@@ -22,12 +22,13 @@ cd /lizardfs/guarracino/sorting-paper/graphs
 # Compute the layout (both LAY and TSV formats) with default parameters
 mkdir -p /lizardfs/guarracino/sorting-paper/layouts
 cd /lizardfs/guarracino/sorting-paper/layouts
-(seq 1 22; echo X; echo Y) | while read i; do echo $i; sbatch -p 386mem -x octopus07,octopus10 -c 48 --wrap "hostname; cd /scratch; cp /lizardfs/guarracino/sorting-paper/graphs/chr${i}.hprc-v1.0-pggb.og chr${i}.hprc-v1.0-pggb.og; \time -v $ODGI layout -i chr${i}.hprc-v1.0-pggb.og -o chr${i}.hprc-v1.0-pggb.lay -T chr${i}.hprc-v1.0-pggb.tsv -t 48 --temp-dir /scratch/ -P; rm chr${i}.hprc-v1.0-pggb.og; mv chr${i}.hprc-v1.0-pggb.* /lizardfs/guarracino/sorting-paper/layouts/"; done
+(seq 1 22; echo X; echo Y) | while read i; do echo $i; sbatch -p 386mem -x octopus07,octopus10 -c 48 --wrap "hostname; cd /scratch; cp /lizardfs/guarracino/sorting-paper/graphs/chr${i}.hprc-v1.0-pggb.og chr${i}.hprc-v1.0-pggb.og; \time -v $ODGI layout -i chr${i}.hprc-v1.0-pggb.og -o chr${i}.hprc-v1.0-pggb.lay -T chr${i}.hprc-v1.0-pggb.tsv -t 48 --temp-dir /scratch/; rm chr${i}.hprc-v1.0-pggb.og; mv chr${i}.hprc-v1.0-pggb.* /lizardfs/guarracino/sorting-paper/layouts/"; done
 
 # Compute the layout (both LAY and TSV formats) with harder parameters (-x 300 -G 30 -I 50000 -l 5000)
 cd /lizardfs/guarracino/sorting-paper/layouts
[…]

### 2023-07-04 manuscript/main.tex
less hogwild

--- 2023-07-01 manuscript/main.tex
@@
     %\title[Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
-    %\title[Pangenome graph layout by Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by Path-Guided Stochastic Gradient Descent}
+    \title[Pangenome graph layout by Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by Path-Guided Stochastic Gradient Descent}
     %\title[Unveiling genome variation by sequence-guided pangenome graph layouts]{Unveiling genome variation by sequence-guided pangenome graph layouts}
@@
     %\title[Genome-guided pangenome graph layouts]{Genome-guided pangenome graph layouts}
-	\title[Sequence-guided pangenome graph layouts]{Sequence-guided pangenome graph layouts}
+	%\title[Sequence-guided pangenome graph layouts]{Sequence-guided pangenome graph layouts}
 
@@
 	To address these challenges, we propose a novel graph layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-	PG-SGD uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel by applying the HOWGWILD! strategy.
+	PG-SGD uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel by applying a modified HOWGWILD! strategy.
 	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, revealing their biological features. \\
@@
 
-	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in parallel with a HOGWILD! strategy.
+	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in parallel with a modified HOGWILD! strategy.
 	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
@@
 	Originating from empirical inspection of word frequency tables, Zipf's law states that a word with rank $n$ occurs $1/n$ times as the most frequent one.
-	It's mathematical formulation is called the Zipfian distribution and it rank-frequency distribution is related to the inverse power law.
+	It's mathematical formulation is called the Zipfian distribution and its rank-frequency distribution is related to the inverse power law.
 	Sampling $s_j$ from a Zipfian distribution fixed in the $s_i$'s path position space increases the possibility to draw a nucleotide position close to $s_i$.
@@
 	In the 1D implementation, nodes are placed in the same order as they appear in the input graph by default, although support for random layout initialization is also provided.
-	his random placement ensures that each run of the PG-SGD method could potentially yield a different result, thereby increasing the likelihood of obtaining a meaningful layout.
+	This random placement ensures that each run of the PG-SGD method could potentially yield a different result, thereby increasing the likelihood of obtaining a meaningful layout.
 	In 2D, we offer several layout initialization techniques.
@@
 
-	Our implementation is multithreaded and follows the HOGWILD! strategy \citep{Recht2011}, that is threads perform layout updates without any locking.
-	This is possible because pangenome graphs are tipically sparse \citep{Guarracino2022}, with low average node degree, meaning most updates only modify small parts of the entire layout.
-	With respect to the HOGWILD! SGD algorithm, which writes the layout updates to a shared non-atomic integer vector, PG-SGD stores node coordinates in an atomic integer vector.
-	We prefer an atomic vector to ensure memory safety because, in our experience, there is no performance loss with respect to the non-atomic counterpart.
+	Our implementation is multithreaded and uses shared memory for the layout vector following the HOGWILD!
+strategy \citep{Recht2011}: Threads perform layout updates without any locking for additional speed up.
+	This is possible because pangenome graphs are typically sparse \citep{Guarracino2022}, with low average node degree, meaning most updates only modify small parts of the entire layout.
+	The HOGWILD! SGD algorithm writes the layout updates to a shared non-atomic double vector, whereas PG-SGD stores node coordinates in a vector of atomic doubles.
+	Such a vector ensures that no memory overwrites are possible. Our tests revealed basically no performance loss with respect to the non-atomic counterpart.
 
@@
 	Our future research involves leveraging these graph projections to detect structural variants and to identify and correct assembly errors
-	Moreover, we are considering extending the algorithm to RNA and protein sequences to support pantranscriptomes and panproteomes, respectively.
+	Moreover, we are considering extending the algorithm to RNA and protein sequences to support pantranscriptome graphs \citep{sibbesen_haplotype-aware_2023} and panproteome graphs \citep{dabbaghie_panpa:_2023}, respectively.
 	%Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.

### 2023-07-13 manuscript/main.tex
fix steps

@@
 		Each node $v_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{v_i}$.
-		The node strand $s$ represents node orientation.
-		Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands ($e_i = ( s_a, s_b )$), defining the graph topology.
-		Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ describe walks over node strands ($p_i = s_1 \ldots s_{|p_i|}$), representing the genomes embedded in the graph.
-		\FIXME{EXPLAIN THE STEPS}
+		The node strand $o$ represents node orientation.
+		Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands ($e_i = ( o_a, o_b )$), defining the graph topology.
+		Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ are series of connected steps $s_i$ that refer to node strands in the graph ($p_i = s_1 \ldots s_{|p_i|}$); the paths represent the genomes embedded in the graph.
 	\end{definition}

### 2023-07-13 manuscript/main.tex
practical
reduce missing stuff

@@
 	So there is a high chance to use small nucleotide distances $nd_{ij}$ to refine the layout of nodes comprising a few base pairs.
-	\FIXME{TO IMPROVE, AS IT DOES NOT EXPLAIN WHY ZIPFIAN AND NOT OTHER FUNCTIONS}
 
@@
     \subsection{Performance}
-	\FIXME{Describe the dataset (HPRC chromosomes), runtime and memory usage (specifying the fastest/slowest chromosome pangenome graph layout and the thinnest/fattest in memory), just text no with a table}
-    \FIXME{A comparison with Bandage might be worth, to give readers an idea of the scalability}
+	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} to show the scalability and practicality of the algorithm with a real pangenome.
+	Experiments were conducted on computer with \FIXME{XXX GB of RAM and XXXAMD/INTELXXX processors with XXX vCPUs}.
+	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
+	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
+	See \FIXME{SUPPLEMENTARY TABLE XXX WITH GRAPH LENGTH AND NUMBER OF NODES, EDGES AND STEPS} for graph statistics).
+	When applied to these pangenome graphs, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16.
+	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
+	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
+	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with \FIXME{chromosome XXX}.
+	Given its scalability, PG-SGD can even be applied to the full graph with all chromosomes together \FIXME{SUPPLEMENTARY TABLE WITH RUNTIME, MEMORY USAGE; SUPPLEMENTARY FIGURE whe}.
 
     \subsection{Pangenome graphs layouts reveal biology features}
-	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.
-	In the supp a huge picture with a "karyotype", that is all chromosome layout together sorted by chromosome number, might be hot}
+	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent.
+	We show how PG-SGD allows us quickly gain insight into biological data by looking at the graph layout structure.
+	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.}
 
@@
 	Our implementation efficiently computes the layout of pangenome graphs representing thousands of whole genomes.
-	\FIXME{Discuss differences to other existing methods? are there possible improvements for our method?}
 

### 2023-07-14 manuscript/main.tex
layout table stats
progress the evaluation

@@
 \usepackage[inline]{enumitem}
+
+% table stuff
+\usepackage{amsfonts}
+\usepackage{booktabs}
+\usepackage{siunitx}
+\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
+\usepackage{float}
+\floatstyle{plaintop}
+\restylefloat{table}
 
@@
 	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} to show the scalability and practicality of the algorithm with a real pangenome.
-	Experiments were conducted on computer with \FIXME{XXX GB of RAM and XXXAMD/INTELXXX processors with XXX vCPUs}.
+	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads 2* AMD EPYC 7343 processors with 512 GB RAM each) and 4 \texttt{HighMem} nodes (64 cores / 128 threads 2* AMD EPYC 7513 processors with 2048 GB RAM each).
+The high-mem nodes were only used for the calculations of the 2D layouts of the full HPRC graph including all chromosomes.
 	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
 	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
-	See \FIXME{SUPPLEMENTARY TABLE XXX WITH GRAPH LENGTH AND NUMBER OF NODES, EDGES AND STEPS} for graph statistics).
-	When applied to these pangenome graphs, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16.
+	See Supplementary Table~\ref{tab:layout} for graph statistics.
+	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16.
 	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
 	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
-	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with \FIXME{chromosome XXX}.
+	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with chromosome 1. An overview of these evaluations is given in Supplementary Table~\ref{tab:layout}.
 	Given its scalability, PG-SGD can even be applied to the full graph with all chromosomes together \FIXME{SUPPLEMENTARY TABLE WITH RUNTIME, MEMORY USAGE; SUPPLEMENTARY FIGURE whe}.
@@
 	    \FIXME{reference-guided path-guided stochastic gradient descent}
+	    \subsection{Performance evaluation}
+	    \label{sec:performance}
+	    \input{table_layout}
 	\end{appendices}

### 2023-07-14 manuscript/main.tex
tweaks

--- 2023-07-13 manuscript/main.tex
@@
 	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} to show the scalability and practicality of the algorithm with a real pangenome.
-	Experiments were conducted on computer with \FIXME{XXX GB of RAM and XXXAMD/INTELXXX processors with XXX vCPUs}.
+	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads AMD EPYC 7343 processors with 512 GB RAM each) and 4 \texttt{HighMem} nodes (64 cores / 128 threads AMD EPYC 7513 processors with 2048 GB RAM each).
 	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
 	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
-	See \FIXME{SUPPLEMENTARY TABLE XXX WITH GRAPH LENGTH AND NUMBER OF NODES, EDGES AND STEPS} for graph statistics).
-	When applied to these pangenome graphs, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16.
+	See Supplementary Table~\ref{tab:layout} for graph statistics.
+	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16 (Supplementary Table~\ref{tab:layout}).
 	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
 	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
-	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with \FIXME{chromosome XXX}.
-	Given its scalability, PG-SGD can even be applied to the full graph with all chromosomes together \FIXME{SUPPLEMENTARY TABLE WITH RUNTIME, MEMORY USAGE; SUPPLEMENTARY FIGURE whe}.
+	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with chromosome 1, the biggest one in humans.
+	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}).
+	Overall, PG-SGD is ~\FIXME{XX}X faster than BandageNG \FIXME{HOW TO CITE IT?}, the current state-of-the-art for graph visualization, while using ~{}X less memory.
 
@@
 	We show how PG-SGD allows us quickly gain insight into biological data by looking at the graph layout structure.
-	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.}
+	\FIXME{Describe a full chromosome and/or 1/a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the human paper or PGR-TK.}
 
 	% \iffalse
-	% Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
+	% Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full human graph.
     % \\
@@
 	%Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
-	%Problem space: We take a look at all paths (FULL HPRC GRAPH ~8 BILLION STEPS ACROSS ~34 THOUSAND PATHS) and not at all nodes (FULL HPRC GRAPH ~100 MILLION).
-So we have a very large problem space compared to just looking at the nodes.
 	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
 	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph}).
+
+	\FIXME{@Jiajie + Niklas: just 2/3 sentences about going GPU?}
 
@@
 	%Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.
-	\FIXME{@Jiajie + Niklas: What about going GPU?}
 
@@
 
-	We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
-	We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
+	%We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
+	%We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
 	A special thanks goes to Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.
@@
 	    \FIXME{reference-guided path-guided stochastic gradient descent}
+	    \subsection{Performance evaluation}
+	    \label{sec:performance}
+	    \input{table_layout}
+		\subsection{Graph layouts}
+		\FIXME{chr1}
+		\FIXME{chr2}
+		\FIXME{...}
+		\FIXME{FIGURE WITH ALL CHROMOSOMES TOGETHER}
 	\end{appendices}

### 2023-07-15 scripts/hprc.sh
extractions

@@ -27,3 +27,58 @@ cd /lizardfs/guarracino/sorting-paper/layouts
 # Draw the layouts
 cd /lizardfs/guarracino/sorting-paper/layouts
 (seq 1 22; echo X; echo Y) | while read i; do echo $i; sbatch -p workers -c 48 --wrap "hostname; cd /scratch; cp /lizardfs/guarracino/sorting-paper/graphs/chr${i}.hprc-v1.0-pggb.og chr${i}.hprc-v1.0-pggb.og; cp /lizardfs/guarracino/sorting-paper/layouts/chr${i}.hprc-v1.0-pggb.lay chr${i}.hprc-v1.0-pggb.lay; \time -v $ODGI draw -i chr${i}.hprc-v1.0-pggb.og -c chr${i}.hprc-v1.0-pggb.lay -p chr${i}.hprc-v1.0-pggb.draw.png -H 1000; \time -v $ODGI draw -i chr${i}.hprc-v1.0-pggb.og -c chr${i}.hprc-v1.0-pggb.lay -p chr${i}.hprc-v1.0-pggb.draw.C.png -C -w 20 -H 1000; rm chr${i}.hprc-v1.0-pggb.og; rm chr${i}.hprc-v1.0-pggb.lay; mv chr${i}.hprc-v1.0-pggb.draw*png /lizardfs/guarracino/sorting-paper/layouts/"; done
+
+# MHC
+odgi extract -i chr6.hprc-v1.0-pggb.og -d 1000000 -r grch38#chr6:29000000-34000000 -o - -t 48 -P | odgi sort -i - -o chr6.hprc-v1.0-pggb.MHC.og -O
+
+# RCCE
+odgi extract -i chr6.hprc-v1.0-pggb.og -d 1000000 -r grch38#chr6:31976719-32117146 -o - -t 48 -P | odgi sort -i - -o chr6.hprc-v1.0-pggb.RCCE.og -O
+
+# LPA
+odgi extract -i chr6.hprc-v1.0-pggb.og -d 1000000 -r grch38#chr6:160529904-160666180 -o - -t 48 -P | odgi sort -i - -o chr6.hprc-v1.0-pggb.LPA.og -O
+
+# AMY
+odgi extract -i chr1.hprc-v1.0-pggb.og -d 1000000 -r grch38#chr1:103542345-103798299 -o - -t 48 -P | odgi sort -i - -o chr1.hprc-v1.0-pggb.AMY.og -O
+
[…]

### 2023-07-17 manuscript/main.tex

@@
     \subsection{Performance}
-	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} to show the scalability and practicality of the algorithm with a real pangenome.
-	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads AMD EPYC 7343 processors with 512 GB RAM each) and 4 \texttt{HighMem} nodes (64 cores / 128 threads AMD EPYC 7513 processors with 2048 GB RAM each).
+	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm with a real pangenome.
+	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads with two AMD EPYC 7343 processors with 512 GB RAM each) and 4 \texttt{HighMem} nodes (64 cores / 128 threads with two AMD EPYC 7513 processors with 2048 GB RAM each).
 	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
@@
 	%We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
-	A special thanks goes to Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.
+	We thank Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.
 
@@
 	A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
-	JNM.S., J.L., and Z.Z. acknowledge funding from the NSF PPoSS Award \#2118709.
+	JNM.S., J.L., Z.Z., P.P., and E.G. acknowledge funding from the NSF PPoSS Award \#2118709.
 

### 2023-07-17 scripts/cfc/cfc_extracts_lay.sh
add some dirty scripts

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+singularity exec /nfsmounts/container/pg-sgd.img odgi layout -i /sfs/9/ws/afaaj01-sorting-paper/extracts/$1 -o /sfs/9/ws/afaaj01-sorting-paper/layouts/$1.lay -t 64 -C "$TMPDIR" -P

### 2023-07-17 scripts/cfc/cfc_sing.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+/usr/bin/time --verbose singularity exec /nfsmounts/container/pg-sgd.img odgi layout -i /sfs/9/ws/afaaj01-sorting-paper/graphs/chr"$1".hprc-v1.0-pggb.gfa -o /sfs/9/ws/afaaj01-sorting-paper/layouts/chr"$1".hprc-v1.0-pggb.gfa.lay -t 64 -C "$TMPDIR" 2> /sfs/9/ws/afaaj01-sorting-paper/times/odgi_layout_"$1".time

### 2023-07-17 scripts/cfc/cfc_slurm.sh

@@ -1,0 +1,52 @@
+# Allocate a workspace folder for 30 days
+# ws_allocate sorting-paper 30
+
+#mkdir -p /sfs/9/ws/afaaj01-sorting-paper/graphs
+cp cfc_sing* /sfs/9/ws/afaaj01-sorting-paper/
+cd /sfs/9/ws/afaaj01-sorting-paper/graphs
+#(seq 1 22; echo X; echo Y; echo M) | while read i; do echo $i; wget -c https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/freeze/freeze1/pggb/chroms/chr${i}.hprc-v1.0-pggb.gfa.gz; gunzip chr${i}.hprc-v1.0-pggb.gfa.gz; done
+
+cd ../
+#mkdir -p layouts
+#mkdir -p times
+#mkdir -p stats
+#(seq 1 22; echo X; echo Y; echo M) | while read i; do echo $i; sbatch -p qbic --mem=100000 -J pg-sgd -c 32 -t 600 --wrap "/usr/bin/time --verbose singularity exec /nfsmounts/container/pg-sgd.img odgi stats -i /sfs/9/ws/afaaj01-sorting-paper/graphs/chr${i}.hprc-v1.0-pggb.gfa -S -t 32 > /sfs/9/ws/afaaj01-sorting-paper/stats/odgi_stats_${i}.tsv"; done;
+#(seq 1 22; echo X; echo Y; echo M) | while read i; do echo $i; sbatch -p qbic --mem=100000 -J pg-sgd -c 64 -t 600 --wrap "./cfc_sing.sh $i"; done;
+
+#(seq 1 22; echo X; echo Y; echo M) | while read i; do echo $i; sbatch -p qbic --mem=100000 -J pg-sgd -c 64 -t 600 --wrap "/usr/bin/time --verbose singularity exec /nfsmounts/container/pg-sgd.img odgi build -g /sfs/9/ws/afaaj01-sorting-paper/graphs/chr${i}.hprc-v1.0-pggb.gfa -o /sfs/9/ws/afaaj01-sorting-paper/graphs/chr${i}.hprc-v1.0-pggb.og -t 64"; done;
+mkdir -p extracts
+#echo MHC
[…]

### 2023-07-19 scripts/container/pg-sgd_container.sh
add container scripts

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+docker build -t ${USER}/pg-sgd:latest .
+docker run -d -p 5000:5000 --name registry registry:2
+docker image tag $(docker images | grep pg-sgd | grep latest | grep ${USER} | tr -s ' ' | cut -f 3 -d ' ') localhost:5000/pg-sgd
+docker push localhost:5000/pg-sgd
+SINGULARITY_NOHTTPS=true singularity build --force pg-sgd.img docker://localhost:5000/pg-sgd
+docker container stop registry && docker container rm -v registry

### 2023-07-21 manuscript/main.tex
add 32T time odgi layout

@@
 	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm with a real pangenome.
-	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads with two AMD EPYC 7343 processors with 512 GB RAM each) and 4 \texttt{HighMem} nodes (64 cores / 128 threads with two AMD EPYC 7513 processors with 2048 GB RAM each).
+	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads with two AMD EPYC 7343 processors with 512 GB RAM) and 4 \texttt{HighMem} nodes (64 cores / 128 threads with two AMD EPYC 7513 processors with 2048 GB RAM).
 	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.

### 2023-07-21 manuscript/main.tex

@@
 	See Supplementary Table~\ref{tab:layout} for graph statistics.
-	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16 (Supplementary Table~\ref{tab:layout}).
+	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
 	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
 	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
-	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with chromosome 1, the biggest one in humans.
+	Memory consumption is 29.66 GB of RAM on average, with the memory peak again occurring with chromosome 16, presumably due to the memory expensive path index building phase.
 	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}).
-	Overall, PG-SGD is ~\FIXME{XX}X faster than BandageNG \FIXME{HOW TO CITE IT?}, the current state-of-the-art for graph visualization, while using ~{}X less memory.
+BandageNG\footnote{\url{https://github.com/asl/BandageNG} (last accessed Jul 2023)}, the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
+	Overall, PG-SGD is $\approx$8X faster on average than BandageNG while using $\approx$2X less memory.
+	% avg time bng 388 min, avg gb ng 66.84
 
@@
 	% Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full human graph.
-    % \\
+    % \\X
     % 1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}.
@@
 	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
-	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph}).
+	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent\footnote{\url{https://github.com/chfi/waragraph} (last accessed Jul 2023)}.
 
-	\FIXME{@Jiajie + Niklas: just 2/3 sentences about going GPU?}
+	\FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
 
@@
 		\subsection{Graph layouts}
-		\FIXME{chr1}
-		\FIXME{chr2}
-		\FIXME{...}
-		\FIXME{FIGURE WITH ALL CHROMOSOMES TOGETHER}
+		\label{sec:layouts}
+		\input{fig_2d_layouts}
+		\FIXME{1D layouts?}
 	\end{appendices}

### 2023-07-24 manuscript/fig_2d_layouts.tex

\begin{figure*}[!htb]
	\centering
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/2D_all_chroms.png}
		\label{fig:sfig1}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_annotated.png}
		\label{fig:sfig2}
	\end{subfigure}
	\\
\begin{subfigure}{1.0\textwidth}
	\centering
	\caption{}
	\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_genes_annotated.png}
	\label{fig:sfig3}
\end{subfigure}
	\\
\begin{subfigure}{1.0\textwidth}
	\centering
	\caption{}
	\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.C4_gene_annotated.png}
	\label{fig:sfig4}
\end{subfigure}	
	\caption{2D visualizations of all chromosomes of the Human Pangenome Resource Consortium (HPRC) Pangenome Graph Builder (PGGB) pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4). \textbf{(a)} \textit{odgi draw} layout of the HPRC pangenome graph 90 haplotypes. Displayed are all 24 autosomes and the mitochondrial chromosome. \textbf{(b)} \textit{gfaestus} screenshot of the chromosome 6 layout highlighting the MHC. \textbf{(c)} \textit{gfaestus} screenshot of the MHC highlighting all genes in this region. \textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically highlighting genes C4A and C4B.}
	\label{fig:2d_layouts}
\end{figure*}

### 2023-07-24 manuscript/main.tex

@@
     \subsection{Performance}
-	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm with a real pangenome.
+	We apply the 2D PG-SGD to the recent draft human pangenome reference \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm with a real pangenome.
 	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads with two AMD EPYC 7343 processors with 512 GB RAM) and 4 \texttt{HighMem} nodes (64 cores / 128 threads with two AMD EPYC 7513 processors with 2048 GB RAM).
@@
 	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}).
+Under one day a 2D layout was calculated.
 BandageNG\footnote{\url{https://github.com/asl/BandageNG} (last accessed Jul 2023)}, the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
-	Overall, PG-SGD is $\approx$8X faster on average than BandageNG while using $\approx$2X less memory.
+	On average, PG-SGD is $\approx$8X faster than BandageNG while using $\approx$2X less memory.
 	% avg time bng 388 min, avg gb ng 66.84
@@
 	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent. 
-	We show how PG-SGD allows us quickly gain insight into biological data by looking at the graph layout structure.
-	\FIXME{Describe a full chromosome and/or 1/a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the human paper or PGR-TK.}
+	We show how 2D PG-SGD allows us quickly gain insight into biological data by looking at the graph layout structure. 
+	To demonstrate this, we provide 2D drawings of various 2D PG-SGD layouts (see Figure~\ref{fig:2d_layouts})
+	In Figure~\ref{fig:sfig1} all chromosomes of the HPRC graph a drawn in 2D. 
+	One can easily spot all the large scale structural variations including the hairball centromeres. 
+	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:sfig2}), the 2D layout reveals the positions and complexity of all MHC genes (Figure~\ref{fig:sfig3}).
+In Figure~\ref{fig:sfig4} the C4A and C4B genes are highlighted.
+	The layout is exactly the same as the one produced by Bandage and displayed in~\citep{Guarracino2022}.
 
@@
 	Graph visualization is key for understanding genome variations and the large-scale layouts produced by PG-SGD offer an unprecedented high-level perspective on variation in pangenomes.
-	We implemented PG-SGD to generate layouts in 1 and 2 dimensions.
+	We implemented PG-SGD to generate layouts in 1D and 2D.
 	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of human acrocentric chromosomes \citep{Guarracino2023}.
@@
 	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent\footnote{\url{https://github.com/chfi/waragraph} (last accessed Jul 2023)}.
+
+	\FIXME{Talk about the performance evaluation here?}
 

### 2023-07-24 manuscript/fig_2d_layouts.tex
cleaner figure

@@
 		\caption{}
-		\includegraphics[width=1.0\linewidth]{fig/2D/2D_all_chroms.png}
+		\includegraphics[width=1.0\linewidth]{fig/2D/hprc-v1.0-pggb.og.lay.H4000.r4.0-highlighted.png}
 		\label{fig:sfig1}
@@
 \end{subfigure}	
-	\caption{2D visualizations of all chromosomes of the Human Pangenome Resource Consortium (HPRC) Pangenome Graph Builder (PGGB) pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4).
+	\caption{2D visualizations of all chromosomes of the Human Pangenome Resource Consortium (HPRC) 90 haplotypes pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4).
 \textbf{(a)} \textit{odgi draw} layout of the HPRC pangenome graph 90 haplotypes.
 Displayed are all 24 autosomes and the mitochondrial chromosome.
-\textbf{(b)} \textit{gfaestus} screenshot of the chromosome 6 layout highlighting the MHC.
-\textbf{(c)} \textit{gfaestus} screenshot of the MHC highlighting all genes in this region.
-\textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically highlighting genes C4A and C4B.}
+\textbf{(b)} \textit{gfaestus} screenshot of the chromosome 6 layout.
+Colored in blue is the MHC.
+\textbf{(c)} \textit{gfaestus} screenshot of the MHC.
+All MHC genes are color annotated.
+\textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically color highlighting genes C4A and C4B.}
 	\label{fig:2d_layouts}

### 2023-07-24 manuscript/main.tex
cleaner figure

@@
 
-	We implemented PG-SGD in ODGI \citep{Guarracino2022}. % which serves as an interface to graphs in the GFA (\url{https://github.com/GFA-spec/GFA-spec}) format. 
-	Our path index is a strict subset of the XG index~\citep{Garrison:2018} where we specifically avoid using succinct SDSL data structures \citep{Gog2014}.
+	We implemented PG-SGD in ODGI \citep{Guarracino2022}, specifically the 1D version in \textit{odgi sort} and the 2D version in \textit{odgi layout}.
+% which serves as an interface to graphs in the GFA (\url{https://github.com/GFA-spec/GFA-spec}) format.
+	Our path index is a strict subset of the XG index~\citep{Garrison:2018} where we avoid using succinct SDSL data structures \citep{Gog2014} in particular.
 	Instead, we only use bit compressed integer vectors, enabling efficient retrieval of path positions to quickly compute nucleotide distances without the need to store all pairwise distances between nodes in memory.
-	This allows PG-SGD to scale on large pangenome graphs representing thousands of whole genomes.
+	This allows our implementation to scale on large pangenome graphs representing thousands of whole genomes.
 
@@
 strategy \citep{Recht2011}: Threads perform layout updates without any locking for additional speed up.
-	This is possible because pangenome graphs are typically sparse \citep{Guarracino2022}, with low average node degree, meaning most updates only modify small parts of the entire layout.
+	This is possible because pangenome graphs are typically sparse \citep{Guarracino2022}, with low average node degree, meaning the updates only modify small parts of the entire layout.
 	The HOGWILD! SGD algorithm writes the layout updates to a shared non-atomic double vector, whereas PG-SGD stores node coordinates in a vector of atomic doubles.
@@
 Under one day a 2D layout was calculated.
-BandageNG\footnote{\url{https://github.com/asl/BandageNG} (last accessed Jul 2023)}, the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
+\textit{BandageNG} (\url{https://github.com/asl/BandageNG}, last accessed Jul 2023), the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
 	On average, PG-SGD is $\approx$8X faster than BandageNG while using $\approx$2X less memory.
@@
 In Figure~\ref{fig:sfig4} the C4A and C4B genes are highlighted.
-	The layout is exactly the same as the one produced by Bandage and displayed in~\citep{Guarracino2022}.
+	The layout is the same as the one produced by \textit{Bandage}~\citep{Wick_2015} and displayed in~\citep{Guarracino2022}.
 
@@
 	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
-	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent\footnote{\url{https://github.com/chfi/waragraph} (last accessed Jul 2023)}.
+	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph)} last accessed Jul 2023).
 
@@
 		\label{sec:layouts}
+		\FIXME{Talk about the 2D layouts in brief. What about the SMT thingy?}
+
+		\FIXME{What about 1D visualizations?}
+
 		\input{fig_2d_layouts}
+
 		\FIXME{1D layouts?}

### 2023-07-28 manuscript/main.tex
edits

@@
 	Originating from empirical inspection of word frequency tables, Zipf's law states that a word with rank $n$ occurs $1/n$ times as the most frequent one.
-	It's mathematical formulation is called the Zipfian distribution and its rank-frequency distribution is related to the inverse power law.
-	Sampling $s_j$ from a Zipfian distribution fixed in the $s_i$'s path position space increases the possibility to draw a nucleotide position close to $s_i$.
+    This law is modeled by the Zipf distribution, whose rank-frequency distribution is related to the inverse power law. % repetitive
+	Sampling $s_j$ from a Zipf distribution fixed in the $s_i$'s path position space increases the possibility to draw a nucleotide position close to $s_i$.
 	So there is a high chance to use small nucleotide distances $nd_{ij}$ to refine the layout of nodes comprising a few base pairs.
+    The Zipf distribution is also long tailed, with many occurrences of low frequency events, which ensures that longer-range structures are represented the output embedding space.
 
-	To provide balance between global and local layout updates, in half of the updates ($flip$ flag in Algorithm~\ref{alg:pg_sgd}) the $s_j$ is sampled uniformly instead from a Zipfian distribution (uniform sampling favours global updates).
+    While in principle, Zipf's long tail would lead to mixing of positional embedding information across entire chromosomes, practice demonstrated that very long range correlations were not captured sufficiently, leading to collapsed layouts for otherwise linear structures.
+    To provide balance between global and local layout updates, in half of the updates ($flip$ flag in Algorithm~\ref{alg:pg_sgd}) the $s_j$ is sampled uniformly instead from a Zipf distribution (uniform sampling favours global updates).
 	Additionally, to boost the local linearity (in 1D) or planarity (in 2D) of the graph layout, a \textit{cooling} phase skewes the Zipfian distribution after half the number of iterations, so that the possibility of a a small sampled nucleotide distance is significantly increased.
@@
 	Another possibility is to arrange nodes along a Hilbert curve, which often favors the creation of planar final layouts.
+    We additionally support fixing a set of node coordinates, such as those in a reference genome, in their order and position, which allows us to build reference-focused views of complex pangenomes.
+\FIXME{Is this only for 1D?
+It would be \textit{amazing} if it were possible to do in 2D, and it should be easy.
+-EG}
 
@@
 Explanation of 2D graph updating in Figures \ref{fig:2d_before_update}-\ref{fig:2d_after_update}.
-Zipfian distribution.
+Zipf distribution.
 

### 2023-07-28 manuscript/main.tex
feedback jiajie

@@
 	\FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
+	As can especially be observed in the chromosome 16 evaluation results, the implementation is memory-bound.
+The current data structure doesn't consider the memory access pattern.
+Therefore, a CPU cache optimized implementation can make use of the massive parallelized calculations on a GPU.
 

### 2023-07-28 manuscript/main.tex

@@
         Zhiru Zhang\,$^{\orcidlink{0000-0002-0778-0308}6}$,
+        Jörg Hagmann\,${\orcidlink{0000-0002-7232-3103}^7}$,
         Sven Nahnsen\,$^{\orcidlink{0000-0002-4375-0691}1,2}$,
@@
         $^6$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
+        $^7$Computomics GmbH, Eisenbahnstr. 1, 72072 Tübingen, Baden-Württemberg, Germany \\
     }
@@
 	Pangenome graphs can represent the full genetic diversity between multiple genomes, but their layouts may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
-	These structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
+	These structures hamper downstream analyses, visualization, and interpretation. \\
 	\textbf{Results:}
@@
 	such as gene annotation, read mapping, and variant detection \citep{Singh2022}.
-	However, with the availability of hundreds to thousands of high quality genomes per species, this linear model has become obsolete.
+	However, with the availability of hundreds to thousands of high quality genomes per species, this linear model will soon hit its limit.
 	A single genome can not faithfully represent the genetic diversity of a species, leading to the reference bias problem \citep{Ballouz2019}.
@@
 
-	A pangenome graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space such that a human accessible representation of genetic variation across multiple genomes can be drawn.
-	Layout algorithms aim to find suitable node coodinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote intuitive understanding by displaying the graph layout.
-	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetically-pleasing layouts.
+	A pangenome graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space such that a human readible representation of genetic variation across multiple genomes can be drawn. 
+	Layout algorithms aim to find suitable node coodinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding by displaying the graph layout.
+	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts.
 	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations can alleviate such a problem \citep{Zheng2019}.
-	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the HOGWILD!
-SGD is a highly parallelizable, therefore scalable solution that can be applied when the optimization problem is sparse \citep{Recht2011}.
+	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions. Specifically, in pangenome graphs we move a single pair of nodes at a time.
+
+	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the lock-free HOGWILD!
+is a highly parallelizable, therefore scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.
+
 	In practice, multidimensional scaling (MDS) is applied to minimize the difference between the visual distance and theoretical graph distance.
@@
 	PG-SGD's pseudocode is shown in Algorithm~\ref{alg:pg_sgd} and key steps are sketched in Figure \FIXME{Add puritan figure.}.
-	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
+	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a random path that traverses them.
 	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
 	$v_i$ is the node associated with the step $s_i$ sampled uniformly from all the steps in $\mathcal{P}$.
-	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by following a Zipfian distribution \citep{Zipf1932}.
+	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by drawing a Zipfian distribution \citep{Zipf1932}.
 	The difference between $nd_{ij}$ and $ld_{ij}$ guides the update of the node coordinates in the layout.
@@
     To provide balance between global and local layout updates, in half of the updates ($flip$ flag in Algorithm~\ref{alg:pg_sgd}) the $s_j$ is sampled uniformly instead from a Zipf distribution (uniform sampling favours global updates).
-	Additionally, to boost the local linearity (in 1D) or planarity (in 2D) of the graph layout, a \textit{cooling} phase skewes the Zipfian distribution after half the number of iterations, so that the possibility of a a small sampled nucleotide distance is significantly increased.
+	Additionally, to boost the local linearity (in 1D) or planarity (in 2D) of the graph layout, a \textit{cooling} phase skewes the Zipfian distribution after half the number of iterations, so that the possibility of a small sampled nucleotide distance is much increased.
 
@@
 % which serves as an interface to graphs in the GFA (\url{https://github.com/GFA-spec/GFA-spec}) format.
-	Our path index is a strict subset of the XG index~\citep{Garrison:2018} where we avoid using succinct SDSL data structures \citep{Gog2014} in particular.
+	For the efficient retrieval of path nucleotide positions, we implemented a path index. 
+	It is a strict subset of the XG index~\citep{Garrison:2018} where we avoid using succinct SDSL data structures \citep{Gog2014} in particular.
 	Instead, we only use bit compressed integer vectors, enabling efficient retrieval of path positions to quickly compute nucleotide distances without the need to store all pairwise distances between nodes in memory.
@@
 	In the 1D implementation, nodes are placed in the same order as they appear in the input graph by default, although support for random layout initialization is also provided.
-	This random placement ensures that each run of the PG-SGD method could potentially yield a different result, thereby increasing the likelihood of obtaining a meaningful layout.
+	%Random placement ensures that each run of the PG-SGD method could potentially yield a different result, thereby increasing the likelihood of obtaining a meaningful layout.
+@Andrea: What is it that you wanted to say here?
 	In 2D, we offer several layout initialization techniques.
@@
 	The HOGWILD! SGD algorithm writes the layout updates to a shared non-atomic double vector, whereas PG-SGD stores node coordinates in a vector of atomic doubles.
-	Such a vector ensures that no memory overwrites are possible. Our tests revealed basically no performance loss with respect to the non-atomic counterpart.
+	Such a vector prevents memory overwrites. Our tests revealed basically no performance loss with respect to the non-atomic counterpart.
 
@@
     \subsection{Performance}
-	We apply the 2D PG-SGD to the recent draft human pangenome reference \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm with a real pangenome.
+	We apply the 2D PG-SGD to the human pangenome \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm.
 	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads with two AMD EPYC 7343 processors with 512 GB RAM) and 4 \texttt{HighMem} nodes (64 cores / 128 threads with two AMD EPYC 7513 processors with 2048 GB RAM).
 	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
-	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
-	See Supplementary Table~\ref{tab:layout} for graph statistics.
+	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references (see Supplementary Table~\ref{tab:layout} for graph statistics).
 	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
-	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
+	This is expected since chromosome 16 has one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
 	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
@@
 	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}). 
-Under one day a 2D layout was calculated.
+	A 2D layout was calculated in under one day.
 	\textit{BandageNG} (\url{https://github.com/asl/BandageNG}, last accessed Jul 2023), the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
-	On average, PG-SGD is $\approx$8X faster than BandageNG while using $\approx$2X less memory.
+	On average, PG-SGD is $\sim$8X faster than BandageNG while using $\sim$2X less memory.
 	% avg time bng 388 min, avg gb ng 66.84
@@
 	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent. 
-	We show how 2D PG-SGD allows us quickly gain insight into biological data by looking at the graph layout structure. 
-	To demonstrate this, we provide 2D drawings of various 2D PG-SGD layouts (see Figure~\ref{fig:2d_layouts})
-	In Figure~\ref{fig:sfig1} all chromosomes of the HPRC graph a drawn in 2D. 
-	One can easily spot all the large scale structural variations including the hairball centromeres. 
+	We show how 2D PG-SGD allows us gaining insight into biological data by looking at the graph layout structure. 
+	To demonstrate this, we provide various 2D PG-SGD layouts (see Figure~\ref{fig:2d_layouts}).
+	In Figure~\ref{fig:sfig1} the chromosomes of the HPRC graph show the large scale structural variations including the hairball centromeres. 
 	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:sfig2}), the 2D layout reveals the positions and complexity of all MHC genes (Figure~\ref{fig:sfig3}). 
@@
 
-	Graph visualization is key for understanding genome variations and the large-scale layouts produced by PG-SGD offer an unprecedented high-level perspective on variation in pangenomes.
+	Graph visualization is key for understanding genome variations and the layouts produced by PG-SGD offer an unprecedented high-level perspective on variation in pangenomes.
 	We implemented PG-SGD to generate layouts in 1D and 2D.
@@
 	\section*{Competing interests}
-	The authors declare that they have no competing interests.
+	Author J.H. is employed by Computomics GmbH.
 

### 2023-07-28 manuscript/main.tex
clean

@@
 	So there is a high chance to use small nucleotide distances $nd_{ij}$ to refine the layout of nodes comprising a few base pairs.
-    The Zipf distribution is also long tailed, with many occurrences of low frequency events, which ensures that longer-range structures are represented the output embedding space.
+    The Zipf distribution is also long tailed, with many occurrences of low frequency events, which ensures that longer-range structures are represented in the output embedding space.
 
@@
 	In the 1D implementation, nodes are placed in the same order as they appear in the input graph by default, although support for random layout initialization is also provided.
-	%Random placement ensures that each run of the PG-SGD method could potentially yield a different result, thereby increasing the likelihood of obtaining a meaningful layout.
-@Andrea: What is it that you wanted to say here?
 	In 2D, we offer several layout initialization techniques.

### 2023-08-03 manuscript/fig_sketches.tex
fix merge conflict

--- 2022-09-22 manuscript/fig_algorithm.tex
@@
 		%\hline
-		\begin{subfigure}[t]{0.4\textwidth}
-			\centering
-			\caption{}
-			\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_metrics_explained.pdf}
-			\label{fig:metrics_explained}
-		\end{subfigure}
-		\begin{subfigure}[t]{0.4\textwidth}
-		\centering
-		\caption{}
-		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
-		\label{fig:sorting_explained}
-		\end{subfigure}
-		%\smallskip
-		\begin{subfigure}[t]{0.4\textwidth}
-			\centering
-			\caption{}
-			\includegraphics[width=\linewidth]{fig/algorithm/TODO_pg_sgd_update_sketch.png}
-			\label{fig:pg_sgd_update_sketch}
-		\end{subfigure}
-		\begin{subfigure}[t]{0.4\textwidth}
-			\centering
-			\caption{}
-			\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
-			\label{fig:fig:zipf_curve}
-		\end{subfigure}
+		\includegraphics[width=\linewidth, trim=0cm 8cm 0cm 0cm, clip]{fig/sketches/PG-SGD.drawio.pdf}
 	\caption{
-		Explain metrics and algorithms.
-		\textbf{(a)} Sketch of 1D sorting operation. \textbf{(b)} Sketch of metrics calculation. \textbf{(c)} Term updated sketch \textbf{(d)} Zipfian distribution.
+		2D PG-SGD update operation sketches. \\
+		\FIXME{ADD CAPTION.} \\ 
+		\FIXME{PROVIDE SEVERAL NICE FIGURES SO WE CAN REARRANGE STUFF INTO SUBFIGURES.}
 	}
-	\label{fig:algorithm}
+	\label{fig:sketches}
 \end{figure*}

### 2023-08-07 manuscript/fig_2d_layouts.tex
incorporate feedback from joerg
fix compilation issue

@@
 		Displayed are all 24 autosomes and the mitochondrial chromosome.
+		A red rectangle highlights chromosome 6 which is shown in the subfigure below. 
 		\textbf{(b)} \textit{gfaestus} screenshot of the chromosome 6 layout. 
 		Colored in blue is the MHC. 
+		The hairball in the middle is the centromere. 
+		The black structures in the centromere are edges. 
 		\textbf{(c)} \textit{gfaestus} screenshot of the MHC. 
-All MHC genes are color annotated.
-\textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically color highlighting genes C4A and C4B.}
+		All MHC genes are color annotated and the names of the genes appear as a text overlay. 
+		\textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically color highlighting genes C4A and C4B. 
+		The black lines are the edges of the graph.}
 	\label{fig:2d_layouts}

### 2023-08-07 manuscript/main.tex
fix compilation issue

@@
 
-	PG-SGD's pseudocode is shown in Algorithm~\ref{alg:pg_sgd} and key steps are sketched in Figure \FIXME{Add puritan figure.}.
-	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a random path that traverses them.
+	PG-SGD's pseudocode is shown in Algorithm~\ref{alg:pg_sgd} and 2D key steps are sketched in Figure~\ref{fig:sketches}.
+	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
+
 	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
@@
 	Additionally, to boost the local linearity (in 1D) or planarity (in 2D) of the graph layout, a \textit{cooling} phase skewes the Zipfian distribution after half the number of iterations, so that the possibility of a small sampled nucleotide distance is much increased.
+
+	\FIXME{In the section above we need to clarify that we move the ends of the nodes in 2D, and just the nodes in 1D.}
+
+	\input{fig_sketches}
 
@@
 	%We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
-	%We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
+	%We thank the authors of the pangenome resources made available xon GenBank which have made our experiments possible.
 	We thank Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.

### 2023-08-28 manuscript/fig_1d_sorts.tex
add 1D figure, cleaning

\begin{figure*}[!htpb]
	\centering
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth, trim=0cm 2.5cm 0cm 0cm]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.png}
		\label{fig:1d_fig1}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth, trim=0cm 2.5cm 0cm 0cm]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.png}
		\label{fig:1d_fig2}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth, trim=0cm 2.5cm 0cm 0cm]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.Y.og.png}
		\label{fig:1d_fig3}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth, trim=0cm 2.5cm 0cm 0cm]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og.png}
		\label{fig:1d_fig4}
	\end{subfigure}	
	\caption{\textit{odgi viz} 1D visualizations of a 5 haplotypes subgraph of the Human Pangenome Resource Consortium (HPRC) chromosome 6 pangenome graph. The major histocompatibility complex (MHC) sequence was injected as an extra path. Various node arrangements are shown. 
		\textbf{a-d} A graphs nodes are arranged from left to right forming the pangenome sequence. The black lines under the paths are the links representing the topology. Path names are left.
		\textbf{(a)} \textit{odgi viz} default modality: The colored bars are the paths versus pangenome sequence in a binary matrix. Shown is the subgraph extracted with \textit{odgi extract}. No 1D layout algorithm was applied here.
		\textbf{b-d} \textit{odgi viz} colored by path position. Light grey corresponds to the beginning of a path, black encodes the end of the path.
		\textbf{(b)} The nodes are arranged randomly in 1D.
		\textbf{(c)} The nodes are arranged applying the 1D PG-SGD algorithm.
		\textbf{(d)} The nodes are arranged applying a 1D reference-guided PG-SGD where the nodes of GRCh38 are fixed and only all the othere ones are movable in 1D.}
	\label{fig:1d_sorts}
\end{figure*}

### 2023-08-28 manuscript/main.tex
add 1D figure, cleaning

@@
 	We show how 2D PG-SGD allows us gaining insight into biological data by looking at the graph layout structure. 
-	To demonstrate this, we provide various 2D PG-SGD layouts (see Figure~\ref{fig:2d_layouts}).
-	In Figure~\ref{fig:sfig1} the chromosomes of the HPRC graph show the large scale structural variations including the hairball centromeres. 
-	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:sfig2}), the 2D layout reveals the positions and complexity of all MHC genes (Figure~\ref{fig:sfig3}). 
-	In Figure~\ref{fig:sfig4} the C4A and C4B genes are highlighted. 
+	To demonstrate this, we provide various 2D PG-SGD layouts in Figure~\ref{fig:2d_layouts}.
+	\input{fig_2d_layouts}
+	In Figure~\ref{fig:2d_fig1} the chromosomes of the HPRC graph show the large scale structural variations including the hairball centromeres. 
+	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:2d_fig2}), the 2D layout reveals the positions and complexity of all MHC genes (Figure~\ref{fig:2d_fig3}). 
+	In Figure~\ref{fig:2d_fig4} the C4A and C4B genes are highlighted. 
 	The layout is the same as the one produced by \textit{Bandage}~\citep{Wick_2015} and displayed in~\citep{Guarracino2022}.
+	Complementary, we provide various 1D visualizations of in Supplementary Figure~\ref{fig:1d_sorts}.
 
@@
 	Code and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/pangenome/sorting-paper}.
+	Animations of the algorithm are deposited at \url{https://doi.org/10.5281/zenodo.8288999}.
 
@@
 	    \section{Supplementary data}
-	    %\FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\
-	    %\FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y} \\
-		\FIXME{ADD LINKS TO OUR ANIMATIONS ON ZENODO} \\
-	    \FIXME{From 1D layout to 1D node order} \\
-	    \FIXME{reference-guided path-guided stochastic gradient descent}
+	    \subsection{Quantization}
+	    \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.}
 	    \subsection{Performance evaluation}
 	    \label{sec:performance}
+	    The results of the performance evaluation are given in Table~\ref{tab:layout}.
 	    \input{table_layout}
-		\subsection{Graph layouts}
-		\label{sec:layouts}
-		\FIXME{Talk about the 2D layouts in brief. What about the SMT thingy?}
-
-		\FIXME{What about 1D visualizations?}
-
-		\input{fig_2d_layouts}
-
-		\FIXME{1D layouts?}
+	    \FIXME{What about the SMT thingy?}
+		\subsection{1D visualizations}	
+		The 1D PG-SGD algorithm creates a 1D layout of the nodes of the graph.
+Theoretically, it is possible that 2 nodes have the same 1D coordinate or overlap.
+But, in our 1D visualizations, we arrange the nodes from left to right.
+Therefore, we project the 1D coordinates into a 1D node order: We sort the final layout by graph component, graph position, and node rank.
+		\input{fig_1d_sorts}
 	\end{appendices}

### 2023-08-28 scripts/1d.sh

@@ -1,0 +1,22 @@
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/freeze/freeze1/pggb/chroms/chr6.hprc-v1.0-pggb.gfa.gz
+gunzip chr6.hprc-v1.0-pggb.gfa.gz
+odgi build -g chr6.hprc-v1.0-pggb.gfa -o chr6.hprc-v1.0-pggb.og -P -t 28
+odgi paths -i chr6.hprc-v1.0-pggb.og -f > chr6.hprc-v1.0-pggb.fasta
+samtools faidx chr6.hprc-v1.0-pggb.fasta
+grep 'chm13#chr6\|grch38#chr6\|HG01928#2\|HG01071#2\|HG01106#2' chr6.hprc-v1.0-pggb.fasta.fai | cut -f 1 > 13_paths_to_extract.txt
+grep 'chm13#chr6\|grch38#chr6\|HG01928#2\|HG01071#2\|HG01106#2' chr6.hprc-v1.0-pggb.fasta.fai | cut -f 1,2 | sed 's/\t/\t0\t/g' > 13_paths_to_extract.bed
+odgi extract -i chr6.hprc-v1.0-pggb.og -o chr6.hprc-v1.0-pggb.13paths.og -E -b 13_paths_to_extract.bed -P -p 13_paths_to_extract.txt -t 28
+odgi viz -i chr6.hprc-v1.0-pggb.13paths.og -o chr6.hprc-v1.0-pggb.13paths.og.png
+bedtools merge -i /mnt/vol1/software/odgi/git/master/test/chr6.HLA_genes.bed -d 10000000 > MHC.bed
+echo -e '\tMHC' >> MHC.bed
+cat MHC.bed | tr '\n' ' ' | sed 's/ //g' > MHC.final.bed
+odgi procbed -i chr6.hprc-v1.0-pggb.13paths.og -b MHC.final.bed > MHC.adj.bed
+odgi inject -i chr6.hprc-v1.0-pggb.13paths.og -b MHC.adj.bed -o chr6.hprc-v1.0-pggb.13paths.MHC.og -P -t 28
+odgi viz -i chr6.hprc-v1.0-pggb.13paths.MHC.og -o chr6.hprc-v1.0-pggb.13paths.MHC.og.png
[…]

### 2023-08-28 manuscript/fig_sketches.tex
passthroug

@@
 	\caption{
-		2D PG-SGD update operation sketches. \\
-		\FIXME{ADD CAPTION.} \\ 
+		2D PG-SGD update operation sketches.
+\textbf{(a)} The path information of the graph which is going to be updated.
+Top: Path positions.
+Bottom: Path names with path sequences and the nucleotide distance between two consecutive nodes.
+\textbf{(b-e)} $v_i$ and $v_j$ is the current pair of nodes to update.
+$ld_{ij}$ is the current layout distance.
+$r,-r$ is the current size of the update.
+\textbf{(b)} Initial graph layout highlighting the future update of the two nodes of path \textit{P1}.
+\textbf{(c)} The graph layout after the first update.
+The nodes appear longer now, because we updated at the end of the nodes.
+Highlighted is the future update of the two nodes of path \textit{P2}.
+\textbf{(d)} The graph layout after the second update.
+Highlighted is the future update of the two nodes of \textit{P1}.
+\textbf{(e)} Final graph layout after three updates using the 2D PG-SGD.
+\\
 		\FIXME{PROVIDE SEVERAL NICE FIGURES SO WE CAN REARRANGE STUFF INTO SUBFIGURES.}

### 2023-08-28 manuscript/main.tex
passthroug

@@
 	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
+In 2D we model a layout node having two ends.
+When moving a pair of nodes we actually move one end of each node.
+For more details see Figure~\ref{fig:sketches}.
 
@@
 	Additionally, to boost the local linearity (in 1D) or planarity (in 2D) of the graph layout, a \textit{cooling} phase skewes the Zipfian distribution after half the number of iterations, so that the possibility of a small sampled nucleotide distance is much increased.
-
-	\FIXME{In the section above we need to clarify that we move the ends of the nodes in 2D, and just the nodes in 1D.}
-
 	\input{fig_sketches}
@@
 	Another possibility is to arrange nodes along a Hilbert curve, which often favors the creation of planar final layouts.
-    We additionally support fixing a set of node coordinates, such as those in a reference genome, in their order and position, which allows us to build reference-focused views of complex pangenomes.
-\FIXME{Is this only for 1D?
-It would be \textit{amazing} if it were possible to do in 2D, and it should be easy.
--EG}
+    We additionally support fixing a set of node coordinates, such as those in a reference genome, in their order and position, which allows us to build 1D reference-focused views of complex pangenomes.
+An exapmle visualization is given in Figure~\ref{fig:1d_fig4}.
+    \FIXME{Is this only for 1D? It would be \textit{amazing} if it were possible to do in 2D, and it should be easy. -EG} \\
+    \FIXME{I am afraid this only works for 1D. Or what do you make of \url{https://github.com/pangenome/odgi/pull/491}? -SH}
 
@@
 	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references (see Supplementary Table~\ref{tab:layout} for graph statistics).
-	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
+	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, \textit{odgi layout's} PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
 	This is expected since chromosome 16 has one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
@@
 
-	\FIXME{Talk about the performance evaluation here?}
-
 	\FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
+	The performance analysis shows that our implemented 2D PG-SGD outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
+While \textit{BandageNG} was not able to deliver a layout of the whole HPRC graph within 1 week, our 2D PG-SGD calculated one within one day.
+We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
 	As can especially be observed in the chromosome 16 evaluation results, the implementation is memory-bound.
 The current data structure doesn't consider the memory access pattern.
-Therefore, a CPU cache optimized implementation can make use of the massive parallelized calculations on a GPU.
+Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
+Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.
 
@@
 
-	\section*{Data availability}
+	\section*{Software and data availability}
 
-	Code and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/pangenome/sorting-paper}.
+	Software versios, code, and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/pangenome/sorting-paper}.
 	Animations of the algorithm are deposited at \url{https://doi.org/10.5281/zenodo.8288999}.

### 2023-08-28 manuscript/fig_1d_sorts.tex
update 1D figure

@@
 		\caption{}
-		\includegraphics[width=1.0\linewidth, trim=0cm 2.5cm 0cm 0cm]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.png}
+		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.png}
 		\label{fig:1d_fig1}
@@
 		\caption{}
-		\includegraphics[width=1.0\linewidth, trim=0cm 2.5cm 0cm 0cm]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.png}
+		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.png}
 		\label{fig:1d_fig2}
@@
 		\caption{}
-		\includegraphics[width=1.0\linewidth, trim=0cm 2.5cm 0cm 0cm]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.Y.og.png}
+		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.Y.og.png}
 		\label{fig:1d_fig3}
@@
 		\caption{}
-		\includegraphics[width=1.0\linewidth, trim=0cm 2.5cm 0cm 0cm]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og.png}
+		\includegraphics[width=1.0\linewidth]{fig/1D/chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og.png}
 		\label{fig:1d_fig4}
@@
 		\textbf{(c)} The nodes are arranged applying the 1D PG-SGD algorithm.
-		\textbf{(d)} The nodes are arranged applying a 1D reference-guided PG-SGD where the nodes of GRCh38 are fixed and only all the othere ones are movable in 1D.}
+		\textbf{(d)} The nodes are arranged applying a 1D reference-guided PG-SGD where the nodes of haplotype HG01071 are fixed and only all the othere ones are movable in 1D.
+Although, this is the optimal sort for this haplotype, log's of longer links are now visible compared to the visualization above.
+This indicates a node order that is globally not optimal.}
 	\label{fig:1d_sorts}

### 2023-08-28 scripts/1d.sh

@@ -17,6 +17,6 @@ odgi sort -i chr6.hprc-v1.0-pggb.13paths.MHC.og -r -P -t 28 -o chr6.hprc-v1.0-pg
 odgi viz -i chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og -o chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.png -u -d
 odgi sort -i chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og -o chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.Y.og -p Y -t 28 -P
 odgi viz -i chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.Y.og -o chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.Y.og.png -u -d
-odgi paths -i chr6.hprc-v1.0-pggb.13paths.MHC.og -L | grep "grch" > chr6.grch38.MHC.name
-odgi sort -i chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og -o chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og -P -t 28 -H chr6.grch38.MHC.name -Y
+odgi paths -i chr6.hprc-v1.0-pggb.13paths.MHC.og -L | grep "HG01071" > chr6.HG01071.MHC.name
+odgi sort -i chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og -o chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og -P -t 28 -H chr6.HG01071.MHC.name -Y
 odgi viz -i chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og -o chr6.hprc-v1.0-pggb.13paths.MHC.og.r.og.HY.og.png -d -u

### 2023-08-28 manuscript/fig_1d_sorts.tex
words

@@
 		\textbf{(d)} The nodes are arranged applying a 1D reference-guided PG-SGD where the nodes of haplotype HG01071 are fixed and only all the othere ones are movable in 1D.
-Although, this is the optimal sort for this haplotype, log's of longer links are now visible compared to the visualization above.
+Now all paths of this haplotype are arranged from their lowest to their highest nucleotide position.
+However,lot's of longer links are now visible compared to the node ordering directly above.
 This indicates a node order that is globally not optimal.}

### 2023-09-14 manuscript/fig_sketches.tex

--- 2023-08-03 manuscript/fig_sketches.tex
@@
-\begin{figure*}[!htb]
+\begin{figure}[!htb]
 	\centering
 		%\hline
-		\includegraphics[width=\linewidth, trim=0cm 8cm 0cm 0cm, clip]{fig/sketches/PG-SGD.drawio.pdf}
+		\includegraphics[width=0.8\linewidth, trim=0cm 0cm 0cm 0cm, clip]{fig/sketches/FigureToHeaven.pdf}
 	\caption{
-		2D PG-SGD update operation sketches. \\
-		\FIXME{ADD CAPTION.} \\ 
-		\FIXME{PROVIDE SEVERAL NICE FIGURES SO WE CAN REARRANGE STUFF INTO SUBFIGURES.}
+		2D PG-SGD update operation sketches.
+\textbf{(a)} The path information of the graph.
+\textit{path1} and \textit{path2} both visit the same first node.
+Then their sequence diverges and they visit distinct nodes.
+\textbf{(b-e)} $v_i$/$v_j$ or $v_i$/$v_k$ is the current pair of nodes to update.
+$ld_{ij}$/$ld_{ik}$ is the current layout distance.
+$r,-r$ is the current size of the update.
+\textbf{(b)} Initial graph layout highlighting the future update of the two nodes of \textit{path1}.
+\textbf{(c)} The graph layout after the first update.
+The nodes appear longer now, because we updated at the end of the nodes.
+Highlighted is the future update of the two nodes of \textit{path2}.
+\textbf{(d)} The graph layout after the second update.
+Highlighted is the future update of the two nodes of \textit{path1}.
+\textbf{(e)} Final graph layout after three updates using the 2D PG-SGD.
 	}
 	\label{fig:sketches}
-\end{figure*}
+\end{figure}

### 2023-09-14 manuscript/main.tex

@@
 	Another possibility is to arrange nodes along a Hilbert curve, which often favors the creation of planar final layouts.
-    We additionally support fixing a set of node coordinates, such as those in a reference genome, in their order and position, which allows us to build 1D reference-focused views of complex pangenomes.
-An exapmle visualization is given in Figure~\ref{fig:1d_fig4}.
-    \FIXME{Is this only for 1D? It would be \textit{amazing} if it were possible to do in 2D, and it should be easy. -EG} \\
-    \FIXME{I am afraid this only works for 1D. Or what do you make of \url{https://github.com/pangenome/odgi/pull/491}? -SH}
+    We additionally support fixing a set of node coordinates, such as those in a reference genome, in their order and position, which allows us to build reference-focused views of complex pangenomes.
+An example visualization is given in Figure~\ref{fig:1d_fig4}.
 

### 2023-09-15 manuscript/main.tex

@@
         Jan-Niklas Manuel Schmelzle\,$^{\orcidlink{0000-0001-8566-4049}5,6}$,
-        Jiajie Li\,$^{\orcidlink{0000-0002-9467-0659}6}$,
+        Jiajie Li\,$^{\orcidlink{0000-0002-6775-2843}6}$,
         Zhiru Zhang\,$^{\orcidlink{0000-0002-0778-0308}6}$,
@@
 	To address these challenges, we propose a novel graph layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-	PG-SGD uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel by applying a modified HOWGWILD! strategy.
+	PG-SGD uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel by applying a modified HOGWILD! strategy.
 	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, revealing their biological features. \\
@@
 	A pangenome graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space such that a human readible representation of genetic variation across multiple genomes can be drawn. 
-	Layout algorithms aim to find suitable node coodinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding by displaying the graph layout.
+	Layout algorithms aim to find suitable node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding by displaying the graph layout.
 	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts.

### 2023-09-18 manuscript/main.tex
more german, more italian

@@
 \usepackage{float}
+\usepackage{hyperref}
 \floatstyle{plaintop}
@@
 
-    %\title[Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
     \title[Pangenome graph layout by Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by Path-Guided Stochastic Gradient Descent}
-    %\title[Unveiling genome variation by sequence-guided pangenome graph layouts]{Unveiling genome variation by sequence-guided pangenome graph layouts}
-    %\title[PG-SGD: sequence-guided pangenome graph layouts]{PG-SGD: sequence-guided pangenome graph layouts}
     %\title[Genome-guided pangenome graph layouts via Stochastic Gradient Descent]{Genome-guided pangenome graph layouts via Stochastic Gradient Descent}
@@
 	\textbf{Motivation:}
-	The increasing availability of complete genomes demands models to study genomic variability within entire populations.
-	Pangenome graphs can represent the full genetic diversity between multiple genomes, but their layouts may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
+	The increasing availability of complete genomes demands for models to study genomic variability within entire populations.
+	Pangenome graphs capture the full genetic diversity between multiple genomes, but their layouts may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
 	These structures hamper downstream analyses, visualization, and interpretation. \\
 	\textbf{Results:}
-	To address these challenges, we propose a novel graph layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
-	PG-SGD uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel by applying a modified HOGWILD! strategy.
-	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, revealing their biological features. \\
+	In response, we introduce a novel graph layout algorithm: the Path-Guided Stochastic Gradient Descent (PG-SGD).
+	PG-SGD uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel applying a modified HOGWILD! strategy.
+	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, unveiling their biological features. \\
 	\textbf{Availability:}
-	PG-SGD has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
+	We integrated PG-SGD in \textit{ODGI} which is released as free software under the MIT open source license.
 	Source code is available at \url{https://github.com/pangenome/odgi}. \\% and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
@@
 	\section{Introduction}
-	Reference genomes are the most widely used resource in genetics, serving as a foundation for a variety of analyses
-	such as gene annotation, read mapping, and variant detection \citep{Singh2022}.
-	However, with the availability of hundreds to thousands of high quality genomes per species, this linear model will soon hit its limit.
-	A single genome can not faithfully represent the genetic diversity of a species, leading to the reference bias problem \citep{Ballouz2019}.
-	Rather, a pangenome models the entire set of genomic elements of a given population \citep{Tettelin_2008,cpang2018,Eizenga_2020, Sherman_2020}.
-	Pangenomes can be represented by a sequence graph incorporating sequences as nodes and their connections as edges \citep{Hein1989}.
-	These nodes are shared for identical sequences, such as homologs, paralogs, and orthologs.
-	In the variation graph model \citep{Garrison:2018}, genomes are encoded as paths visiting the nodes in the graph.
+	Reference genomes are widely used in genetics, serving as a foundation for a variety of analyses, including gene annotation, read mapping, and variant detection \citep{Singh2022}.
+	However, this linear model is becoming obsolete given the accessibility to hundreds or even thousands of high-quality genomes.
+	A single genome can not fully represent the genetic diversity of any species, resulting in reference bias \citep{Ballouz2019}.
+	In contrast, a pangenome models the entire set of genomic elements of a given population \citep{Tettelin_2008,cpang2018,Eizenga_2020, Sherman_2020}.
+	Pangenomes can be represented as a sequence graph incorporating sequences as nodes and their relationships as edges \citep{Hein1989}.
+	%These nodes are shared for identical sequences, such as homologs, paralogs, and orthologs.
+	In the variation graph model \citep{Garrison:2018}, genomes are encoded as paths traversing the nodes in the graph.
 	%A bidirected graph can even contain both strands of DNA as paths. \\
 
-	A pangenome graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space such that a human readible representation of genetic variation across multiple genomes can be drawn. 
-	Layout algorithms aim to find suitable node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding by displaying the graph layout.
+	A pangenome graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space to produce a human-readable visualization of genetic variation between multiple genomes.
+	Layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive graph understanding.
 	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts.
-	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations can alleviate such a problem \citep{Zheng2019}.
-	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions. Specifically, in pangenome graphs we move a single pair of nodes at a time.
+	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
+	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.
+	%Specifically, in pangenome graphs we move a single pair of nodes at a time.
 
 	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the lock-free HOGWILD!
-is a highly parallelizable, therefore scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.
+method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.
 
@@
 	This can be accomplished by using pairwise node distances to minimize an energy function.
-	%\FIXME{
-	%	Existing layout algorithms range from simple, force-directed methods to more complex, hierarchical layouts, each presenting unique strengths and drawbacks.
-	%	Force-directed methods offer an aesthetic layout based on a physics simulation but may struggle with scalability for larger graphs (XXX WE NEED CITATIONS XXX).
-	%	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
-	%}
 	Since pangenome graphs represent genomes as paths in the graph, a reasonable distance metric would be the nucleotide distance between a pair of nodes traversed by the same path.
-	To our knowledge, no algorithm takes into account such biological information to compute the pangenome graph layouts.
 
@@
 	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
-	PG-SGD can be extended in any number of dimensions. We provide in the ODGI toolkit \citep{Guarracino2022} implementations to produce layouts in 1 dimension (1D) and 2 dimensions (2D).
-	% We may want to adjust some citations using https://tex.stackexchange.com/questions/585942/how-to-display-the-first-two-authors-in-a-citation-call-out.
+	To our knowledge, no algorithm takes into account such biological information to compute the graph layouts.
+	PG-SGD can be extended in any number of dimensions.
+	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.
 
 	\section{Algorithm}
-	%\FIXME{WE ONLY WANT TO TALK ABOUT LAYOUTS AND PRODUCING LAYOUTS, NOT ABOUT GRAPH SORTING.
-PG-SGD PRODUCES 1D AND 2D LAYOUTS.
-IN 1D WE SORT A GRAPH BY PROJECTING THE ACTUAL 1D LAYOUT INTO A NODE ORDER!}
 
-	PG-SGD is inspired by \cite{Zheng2019}, but we designed the algorithm to work on pangenomes in the form of a variation graph (Definition \ref{def:vg}).
+	While PG-SGD is inspired by \cite{Zheng2019}, we designed the algorithm to work on the variation graph model (Definition \ref{def:vg}).
 
@@
 		Each node $v_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{v_i}$.
-		The node strand $o$ represents node orientation.
+		The node strand $o$ represents the node orientation.
 		Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands ($e_i = ( o_a, o_b )$), defining the graph topology.
@@
 
-	PG-SGD's pseudocode is shown in Algorithm~\ref{alg:pg_sgd} and 2D key steps are sketched in Figure~\ref{fig:sketches}.
+	We report PG-SGD's pseudocode in Algorithm~\ref{alg:pg_sgd} and its schematic in Figure~\ref{fig:sketches}.
 	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.
-In 2D we model a layout node having two ends.
-When moving a pair of nodes we actually move one end of each node.
-For more details see Figure~\ref{fig:sketches}.
-
-	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
+	In the 2D layouts, nodes have two ends.
+	When moving a pair of nodes, we actually move one end of each node.
 	$v_i$ is the node associated with the step $s_i$ sampled uniformly from all the steps in $\mathcal{P}$.
@@
 	The difference between $nd_{ij}$ and $ld_{ij}$ guides the update of the node coordinates in the layout.
-	The magnitude $r$ of the update depends on the learning rate $\mu$ of the SGD.
+	The magnitude $r$ of the update depends on the learning rate $\mu$.
 	The number of iterations steers the annealing step size $\eta$ which determines the learning rate $\mu$.
@@
 	A large $\eta$ in the first iterations leads to a globally linear (in 1D) or planar (in 2D) layout.
-	By decreasing $\eta$, the layout adjustments become more localized, ensuring that the nodes are positioned in a way to best reflect the nucleotide distances in the paths (i.e., in the genomes).
-% LIAR: down to the single base pair level.
+	By decreasing $\eta$, the layout adjustments become more localized, ensuring that the nodes are positioned to best reflect the nucleotide distances in the paths (i.e., in the genomes).
 
@@
 	%In standard SGD layout algorithms, the problem space is given by pairwise node comparisons.
-In PG-SGD, the problem space is given by the pairwise path step comparisons.
-The number of path steps in large variation graphs can be several orders of magnitude higher than the number of nodes.
-Therefore, the number of term updates in PG-SGD is bound to the number of path steps in the graph.
+	%In PG-SGD, the problem space is given by the pairwise path step comparisons.
+	%The number of path steps in large variation graphs can be several orders of magnitude higher than the number of nodes.
+	%Therefore, the number of term updates in PG-SGD is bound to the number of path steps in the graph.
 	%Should the delta of $nd_{ij}$ and $ld_{ij}$ become 0 before the annealing schedule is finished, the algorithm stops.
@@
 	Originating from empirical inspection of word frequency tables, Zipf's law states that a word with rank $n$ occurs $1/n$ times as the most frequent one.
-    This law is modeled by the Zipf distribution, whose rank-frequency distribution is related to the inverse power law. % repetitive
+    This law is modeled by the Zipf distribution.%, whose rank-frequency distribution is related to the inverse power law. % repetitive
 	Sampling $s_j$ from a Zipf distribution fixed in the $s_i$'s path position space increases the possibility to draw a nucleotide position close to $s_i$.
 	So there is a high chance to use small nucleotide distances $nd_{ij}$ to refine the layout of nodes comprising a few base pairs.
-    The Zipf distribution is also long tailed, with many occurrences of low frequency events, which ensures that longer-range structures are represented in the output embedding space.
+    The Zipf distribution is also long-tailed, with many occurrences of low frequency events.
+	However, extremely long-range correlations might not be captured sufficiently, resulting in collapsed layouts for structures that are otherwise linear.
+    To provide balance between global and local layout updates, in half of the updates ($flip$ flag in Algorithm~\ref{alg:pg_sgd}), the $s_j$ is sampled uniformly instead from a Zipf distribution, with uniform sampling being more favorable for global updates.
+	Furthermore, to enhance local linearity (in 1D) or planarity (in 2D) of the graph layout, a \textit{cooling} phase skews the Zipfian distribution after half of iterations have been completed.
+	This increases the likelihood of sampling smaller nucleotide distances for the layout updates.
 
-    While in principle, Zipf's long tail would lead to mixing of positional embedding information across entire chromosomes, practice demonstrated that very long range correlations were not captured sufficiently, leading to collapsed layouts for otherwise linear structures.
-    To provide balance between global and local layout updates, in half of the updates ($flip$ flag in Algorithm~\ref{alg:pg_sgd}) the $s_j$ is sampled uniformly instead from a Zipf distribution (uniform sampling favours global updates).
-	Additionally, to boost the local linearity (in 1D) or planarity (in 2D) of the graph layout, a \textit{cooling} phase skewes the Zipfian distribution after half the number of iterations, so that the possibility of a small sampled nucleotide distance is much increased.
 	\input{fig_sketches}
@@
 
-	We implemented PG-SGD in ODGI \citep{Guarracino2022}, specifically the 1D version in \textit{odgi sort} and the 2D version in \textit{odgi layout}.
+	We implemented PG-SGD in ODGI \citep{Guarracino2022}: the 1D version can be found in \textit{odgi sort} and the 2D version in \textit{odgi layout}.
 % which serves as an interface to graphs in the GFA (\url{https://github.com/GFA-spec/GFA-spec}) format.
-	For the efficient retrieval of path nucleotide positions, we implemented a path index. 
-	It is a strict subset of the XG index~\citep{Garrison:2018} where we avoid using succinct SDSL data structures \citep{Gog2014} in particular.
-	Instead, we only use bit compressed integer vectors, enabling efficient retrieval of path positions to quickly compute nucleotide distances without the need to store all pairwise distances between nodes in memory.
-	This allows our implementation to scale on large pangenome graphs representing thousands of whole genomes.
+	To efficiently retrieve path nucleotide positions, we implemented a path index.
+	This index is a strict subset of the XG index~\citep{Garrison:2018} where we avoid to use succinct SDSL data structures \citep{Gog2014}.
+	Instead, we rely on bit-compressed integer vectors, enabling efficient retrieval of path nucleotide positions to quickly compute nucleotide distances without having to store all pairwise distances between nodes in memory.
+	This approach ensures to scale on large pangenome graphs representing thousands of whole genomes.
 
 	Graph layout initialization can significantly influence the quality of the final layout.
-	In the 1D implementation, nodes are placed in the same order as they appear in the input graph by default, although support for random layout initialization is also provided.
+	In the 1D implementation, by default, nodes are placed in the same order as they appear in the input graph, although we also provide support for random layout initialization.
 	In 2D, we offer several layout initialization techniques.
-	One option is to position nodes in the first layout dimension according to their order in the input graph and to add either uniform or Gaussian noise in the second dimension.
-	Another possibility is to arrange nodes along a Hilbert curve, which often favors the creation of planar final layouts.
-    We additionally support fixing a set of node coordinates, such as those in a reference genome, in their order and position, which allows us to build reference-focused views of complex pangenomes.
-An example visualization is given in Figure~\ref{fig:1d_fig4}.
+	One approach places nodes in the first layout dimension according to their order in the input graph, adding either uniform or Gaussian noise in the second dimension.
+	Another strategy arranges nodes along a Hilbert curve, an approach that often favors the creation of planar final layouts.
+	We also support fixing node positions to keep nodes in the same order as they are in a selected path, such as a reference genome.
+	This feature allows us to build reference-focused graph layouts (Figure~\ref{fig:1d_fig4}).
 
-	Our implementation is multithreaded and uses shared memory for the layout vector following the HOGWILD!
-strategy \citep{Recht2011}: Threads perform layout updates without any locking for additional speed up.
-	This is possible because pangenome graphs are typically sparse \citep{Guarracino2022}, with low average node degree, meaning the updates only modify small parts of the entire layout.
-	The HOGWILD! SGD algorithm writes the layout updates to a shared non-atomic double vector, whereas PG-SGD stores node coordinates in a vector of atomic doubles.
-	Such a vector prevents memory overwrites. Our tests revealed basically no performance loss with respect to the non-atomic counterpart.
+	Our implementation is multithreaded and uses shared memory for storing the layout in a vector, according to the HOGWILD! strategy \citep{Recht2011}.
+	Threads perform layout updates without any locking for additional speed up.
+	This approach is feasible since pangenome graphs are typically sparse \citep{Guarracino2022}, with low average node degree.
+	As a result, the updates only modify small parts of the entire layout.
+	While the HOGWILD! SGD algorithm writes the layout updates to a shared non-atomic double vector, PG-SGD stores node coordinates in a vector of atomic doubles.
+	This vector prevents any potential memory overwrites.
+	Our tests revealed basically no performance loss with respect to the non-atomic counterpart.
 
-    \iffalse
-    Fig.
+	%Fig.
 1: Describe how our approach works, especially a single update operation (Fig.
@@
 Zipf distribution.
-
-    \input{fig_sketches}
-    \fi
+	%\input{fig_sketches}
 
@@
 	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references (see Supplementary Table~\ref{tab:layout} for graph statistics).
-	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, \textit{odgi layout's} PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
+	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, 2D PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
 	This is expected since chromosome 16 has one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
 	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
-	Memory consumption is 29.66 GB of RAM on average, with the memory peak again occurring with chromosome 16, presumably due to the memory expensive path index building phase.
+	Memory consumption is 29.66 GB of RAM on average, with the memory peak again occurring with chromosome 16, due to the path index building phase.
 	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}). 
-	A 2D layout was calculated in under one day.
+	%A 2D layout was calculated in under one day.
 	\textit{BandageNG} (\url{https://github.com/asl/BandageNG}, last accessed Jul 2023), the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
@@
 	We show how 2D PG-SGD allows us gaining insight into biological data by looking at the graph layout structure. 
-	To demonstrate this, we provide various 2D PG-SGD layouts in Figure~\ref{fig:2d_layouts}.
-	\input{fig_2d_layouts}
-	In Figure~\ref{fig:2d_fig1} the chromosomes of the HPRC graph show the large scale structural variations including the hairball centromeres. 
-	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:2d_fig2}), the 2D layout reveals the positions and complexity of all MHC genes (Figure~\ref{fig:2d_fig3}). 
+	%To demonstrate this, we provide various 2D PG-SGD layouts in Figure~\ref{fig:2d_layouts}.
+
+	In Figure~\ref{fig:2d_fig1}, the chromosomes of the HPRC graph show the large scale structural variations in the centromeres.
+	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:2d_fig2}), the 2D layout reveals the positions and diversity of all MHC genes (Figure~\ref{fig:2d_fig3}).
 	In Figure~\ref{fig:2d_fig4} the C4A and C4B genes are highlighted. 
-	The layout is the same as the one produced by \textit{Bandage}~\citep{Wick_2015} and displayed in~\citep{Guarracino2022}.
+	%The layout is the same as the one produced by \textit{Bandage}~\citep{Wick_2015} and displayed in~\citep{Guarracino2022}.
 	Complementary, we provide various 1D visualizations of in Supplementary Figure~\ref{fig:1d_sorts}.
@@
 
-	Graph visualization is key for understanding genome variations and the layouts produced by PG-SGD offer an unprecedented high-level perspective on variation in pangenomes.
+	Graph visualization is key for understanding genome variations and the layouts produced by PG-SGD offer an unprecedented high-level perspective on pangenome variation.
 	We implemented PG-SGD to generate layouts in 1D and 2D.
@@
 	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
-	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph)} last accessed Jul 2023).
+	Our algorithm will underpin new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph)} last accessed Jul 2023).
 
 	\FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
-	The performance analysis shows that our implemented 2D PG-SGD outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
+	The performance analysis shows that our 2D implementation outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
 	While \textit{BandageNG} was not able to deliver a layout of the whole HPRC graph within 1 week, our 2D PG-SGD calculated one within one day.
-We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
-	As can especially be observed in the chromosome 16 evaluation results, the implementation is memory-bound.
+	%We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
+	As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
 	The current data structure doesn't consider the memory access pattern.
@@
 
-	%We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
-	%We thank the authors of the pangenome resources made available xon GenBank which have made our experiments possible.
 	We thank Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.
@@
 
-	Software versios, code, and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/pangenome/sorting-paper}.
+	Software versions, code, and links to data resources used to prepare this manuscript can be found at \url{https://github.com/pangenome/sorting-paper}.
 	Animations of the algorithm are deposited at \url{https://doi.org/10.5281/zenodo.8288999}.
@@
 	    \section{Supplementary data}
-	    \subsection{Quantization}
-	    \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.}
+		\input{fig_2d_layouts}
+	    %\subsection{Quantization} % DURING REVISION, IF NEEDED
+	    %More details about your clever quantization.
 	    \subsection{Performance evaluation}

### 2023-09-18 manuscript/main.tex
move 2D fig into main part

@@
 \restylefloat{table}
+
+% supplement stuff
+\newcommand{\beginsupplement}{%
+	\setcounter{table}{0}
+	\renewcommand{\thetable}{S\arabic{table}}%
+	\setcounter{figure}{0}
+	\renewcommand{\thefigure}{S\arabic{figure}}%
+}
 
@@
 	%To demonstrate this, we provide various 2D PG-SGD layouts in Figure~\ref{fig:2d_layouts}.
-
+			\input{fig_2d_layouts}
 	In Figure~\ref{fig:2d_fig1}, the chromosomes of the HPRC graph show the large scale structural variations in the centromeres.
@@
 
-	\begin{appendices}
-	    \section{Supplementary data}
-		\input{fig_2d_layouts}
+	\clearpage
+	\setcounter{page}{1}
+
+	\beginsupplement
+
+	\section{Supplement}
+
+	 \subsection{Supplementary data}
 	 %\subsection{Quantization} % DURING REVISION, IF NEEDED
 	 %More details about your clever quantization.
-	    \subsection{Performance evaluation}
+	 \subsubsection{Performance evaluation}
 	 \label{sec:performance}
@@
 	 \input{table_layout}
-	    \FIXME{What about the SMT thingy?}
-		\subsection{1D visualizations}	
+	 %\FIXME{What about the SMT thingy?} % DURING REVISION, IF NEEDED
+	\subsubsection{1D visualizations}	
 	The 1D PG-SGD algorithm creates a 1D layout of the nodes of the graph.
@@
 	\input{fig_1d_sorts}
-	\end{appendices}
 

### 2023-09-20 manuscript/main.tex
cosmetics

@@
         $^6$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
-        $^7$Computomics GmbH, Eisenbahnstr. 1, 72072 Tübingen, Baden-Württemberg, Germany \\
+        $^7$Computomics GmbH, Eisenbahnstr. 1, 72072 Tübingen, Germany \\
     }
@@
 
-	\FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
+	% \FIXME{@Jiajie + Niklas: Maybe address our SMT issue with your cache locality as a potential solution. Additionally just 2/3 sentences about going GPU?}
 	The performance analysis shows that our 2D implementation outperforms \textit{BandageNG} when handling large, complex pangenome graphs.
@@
 	%We expect that our implementation will be able to handle the next phase of the HPRC, 24 chromosomal pangenome graphs constructed from 300 individuals, without any problems.
-	As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
-	The current data structure doesn't consider the memory access pattern.
-	Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
-	Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.
+	There are some possible optimization approaches for future work to further improve the performance of PG-SGD, making it possible for interactive use. 
+	The data structure could be optimized to improve ache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU.  
+	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
+	% The current data structure doesn't consider the memory access pattern.
+	% Therefore, a CPU cache optimized implementation could make use of the massive parallelized calculations on a GPU.
+	% Such efforts are currently in progress \url{https://github.com/nsmlzl/odgi/tree/layout_cache_optimized_cpu}.
 
@@
 	This enables the application of machine learning algorithms that use the graph layout for variant detection and classification. 
-	Our future research involves leveraging these graph projections to detect structural variants and to identify and correct assembly errors
+	Our future research involves leveraging these graph projections to detect structural variants and to identify and correct assembly errors. 
 	Moreover, we are considering extending the algorithm to RNA and protein sequences to support pantranscriptome graphs \citep{sibbesen_haplotype-aware_2023} and panproteome graphs \citep{dabbaghie_panpa:_2023}, respectively.

### 2023-09-22 manuscript/main.tex
typo
integrate Pjotr's feedback

@@
 	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.
+The respective algorithms were already successfully applied to build and visualize large scale pangenome graphs of the Human Pangenome Resource Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.
 
@@
 	There are some possible optimization approaches for future work to further improve the performance of PG-SGD, making it possible for interactive use. 
-	The data structure could be optimized to improve ache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU.  
+	The data structure could be optimized to improve cache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU.  
 	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.
@@
 	A.G. acknowledges efforts by Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy.
-	JNM.S., J.L., Z.Z., P.P., and E.G. acknowledge funding from the NSF PPoSS Award \#2118709.
+	JNM.S., J.L., Z.Z., P.P., and E.G.
+acknowledge funding from the NSF PPoSS Award \#2118709.
+The authors gratefully acknowledge support from National Institutes of Health/NIDA U01DA047638 (E.G.), National Institutes of Health/NIGMS R01GM123489 (E.G.
+and P.P.)
 

### 2023-09-22 manuscript/fig_1d_sorts.tex
each sentence in one line
integrate feedback from Niklas

@@
 	\end{subfigure}	
-	\caption{\textit{odgi viz} 1D visualizations of a 5 haplotypes subgraph of the Human Pangenome Resource Consortium (HPRC) chromosome 6 pangenome graph.
+	\caption{\textit{odgi viz} 1D visualizations of a 5 haplotypes subgraph of the Human Pangenome Reference Consortium (HPRC) chromosome 6 pangenome graph.
 The major histocompatibility complex (MHC) sequence was injected as an extra path.
 Various node arrangements are shown.
-		\textbf{a-d} A graphs nodes are arranged from left to right forming the pangenome sequence. The black lines under the paths are the links representing the topology. Path names are left.
+		\textbf{a-d} A graphs nodes are arranged from left to right forming the pangenome sequence. The black lines under the paths are the links representing the topology. 
+		Path names are left.
 		\textbf{(a)} \textit{odgi viz} default modality: The colored bars are the paths versus pangenome sequence in a binary matrix. 
-Shown is the subgraph extracted with \textit{odgi extract}.
-No 1D layout algorithm was applied here.
-		\textbf{b-d} \textit{odgi viz} colored by path position. Light grey corresponds to the beginning of a path, black encodes the end of the path.
+		Shown is the subgraph extracted with \textit{odgi extract}. No 1D layout algorithm was applied here.
+		\textbf{b-d} \textit{odgi viz} colored by path position. 
+		Light grey corresponds to the beginning of a path, black encodes the end of the path.
 		\textbf{(b)} The nodes are arranged randomly in 1D.
@@
 		Now all paths of this haplotype are arranged from their lowest to their highest nucleotide position. 
-However,lot's of longer links are now visible compared to the node ordering directly above.
-This indicates a node order that is globally not optimal.}
+		However,lot's of longer links are now visible compared to the node ordering directly above. This indicates a node order that is globally not optimal.}
 	\label{fig:1d_sorts}

### 2023-09-22 manuscript/fig_2d_layouts.tex
add 1D figure, cleaning
integrate feedback from Niklas

@@
-\begin{figure*}[!htb]
+\begin{figure*}[!htpb]
 	\centering
@@
 		\includegraphics[width=1.0\linewidth]{fig/2D/hprc-v1.0-pggb.og.lay.H4000.r4.0-highlighted.png}
-		\label{fig:sfig1}
+		\label{fig:2d_fig1}
 	\end{subfigure}
@@
 		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_annotated.png}
-		\label{fig:sfig2}
+		\label{fig:2d_fig2}
 	\end{subfigure}
@@
 		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_genes_annotated.png}
-		\label{fig:sfig3}
+		\label{fig:2d_fig3}
 	\end{subfigure}
@@
 		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.C4_gene_annotated.png}
-		\label{fig:sfig4}
+		\label{fig:2d_fig4}
 	\end{subfigure}	
-	\caption{2D visualizations of all chromosomes of the Human Pangenome Resource Consortium (HPRC) 90 haplotypes pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4).
+	\caption{2D visualizations of all chromosomes of the Human Pangenome Reference Consortium (HPRC) 90 haplotypes pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4).
 		\textbf{(a)} \textit{odgi draw} layout of the HPRC pangenome graph 90 haplotypes. 

### 2023-09-22 manuscript/main.tex
integrate feedback from Niklas

@@
 	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.
-The respective algorithms were already successfully applied to build and visualize large scale pangenome graphs of the Human Pangenome Resource Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.
+The respective algorithms were already successfully applied to build and visualize large scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.
 
@@
 	In the 2D layouts, nodes have two ends.
-	When moving a pair of nodes, we actually move one end of each node.
+	When moving a pair of nodes, we actually move one end of each node. For clarification, an example is given in Figure~\ref{fig:sketches}.
 	$v_i$ is the node associated with the step $s_i$ sampled uniformly from all the steps in $\mathcal{P}$.
-	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by drawing a Zipfian distribution \citep{Zipf1932}.
+	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by drawing a uniform or a Zipfian distribution \citep{Zipf1932}.
 	The difference between $nd_{ij}$ and $ld_{ij}$ guides the update of the node coordinates in the layout.
@@
 
-    \subsection{Pangenome graphs layouts reveal biology features}
+    \subsection{Pangenome graph layouts reveal biology features}
 	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent. 
@@
 	%The layout is the same as the one produced by \textit{Bandage}~\citep{Wick_2015} and displayed in~\citep{Guarracino2022}.
-	Complementary, we provide various 1D visualizations of in Supplementary Figure~\ref{fig:1d_sorts}.
+	Complementary, we provide various 1D visualizations in Supplementary Figure~\ref{fig:1d_sorts}.
 
@@
 	%Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
-	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
+	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graphs with annotation.
 	Our algorithm will underpin new pangenome graph browsers for studying graph layouts and the genome variation they represent (\url{https://github.com/chfi/waragraph)} last accessed Jul 2023).

### 2023-09-22 manuscript/pg-sgd_pseudocode.tex
pseudocode feedback from Niklas

@@
 				}
-				$\delta \gets \mu \cdot \frac{ld -nd_{ij}}{2}$ \tcp{the actual delta}
+				$\delta \gets abs(\mu \cdot \frac{ld_{ij} -nd_{ij}}{2})$ \tcp{the actual delta}
 				\uIf{$\delta <= 0$} {
@@
 				$r \gets \delta - ld_{ij}$ \tcp{size of the update}
-				$l_i \gets l_i - r\cdot ld_{ij}$ \tcp{update $v_i$ coordinates}
+				$l_i \gets l_i + r\cdot ld_{ij}$ \tcp{update $v_i$ coordinates}
 				$l_j \gets l_j - r\cdot ld_{ij}$ \tcp{update $v_j$ coordinates}

### 2023-09-22 manuscript/main.tex
small stuff

@@
 	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.
-The respective algorithms were already successfully applied to build and visualize large scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.
+	These algorithms have already been successfully applied to construct and visualize large-scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.
 

### 2023-09-25 manuscript/pg-sgd_pseudocode.tex
fix abs(delta)
1D
update pseudocode caption
make Niklas hoppy

--- 2023-07-01 manuscript/pg-sgd_pseudocode.tex
@@
 				}
-				$\delta \gets \mu \cdot \frac{ld -nd_{ij}}{2}$ \tcp{the actual delta}
-				\uIf{$\delta <= 0$} {
+				$\delta \gets \mu \cdot \frac{ld_{ij} -nd_{ij}}{2}$ \tcp{the actual delta}
+				\uIf{abs($\delta) <= 0$} {
 					$STOP$ \tcp{we can't optimize more}
@@
 				% TODO potentially store new delta max?
-				$r \gets \delta - ld_{ij}$ \tcp{size of the update}
-				$l_i \gets l_i - r\cdot ld_{ij}$ \tcp{update $v_i$ coordinates}
-				$l_j \gets l_j - r\cdot ld_{ij}$ \tcp{update $v_j$ coordinates}
+				$r \gets \frac{\delta}{ld_{ij}}$ \tcp{size of the update}
+				$r_x \gets r \cdot ld_{ij}$ \tcp{update size normalized by layout distance}
+				$l_i \gets l_i + r_{x}$ \tcp{update $v_i$ coordinates}
+				$l_j \gets l_j - r_{x}$ \tcp{update $v_j$ coordinates}
 			}
@@
 	}
-	\caption{Pseudocode of PG-SGD.}
+	\caption{Pseudocode of PG-SGD in 1D.}
 	\label{alg:pg_sgd}

### 2023-09-25 manuscript/pg-sgd_pseudocode.tex
niklas is happy now

@@
 				$r \gets \frac{\delta}{ld_{ij}}$ \tcp{size of the update}
-				$r_x \gets r \cdot ld_{ij}$ \tcp{update size normalized by layout distance}
+				$r_x \gets r \cdot (l_{i} - l_{j})$ \tcp{update size normalized by layout distance}
 				$l_i \gets l_i + r_{x}$ \tcp{update $v_i$ coordinates}

### 2023-10-09 manuscript/main.tex
no bioinf
fix abs(delta)
initial

@@
         Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}3,4,\dagger}$,
-        Jan-Niklas Manuel Schmelzle\,$^{\orcidlink{0000-0001-8566-4049}5,6}$,
+        Jan-Niklas M. Schmelzle\,$^{\orcidlink{0000-0001-8566-4049}5,6}$,
         Jiajie Li\,$^{\orcidlink{0000-0002-6775-2843}6}$,
@@
 	\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
-	\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
+	%\textbf{Supplementary information:} Supplementary data are available online. %at \textit{Bioinformatics} online.
 	}

### 2023-10-13 manuscript/main.tex
add feedback from Sven

@@
         Jörg Hagmann\,${\orcidlink{0000-0002-7232-3103}^7}$,
-        Sven Nahnsen\,$^{\orcidlink{0000-0002-4375-0691}1,2}$,
+        Sven Nahnsen\,$^{\orcidlink{0000-0002-4375-0691}1,2,8}$,
         Pjotr Prins\,$^{\orcidlink{0000-0002-8021-9162}3}$,
@@
         $^7$Computomics GmbH, Eisenbahnstr. 1, 72072 Tübingen, Germany \\
+        $^8$M3 Research Center, University Hospital Tübingen, 72076 Tübingen, Germany \\
     }
@@
 	\section{Introduction}
-	Reference genomes are widely used in genetics, serving as a foundation for a variety of analyses, including gene annotation, read mapping, and variant detection \citep{Singh2022}.
+	Reference genomes are widely used in genomics, serving as a foundation for a variety of analyses, including gene annotation, read mapping, and variant detection \citep{Singh2022}.
 	However, this linear model is becoming obsolete given the accessibility to hundreds or even thousands of high-quality genomes.

### 2023-10-16 manuscript/main.tex
suggestions from Erik

@@
 	The increasing availability of complete genomes demands for models to study genomic variability within entire populations.
-	Pangenome graphs capture the full genetic diversity between multiple genomes, but their layouts may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
-	These structures hamper downstream analyses, visualization, and interpretation. \\
+	Pangenome graphs capture the full genomic diversity between multiple genomes. 
+	In order to understand them, we need to see them. 
+	For visualization, we need a human readable graph layout: A graph embedding in low (e.g. two) dimensional depictions.
+	Due to a pangenome graph's potential excessive size, this is a grand challenge.
+	%but their layouts may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
+	%These structures hamper downstream analyses, visualization, and interpretation. 
+	\\
 	\textbf{Results:}
 	In response, we introduce a novel graph layout algorithm: the Path-Guided Stochastic Gradient Descent (PG-SGD).
-	PG-SGD uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel applying a modified HOGWILD! strategy.
-	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, unveiling their biological features. \\
+	PG-SGD uses the genomic paths in the pangenome graph as an embedded positional system to sample genomic distances between pairs of nodes, avoiding the quadratic cost seen in previous versions of graph drawing by Stochastic Gradient Descent (SGD).
+	We show that our implementation efficiently computes the low dimensional layouts of gigabase-scale pangenome graphs, unveiling their biological features. \\
 	\textbf{Availability:}
@@
 	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.
+
+	However, \cite{Zheng2019}'s algorithm has a quadratic up front cost in the number of nodes to find pairwise distances to guide the layout, making it impossible to apply to pangenome graphs with millions of nodes.
 	%Specifically, in pangenome graphs we move a single pair of nodes at a time.
@@
 	Since pangenome graphs represent genomes as paths in the graph, a reasonable distance metric would be the nucleotide distance between a pair of nodes traversed by the same path.
+Such path sampling would overcome the quadratic costs of previous versions of graph drawing by SGD.
 
-	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in parallel with a modified HOGWILD! strategy.
+	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to use the paths as an embedded positional system to find distances between nodes, moving pairs of nodes in parallel with a modified HOGWILD!
+strategy.
 	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.

### 2023-10-16 manuscript/main.tex
tweak

@@
 	In response, we introduce a novel graph layout algorithm: the Path-Guided Stochastic Gradient Descent (PG-SGD).
-	PG-SGD uses the genomic paths in the pangenome graph as an embedded positional system to sample genomic distances between pairs of nodes, avoiding the quadratic cost seen in previous versions of graph drawing by Stochastic Gradient Descent (SGD).
-	We show that our implementation efficiently computes the low dimensional layouts of gigabase-scale pangenome graphs, unveiling their biological features. \\
+	PG-SGD uses the genomes, represented in the pangenome graph as paths, as an embedded positional system to sample genomic distances between pairs of nodes.
+	This avoids the quadratic cost seen in previous versions of graph drawing by Stochastic Gradient Descent (SGD).We show that our implementation efficiently computes the low dimensional layouts of gigabase-scale pangenome graphs, unveiling their biological features.
+\\
 	\textbf{Availability:}

### 2023-10-16 manuscript/main.tex
ene mine mous

@@
 	This is expected since chromosome 16 has one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
-	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
+	Repetitive sequences lead to graph nodes with a very high number of path steps, which are computationally expensive to work with \citep{Guarracino2022}.
 	Memory consumption is 29.66 GB of RAM on average, with the memory peak again occurring with chromosome 16, due to the path index building phase.

### 2023-10-16 manuscript/main.tex
fix header

@@
 
-    \subsection{Pangenome graph layouts reveal biology features}
+    \subsection{Pangenome graph layouts reveal biological features}
 	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent. 

### 2023-10-17 manuscript/main.tex
more suggestions

@@
 	The increasing availability of complete genomes demands for models to study genomic variability within entire populations.
-	Pangenome graphs capture the full genomic diversity between multiple genomes. 
+	Pangenome graphs capture the full genomic similarity and diversity between multiple genomes. 
 	In order to understand them, we need to see them. 
 	For visualization, we need a human readable graph layout: A graph embedding in low (e.g. two) dimensional depictions.
-	Due to a pangenome graph's potential excessive size, this is a grand challenge.
+	Due to a pangenome graph's potential excessive size, this is a signifiant challenge.
 	%but their layouts may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
@@
 
-	A pangenome graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space to produce a human-readable visualization of genetic variation between multiple genomes.
-	Layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive graph understanding.
+	A graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space.
+	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
 	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts.
@@
 
+	A \textit{pangenome} graph layout can provide a human-readable visualization of genetic variation between multiple genomes.
 	However, \cite{Zheng2019}'s algorithm has a quadratic up front cost in the number of nodes to find pairwise distances to guide the layout, making it impossible to apply to pangenome graphs with millions of nodes.
+	Also, existing graph layout approaches ignore the biological information inherent in pangenome graphs.
 	%Specifically, in pangenome graphs we move a single pair of nodes at a time.
-
-	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the lock-free HOGWILD!
-method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.
 
@@
 Such path sampling would overcome the quadratic costs of previous versions of graph drawing by SGD.
+
+	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the lock-free HOGWILD!
+method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.
 

### 2023-10-17 manuscript/main.tex
typo
WHY

@@
 	For visualization, we need a human readable graph layout: A graph embedding in low (e.g. two) dimensional depictions.
-	Due to a pangenome graph's potential excessive size, this is a signifiant challenge.
+	Due to a pangenome graph's potential excessive size, this is a significant challenge.
 	%but their layouts may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
@@
 	PG-SGD uses the genomes, represented in the pangenome graph as paths, as an embedded positional system to sample genomic distances between pairs of nodes.
-	This avoids the quadratic cost seen in previous versions of graph drawing by Stochastic Gradient Descent (SGD).We show that our implementation efficiently computes the low dimensional layouts of gigabase-scale pangenome graphs, unveiling their biological features.
+	This avoids the quadratic cost seen in previous versions of graph drawing by Stochastic Gradient Descent (SGD).
+We show that our implementation efficiently computes the low dimensional layouts of gigabase-scale pangenome graphs, unveiling their biological features.
 \\

### 2024-02-06 manuscript/main.tex
the story so far

@@
 	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references (see Supplementary Table~\ref{tab:layout} for graph statistics).
-	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, 2D PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
+	\red{When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, \textit{odgi layout}'s 2D PG-SGD implementation obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).}
 	This is expected since chromosome 16 has one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
@@
 	Memory consumption is 29.66 GB of RAM on average, with the memory peak again occurring with chromosome 16, due to the path index building phase.
-	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}). 
+	\red{Given its scalability, we applied 2D PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}).} 
 	%A 2D layout was calculated in under one day.
-	\textit{BandageNG} (\url{https://github.com/asl/BandageNG}, last accessed Jul 2023), the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
-	On average, PG-SGD is $\sim$8X faster than BandageNG while using $\sim$2X less memory.
+	\red{To compare, \textit{BandageNG} (\url{https://github.com/asl/BandageNG}, last accessed Jul 2023), the current state of the art for graph visualization, was used to calculate a 2D layout of each of the HPRC pangenome graphs.
+	For a fair comparison, we did not rely on \textit{BandageNG}'s interactive GUI application, but we executed \textit{BandageNG layout}, which directly emits a 2D graph layout similar to \textit{odgi layout}.
+	\textit{BandageNG} was not able to produce a layout for the full graph within 7 days, hitting the wall clock time limit of the cluster.
+	On average, PG-SGD is $\sim$8X faster than \textit{BandageNG} while using $\sim$2X less memory.}
 	% avg time bng 388 min, avg gb ng 66.84

### 2024-02-08 manuscript/main.tex
gimme more

@@
 	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
-	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts.
+	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts. %by ....
 	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
@@
 	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
-	To our knowledge, no algorithm takes into account such biological information to compute the graph layouts.
+	\red{To our knowledge, no generic graph layout algorithm takes into account such path encoded biological information when computing a graph's layout.}
 	PG-SGD can be extended in any number of dimensions.
@@
 
-	We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
+	\red{We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
+	Other generic graph layout algorithms, such as the one offered by \textit{BandageNG}, ignore this additional information.}
 	Our implementation efficiently computes the layout of pangenome graphs representing thousands of whole genomes.

### 2024-02-09 manuscript/main.tex
reviewer2 q2
r2 q3

@@
 	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
-	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts. %by ....
-	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
-	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.
+	\red{One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which uses physical simulation to produce aesthetic layouts. %by ....
+	This is prone to get stuck in local minima, but multi-layer strategies such as the Fast Multipole Multilevel (FM$^{3}$)~\citep{Hachul2005} method or stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
+	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.}
 
@@
 	However, \cite{Zheng2019}'s algorithm has a quadratic up front cost in the number of nodes to find pairwise distances to guide the layout, making it impossible to apply to pangenome graphs with millions of nodes.
-	Also, existing graph layout approaches ignore the biological information inherent in pangenome graphs.
+	\red{Also, existing generic graph layout approaches ignore the biological information inherent in pangenome graphs.
+One such bioinformatics tool is \textit{BandageNG}, the current state of the art for genome graph visualization.}
 	%Specifically, in pangenome graphs we move a single pair of nodes at a time.
@@
 
-	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}, but the lock-free HOGWILD!
-method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.
+	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}.
+\red{Although, \textit{BandageNG} applies FM$^{3}$ for layout generation, it's parallelism is bound by the number of connected graph components.
+Alternatively, the lock-free HOGWILD!
+method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.}
 
@@
 	\red{To our knowledge, no generic graph layout algorithm takes into account such path encoded biological information when computing a graph's layout.}
+
 	PG-SGD can be extended in any number of dimensions.
@@
 	These algorithms have already been successfully applied to construct and visualize large-scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.
+	\red{In addition, we show that PG-SGD is almost an order of magnitude faster than \textit{BandageNG}.}
 
@@
 	The data structure could be optimized to improve cache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU. 
+	\red{Since BandageNG was primarily designed for assembly graphs, one may have to adjust its parameters dependent on the input graph, in order to boost the layout generation or to adjust the highlighting of desired graph features.}
 	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.

### 2024-02-09 manuscript/main.tex
r2 q4

@@
 	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
-	\red{One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which uses physical simulation to produce aesthetic layouts. %by ....
+	\red{One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which uses physical simulation to produce aesthetic layouts.
+The classical approach combines repulsive forces on all vertices and attractive forces on adjacent vertices.
+%by ....
 	This is prone to get stuck in local minima, but multi-layer strategies such as the Fast Multipole Multilevel (FM$^{3}$)~\citep{Hachul2005} method or stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
@@
 	\red{Also, existing generic graph layout approaches ignore the biological information inherent in pangenome graphs.
-One such bioinformatics tool is \textit{BandageNG}, the current state of the art for genome graph visualization.}
+One such bioinformatics tool is \textit{BandageNG}, the current state of the art for genome graph visualization.
+It uses FM$^{3}$ which only considers the nodes and edges of a graph.}
 	%Specifically, in pangenome graphs we move a single pair of nodes at a time.
@@
 	The data structure could be optimized to improve cache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU. 
-	\red{Since BandageNG was primarily designed for assembly graphs, one may have to adjust its parameters dependent on the input graph, in order to boost the layout generation or to adjust the highlighting of desired graph features.}
+	\red{In \textit{BandageNG}, one can't select the number of threads for the calculations.
+They are automatically chosen by the number of connected components of the graph to draw.
+This limits its parallelism and leads to an unbalanced workload.
+Since \textit{BandageNG} was primarily designed for assembly graphs, one may have to adjust its parameters dependent on the input graph, in order to boost the layout generation or to adjust the highlighting of desired graph features.}
+
+	\red{The classical force model of state of the art generic graph algorithms, such as FM³ based-ones, places nodes according to their attractive and repulsive forces.
+	This force can be seen as equivalent to how our 2D PG-SGD moves the nodes’ ends in 2D: If the nucleotide distance of the randomly chose path steps is smaller than the layout distance of the nodes’ ends, we move them closer together (“attractive force”), else we move them further away (“repulsive force”).
+	However, the key difference here is that this approach is path-guided using a “biological force”. 
+	So theoretically, it would be possible to combine our approach with a force-directed one. 
+	Combining both methods, we might get the best of both worlds: Multi-threadable PG-SGD iteratively applied to different graph-layout-levels. 
+	We can imagine that such an approach can lead to a further speedup when calculating the layout.
+	However, for generic graphs, this would only work if quantitative path information for each node could be added: We would replace the classical physical simulation approach with our path-guided method.
+	If such information is not available, one could randomly cover the graph with paths. Staying in the pangenomic world there already exists such a tool: \textit{odgi cover}. 
+	However, this is an NP-hard problem and so far didn’t work very well in practice.}
+
+	\red{With assembly graphs we face the same problem: They usually do not carry path information during each assembly step. 
+	One could map the initial assembly back against the assembly graph in order to build paths through the graph. This would allow us to obtain a layout using PG-SGD.}	 
 	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.

### 2024-02-10 manuscript/main.tex
Made in Italy

@@
 
-	\red{The classical force model of state of the art generic graph algorithms, such as FM³ based-ones, places nodes according to their attractive and repulsive forces.
-	This force can be seen as equivalent to how our 2D PG-SGD moves the nodes’ ends in 2D: If the nucleotide distance of the randomly chose path steps is smaller than the layout distance of the nodes’ ends, we move them closer together (“attractive force”), else we move them further away (“repulsive force”).
-	However, the key difference here is that this approach is path-guided using a “biological force”. 
+	\red{The classical force model of state of the art generic graph algorithms, such as FM³-based ones, places nodes according to their attractive and repulsive forces.
+	This force can be seen as equivalent to how our 2D PG-SGD moves the nodes’ ends in 2D: If the nucleotide distance of the randomly chosen path steps is smaller than the layout distance of the nodes' ends, we move them closer together (“attractive force”), else we move them further away (“repulsive force”).
+	However, the key difference here is that this approach is path-guided using a “biological force”, but paths represent biological sequences in pangenome graphs, so it is as if PG-SDG considers a “biological force” for placing the graph nodes.
 	So theoretically, it would be possible to combine our approach with a force-directed one. 
@@
 	We can imagine that such an approach can lead to a further speedup when calculating the layout.
-	However, for generic graphs, this would only work if quantitative path information for each node could be added: We would replace the classical physical simulation approach with our path-guided method.
-	If such information is not available, one could randomly cover the graph with paths. Staying in the pangenomic world there already exists such a tool: \textit{odgi cover}. 
-	However, this is an NP-hard problem and so far didn’t work very well in practice.}
+	However, for generic graphs, this would only work if path information for each node could be added: We would replace the classical physical simulation approach with our path-guided method. 
+	If such information is not available, one could randomly cover the graph with paths.
+	This function is already provided in \textit{odgi cover}. 
+	However, this is an NP-hard problem and so far didn't work very well in practice.}
 

### 2024-02-12 manuscript/main.tex
minor tweaks

@@
 	This force can be seen as equivalent to how our 2D PG-SGD moves the nodes’ ends in 2D: If the nucleotide distance of the randomly chosen path steps is smaller than the layout distance of the nodes' ends, we move them closer together (“attractive force”), else we move them further away (“repulsive force”).
-	However, the key difference here is that this approach is path-guided using a “biological force”, but paths represent biological sequences in pangenome graphs, so it is as if PG-SDG considers a “biological force” for placing the graph nodes.
+	However, the key difference here is that this approach is path-guided: Paths represent biological sequences in pangenome graphs, so it is as if PG-SDG considers a “biological force” for placing the graph nodes.
 	So theoretically, it would be possible to combine our approach with a force-directed one. 
@@
 	\red{With assembly graphs we face the same problem: They usually do not carry path information during each assembly step. 
-	One could map the initial assembly back against the assembly graph in order to build paths through the graph. This would allow us to obtain a layout using PG-SGD.}	 
+	One could map the initial assembly reads back against the assembly graph in order to build paths through the graph. This would allow us to obtain a layout using PG-SGD.}	 
 	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.

### 2024-02-12 manuscript/main.tex
update affiliations

@@
 	\author[Heumos, Guarracino \textit{et~al}.]{
-        Simon~Heumos\,$^{\orcidlink{0000-0003-3326-817X}1,2,\dagger}$,
-        Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}3,4,\dagger}$,
-        Jan-Niklas M. Schmelzle\,$^{\orcidlink{0000-0001-8566-4049}5,6}$,
-        Jiajie Li\,$^{\orcidlink{0000-0002-6775-2843}6}$,
-        Zhiru Zhang\,$^{\orcidlink{0000-0002-0778-0308}6}$,
-        Jörg Hagmann\,${\orcidlink{0000-0002-7232-3103}^7}$,
-        Sven Nahnsen\,$^{\orcidlink{0000-0002-4375-0691}1,2,8}$,
-        Pjotr Prins\,$^{\orcidlink{0000-0002-8021-9162}3}$,
-        Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 3},*}$
+        Simon~Heumos\,$^{\orcidlink{0000-0003-3326-817X}1,2,3,4\dagger}$,
+        Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}5,6,\dagger}$,
+        Jan-Niklas M. Schmelzle\,$^{\orcidlink{0000-0001-8566-4049}7,8}$,
+        Jiajie Li\,$^{\orcidlink{0000-0002-6775-2843}8}$,
+        Zhiru Zhang\,$^{\orcidlink{0000-0002-0778-0308}8}$,
+        Jörg Hagmann\,${\orcidlink{0000-0002-7232-3103}^9}$,
+        Sven Nahnsen\,$^{\orcidlink{0000-0002-4375-0691}1,2,3,4}$,
+        Pjotr Prins\,$^{\orcidlink{0000-0002-8021-9162}5}$,
+        Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 5},*}$
     }
@@
         $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen, Tübingen 72076, Germany \\
-        $^3$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
-        $^4$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
-        $^5$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
-        $^6$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
-        $^7$Computomics GmbH, Eisenbahnstr. 1, 72072 Tübingen, Germany \\
-        $^8$M3 Research Center, University Hospital Tübingen, 72076 Tübingen, Germany \\
+        $^3$M3 Research Center, University Hospital Tübingen, Tübingen 72076 , Germany \\
+        $^4$Institute for Bioinformatics and Medical Informatics (IBMI), University of Tübingen, Tübingen 72076, Germany \\
+        $^5$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
+        $^6$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
+        $^7$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
+        $^8$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
+        $^9$Computomics GmbH, Eisenbahnstr. 1, 72072 Tübingen, Germany \\
     }

### 2024-02-12 scripts/cfc/cfc_sling_32T.sh
mirror scripts

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+/usr/bin/time --verbose singularity exec /nfsmounts/container/pg-sgd.img odgi layout -i /sfs/9/ws/afaaj01-sorting-paper/graphs/chr"$1".hprc-v1.0-pggb.gfa -o /sfs/9/ws/afaaj01-sorting-paper/layouts/chr"$1".hprc-v1.0-pggb.gfa.32T.lay -t 32 -C "$TMPDIR" 2> /sfs/9/ws/afaaj01-sorting-paper/times/odgi_layout_32T_"$1".time

### 2024-02-12 scripts/cfc/hprc-v1.0-bandage-ng.layout.chroms.32T.sh

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+# wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/freeze/freeze1/pggb/hprc-v1.0-pggb.gfa.gz
+# gunzip hprc-v1.0-pggb.gfa.gz
+XDG_RUNTIME_DIR=$TMPDIR
+#/usr/bin/time --verbose singularity exec /nfsmounts/container/pg-sgd.img BandageNG layout /sfs/9/ws/afaaj01-sorting-paper/graphs/chr"$1".hprc-v1.0-pggb.gfa /sfs/9/ws/afaaj01-sorting-paper/layouts/chr"$1".hprc-v1.0-pggb.gfa.bng.32T.layout 2> /sfs/9/ws/afaaj01-sorting-paper/times/bng_layout_"$1".time
+/usr/bin/time --verbose singularity exec /nfsmounts/container/pg-sgd.img BandageNG layout /sfs/9/ws/afaaj01-sorting-paper/graphs/chr"$1".hprc-v1.0-pggb.gfa /sfs/9/ws/afaaj01-sorting-paper/layouts/chr"$1".hprc-v1.0-pggb.gfa.bng.32T.layout 2> /sfs/9/ws/afaaj01-sorting-paper/times/bng_layout_32T_"$1".time
+# /usr/bin/time --verbose singularity exec /nfsmounts/container/pg-sgd.img odgi layout -i /sfs/9/ws/afaaj01-sorting-paper/graphs/chr"$1".hprc-v1.0-pggb.gfa -o /sfs/9/ws/afaaj01-sorting-paper/layouts/chr"$1".hprc-v1.0-pggb.gfa.lay -t 64 -C "$TMPDIR" 2> /sfs/9/ws/afaaj01-sorting-paper/times/odgi_layout_"$1".time
+# /usr/bin/time --verbose srun -p qbic --mem=500000 -J bandage-ng -c 64 -t 10080 /usr/bin/time --verbose singularity exec /nfsmounts/container/pg-sgd.img Bandage image hprc-v1.0-pggb.gfa hprc-v1.0-pggb.gfa.image.png 2> hprc-v1.0.bandage-ng.image.time

### 2024-02-19 manuscript/main.tex
feedback Niklas
supplementary

@@
 	This function is already provided in \textit{odgi cover}. 
-	However, this is an NP-hard problem and so far didn't work very well in practice.}
+	However, this is an NP-hard problem and our preliminary solutions proved ineffective.}
 
@@
 
-	\clearpage
-	\setcounter{page}{1}
+	%\clearpage
+	%\setcounter{page}{1}
 
-	\beginsupplement
+	%\beginsupplement
 
-	\section{Supplement}
-
-	 \subsection{Supplementary data}
+	%\section{Supplement}
+	%
+	% \subsection{Supplementary data}
 	 %\subsection{Quantization} % DURING REVISION, IF NEEDED
 	 %More details about your clever quantization.
-	 \subsubsection{Performance evaluation}
-	 \label{sec:performance}
-	 The results of the performance evaluation are given in Table~\ref{tab:layout}.
-	 \input{table_layout}
+	% \subsubsection{Performance evaluation}
+	% \label{sec:performance}
+	% The results of the performance evaluation are given in Table~\ref{tab:layout}.
+	% \input{table_layout}
 	 %\FIXME{What about the SMT thingy?} % DURING REVISION, IF NEEDED
-	\subsubsection{1D visualizations}	
-	The 1D PG-SGD algorithm creates a 1D layout of the nodes of the graph.
-	Theoretically, it is possible that 2 nodes have the same 1D coordinate or overlap.
-	But, in our 1D visualizations, we arrange the nodes from left to right.
-	Therefore, we project the 1D coordinates into a 1D node order: We sort the final layout by graph component, graph position, and node rank.
-	\input{fig_1d_sorts}
+	%\subsubsection{1D visualizations}	
+	%The 1D PG-SGD algorithm creates a 1D layout of the nodes of the graph.
+	%Theoretically, it is possible that 2 nodes have the same 1D coordinate or overlap.
+	%But, in our 1D visualizations, we arrange the nodes from left to right.
+	%Therefore, we project the 1D coordinates into a 1D node order: We sort the final layout by graph component, graph position, and node rank.
+	%\input{fig_1d_sorts}
 

### 2024-04-19 manuscript/main.tex
remove red

@@
 	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
-	\red{One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which uses physical simulation to produce aesthetic layouts.
+	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which uses physical simulation to produce aesthetic layouts.
 The classical approach combines repulsive forces on all vertices and attractive forces on adjacent vertices.
@@
 	This is prone to get stuck in local minima, but multi-layer strategies such as the Fast Multipole Multilevel (FM$^{3}$)~\citep{Hachul2005} method or stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
-	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.}
+	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.
 
@@
 	However, \cite{Zheng2019}'s algorithm has a quadratic up front cost in the number of nodes to find pairwise distances to guide the layout, making it impossible to apply to pangenome graphs with millions of nodes.
-	\red{Also, existing generic graph layout approaches ignore the biological information inherent in pangenome graphs.
+	Also, existing generic graph layout approaches ignore the biological information inherent in pangenome graphs.
 One such bioinformatics tool is \textit{BandageNG}, the current state of the art for genome graph visualization.
-It uses FM$^{3}$ which only considers the nodes and edges of a graph.}
+It uses FM$^{3}$ which only considers the nodes and edges of a graph.
 	%Specifically, in pangenome graphs we move a single pair of nodes at a time.
@@
 	Typically, force-directed layouts are hard to compute \citep{wang_research_2014}.
-\red{Although, \textit{BandageNG} applies FM$^{3}$ for layout generation, it's parallelism is bound by the number of connected graph components.
+Although, \textit{BandageNG} applies FM$^{3}$ for layout generation, it's parallelism is bound by the number of connected graph components.
 Alternatively, the lock-free HOGWILD!
-method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.}
+method offers a highly parallelizable and thus scalable SGD approach that can be applied when the optimization problem is sparse \citep{Recht2011}.
 
@@
 	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
-	\red{To our knowledge, no generic graph layout algorithm takes into account such path encoded biological information when computing a graph's layout.}
+	To our knowledge, no generic graph layout algorithm takes into account such path encoded biological information when computing a graph's layout.
 
@@
 	These algorithms have already been successfully applied to construct and visualize large-scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.
-	\red{In addition, we show that PG-SGD is almost an order of magnitude faster than \textit{BandageNG}.}
+	In addition, we show that PG-SGD is almost an order of magnitude faster than \textit{BandageNG}.
 
@@
 	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references (see Supplementary Table~\ref{tab:layout} for graph statistics).
-	\red{When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, \textit{odgi layout}'s 2D PG-SGD implementation obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).}
+	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, \textit{odgi layout}'s 2D PG-SGD implementation obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
 	This is expected since chromosome 16 has one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
@@
 	Memory consumption is 29.66 GB of RAM on average, with the memory peak again occurring with chromosome 16, due to the path index building phase.
-	\red{Given its scalability, we applied 2D PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}).} 
+	Given its scalability, we applied 2D PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}).
 	%A 2D layout was calculated in under one day.
-	\red{To compare, \textit{BandageNG} (\url{https://github.com/asl/BandageNG}, last accessed Jul 2023), the current state of the art for graph visualization, was used to calculate a 2D layout of each of the HPRC pangenome graphs.
+	To compare, \textit{BandageNG} (\url{https://github.com/asl/BandageNG}, last accessed Jul 2023), the current state of the art for graph visualization, was used to calculate a 2D layout of each of the HPRC pangenome graphs.
 	For a fair comparison, we did not rely on \textit{BandageNG}'s interactive GUI application, but we executed \textit{BandageNG layout}, which directly emits a 2D graph layout similar to \textit{odgi layout}.
 	\textit{BandageNG} was not able to produce a layout for the full graph within 7 days, hitting the wall clock time limit of the cluster.
-	On average, PG-SGD is $\sim$8X faster than \textit{BandageNG} while using $\sim$2X less memory.}
+	On average, PG-SGD is $\sim$8X faster than \textit{BandageNG} while using $\sim$2X less memory.
 	% avg time bng 388 min, avg gb ng 66.84
@@
 
-	\red{We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
-	Other generic graph layout algorithms, such as the one offered by \textit{BandageNG}, ignore this additional information.}
+	We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
+	Other generic graph layout algorithms, such as the one offered by \textit{BandageNG}, ignore this additional information.
 	Our implementation efficiently computes the layout of pangenome graphs representing thousands of whole genomes.
@@
 	The data structure could be optimized to improve cache performance. Moreover, the high-degree of parallelism could be further exploited by using a GPU. 
-	\red{In \textit{BandageNG}, one can't select the number of threads for the calculations.
+	In \textit{BandageNG}, one can't select the number of threads for the calculations.
 They are automatically chosen by the number of connected components of the graph to draw.
 This limits its parallelism and leads to an unbalanced workload.
-Since \textit{BandageNG} was primarily designed for assembly graphs, one may have to adjust its parameters dependent on the input graph, in order to boost the layout generation or to adjust the highlighting of desired graph features.}
+Since \textit{BandageNG} was primarily designed for assembly graphs, one may have to adjust its parameters dependent on the input graph, in order to boost the layout generation or to adjust the highlighting of desired graph features.
 
-	\red{The classical force model of state of the art generic graph algorithms, such as FM³-based ones, places nodes according to their attractive and repulsive forces.
+	The classical force model of state of the art generic graph algorithms, such as FM³-based ones, places nodes according to their attractive and repulsive forces.
 	This force can be seen as equivalent to how our 2D PG-SGD moves the nodes’ ends in 2D: If the nucleotide distance of the randomly chosen path steps is smaller than the layout distance of the nodes' ends, we move them closer together (“attractive force”), else we move them further away (“repulsive force”).
@@
 	This function is already provided in \textit{odgi cover}. 
-	However, this is an NP-hard problem and our preliminary solutions proved ineffective.}
+	However, this is an NP-hard problem and our preliminary solutions proved ineffective.
 
-	\red{With assembly graphs we face the same problem: They usually do not carry path information during each assembly step. 
-	One could map the initial assembly reads back against the assembly graph in order to build paths through the graph. This would allow us to obtain a layout using PG-SGD.}	 
+	With assembly graphs we face the same problem: They usually do not carry path information during each assembly step. 
+	One could map the initial assembly reads back against the assembly graph in order to build paths through the graph. This would allow us to obtain a layout using PG-SGD.
 	% As especially observed with chromosome 16 evaluation, our implementation is memory-bound.

### 2024-04-19 manuscript/main.tex
update and PDFs files

@@
 
-\copyrightyear{2023} \pubyear{XXXX}
+\copyrightyear{2024} \pubyear{XXXX}
 
@@
 
-	Software versions, code, and links to data resources used to prepare this manuscript can be found at \url{https://github.com/pangenome/sorting-paper}.
+	Software versions, code, and links to data used to prepare this manuscript can be found at \url{https://github.com/pangenome/sorting-paper}.
 	Animations of the algorithm are deposited at \url{https://doi.org/10.5281/zenodo.8288999}.
@@
 
-	%\clearpage
-	%\setcounter{page}{1}
+	\clearpage
+	\setcounter{page}{1}
 
-	%\beginsupplement
+	\beginsupplement
 
-	%\section{Supplement}
-	%
+	\section{Supplement}
+
 	% \subsection{Supplementary data}
@@
 	 %More details about your clever quantization.
-	% \subsubsection{Performance evaluation}
-	% \label{sec:performance}
-	% The results of the performance evaluation are given in Table~\ref{tab:layout}.
-	% \input{table_layout}
+	\subsubsection{Performance evaluation}
+	\label{sec:performance}
+	The results of the performance evaluation are given in Table~\ref{tab:layout}.
+	\input{table_layout}
 	 %\FIXME{What about the SMT thingy?} % DURING REVISION, IF NEEDED
-	%\subsubsection{1D visualizations}	
-	%The 1D PG-SGD algorithm creates a 1D layout of the nodes of the graph.
-	%Theoretically, it is possible that 2 nodes have the same 1D coordinate or overlap.
-	%But, in our 1D visualizations, we arrange the nodes from left to right.
-	%Therefore, we project the 1D coordinates into a 1D node order: We sort the final layout by graph component, graph position, and node rank.
-	%\input{fig_1d_sorts}
+	\subsubsection{1D visualizations}	
+	The 1D PG-SGD algorithm creates a 1D layout of the nodes of the graph.
+	Theoretically, it is possible that 2 nodes have the same 1D coordinate or overlap.
+	But, in our 1D visualizations, we arrange the nodes from left to right.
+	Therefore, we project the 1D coordinates into a 1D node order: We sort the final layout by graph component, graph position, and node rank.
+	\input{fig_1d_sorts}
 
